---
title: "Differential abundance analysis, untargeted approach"
subtitle: "Positive polarity"
author: "Christa Malfertheiner, Johannes Rainer"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r setup, echo = FALSE, results = "asis", warning = FALSE}
library(BiocStyle)
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r parameters, echo = FALSE, warning = FALSE}
## Set general parameters
polarity <- "POS" # specify "POS" or "NEG"
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.7      # cut-off for log2 fold change
set.seed(123)

## Setting golden ratio to save images
phi <- (1+sqrt(5))/2

FILE_NAME <- "untargeted_matrix_pos"

## Define paths:
IMAGE_PATH <- paste0("images/", FILE_NAME, "/")
if (dir.exists(IMAGE_PATH)) unlink(IMAGE_PATH, recursive = TRUE, force = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE, showWarnings = FALSE)

RDATA_PATH <- paste0("data/RData/", FILE_NAME, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

RESULT_PATH <- paste0("data/results/", FILE_NAME, "/")
dir.create(RESULT_PATH, recursive = TRUE, showWarnings = FALSE)

FIG_PATH <- paste0(IMAGE_PATH, "FIG/")
dir.create(FIG_PATH, showWarnings = FALSE)
```

```{r}
## Version of `plot_pca` that uses a *real* minus and not a hyphen.
plot_pca <- function(pc, pch = 16, col = "#000000", pc_x = 1, pc_y = 2, 
                     main = "", labels = NULL, ...) {
    pcSummary <- summary(pc)
    plot(pc$x[, pc_x], pc$x[, pc_y], pch = NA, main = main,
         xlab = paste0("PC", pc_x, ": ",
                       format(pcSummary$importance[2, pc_x] * 100, 
                              digits = 3), " % variance"),
         ylab = paste0("PC", pc_y, ": ",
                       format(pcSummary$importance[2, pc_y] * 100, 
                              digits = 3), " % variance"),
         xaxt = "n", yaxt = "n", ...)
    xat <- axTicks(1, usr = par("usr")[1:2])
    labs <- gsub("-", "\U2212", print.default(xat))
    axis(1, at = xat, labels = labs)
    yat <- axTicks(2, usr = par("usr")[1:2])
    labs <- gsub("-", "\U2212", print.default(yat))
    axis(2, at = yat, labels = labs)    
    grid()
    if (!is.null(labels)) 
        text(pc$x[, pc_x], pc$x[, pc_y], labels = labels, col = col, 
             ...)
    else points(pc$x[, pc_x], pc$x[, pc_y], pch = pch, col = col, 
                ...)
}

```


# Introduction

In this document we perform the differential abundance analysis of the features
for the *MitYOU* project, with the aim of finding significant sample 
source-related features. This task is performed by hypothesis testing,
where we try to identify which metabolites have the most different 
concentrations between plasma samples, venous and capillary blood samples. 
We follow an untargeted approach, the analysis comprises feature pre-filtering,
exploratory analysis and differential abundance analysis.


# Data import

First, we load the required packages and the data, after preprocessing and
normalization. The end result of these steps is a `SummarizedExperiment` that
contains aligned data, where features are grouped (after correspondence), and
that have undergone gap filling, normalization by the median, linear fitting and 
per-feature between-batch normalization to remove any unwanted variability. 
The `SummarizedExperiment` lets us store all the information regarding the 
normalization steps in the form of `assays`, which we are still able to access 
to proceed with the analysis.

```{r load-data, echo = FALSE, warning = FALSE}
library(xcms)
library(limma)
library(pheatmap)
library(writexl)
library(SummarizedExperiment)
library(RColorBrewer)
library(MsFeatures)
library(CompMetaboTools)
library(pander)
setMSnbaseFastLoad(FALSE)
load("data/RData/vams_normalization_pos/res_pos.RData")
res_pos$sample_pair <- paste0(res_pos$source, ".", res_pos$sample)
```

It is important now to remove the `POOL` samples from the dataset, because the
analysis has to be performed only on study samples; the `POOL` samples, though
are still required to evaluate the goodness of the detected features, therefore
they will be stored in a separate `SummarizedExperiment` object that can be
accessed when needed. We also exclude the `RBC` samples from our analysis, as 
these samples showed a noisy signal beforehand. We also store them separately 
in a SummarizedExperiment object.

We assign the colours as seen before.

```{r split-qc, echo = TRUE}
res_qc <- res_pos[, res_pos$source == "all"]
res_pos <- res_pos[, res_pos$source != "all"]
res_rbc <- res_pos[, res_pos$source == "RBC"]
res_pos <- res_pos[, res_pos$source != "RBC"]

res_pos$source <- factor(as.character(res_pos$source))
res_pos$sex <- factor(as.character(res_pos$sex))

col_source <- brewer.pal(6, name = "Set1")[c(6, 2, 1)]
names(col_source) <- c("plasma",        #' yellow
                       "venous",        #' blue
                       "capillary")     #' red
                       
col_sex <- brewer.pal(4, name = "Set1") [c(1, 2, 3)]
names(col_sex) <- c("F",           # red
                    "M",           # blue
                    "POOL")        # green

## Setting golden ratio to save images
phi <- (1+sqrt(5))/2
```

```{r, echo = FALSE}
## Calculate variability across samples/replicates.
samps <- unique(res_pos$sample_pair)
M <- lapply(samps, function(z) {
    idx <- which(res_pos$sample_pair == z)
    if (length(idx) > 1) {
        abs(log2(assay(res_pos, "normalized_filled_imputed")[, idx[1L]]) -
            log2(assay(res_pos, "normalized_filled_imputed")[, idx[2L]]))
    } else NULL
})
names(M) <- samps
M <- do.call(cbind, M)
```

At last we also average the *technical* replicates for each individual to
generate the final data set.

```{r average-replicates}
#' Average 
averageSE <- function(x, column = character(), mainAssay = character()) {
    if (!column %in% colnames(colData(x)))
        stop("Column '", "' not found in 'colData' of 'x'")
    f <- factor(colData(x)[, column], levels = unique(colData(x)[, column]))
    ## new colData: take the first element for each replicate.
    cd <- colData(x)[match(levels(f), f), ]
    rownames(cd) <- cd[, column]
    ## loop over the assays and average them.
    a <- lapply(assays(x), function(z) {
        z <- split.data.frame(t(z), f = f)
        z <- do.call(cbind, lapply(z, colMeans, na.rm = TRUE))
        z[is.na(z)] <- NA
        z
    })
    if (length(mainAssay)) {
        tmp <- split.data.frame(t(assay(x, mainAssay)), f = f)
        tmp <- do.call(cbind, lapply(tmp, function(y) {
            apply(y, MARGIN = 2, FUN = sd, na.rm = TRUE)
        }))
        tmp[is.na(tmp)] <- NA
        a[[paste0(mainAssay, "_sd")]] <- tmp
    }
    SummarizedExperiment(assays = a, rowData = rowData(x),
                         colData = cd, metadata = metadata(x))
}
## Average technical replicates:
res_pos <- averageSE(res_pos, column = "source_sample",
                     mainAssay = "normalized_filled")
```

The samples used in this analysis are listed below.

```{r, echo = FALSE, results = "asis"}
tab <- colData(res_pos)[, c("source", "sex", "age")]
pandoc.table(as.data.frame(tab), style = "rmarkdown", split.tables = Inf,
             caption = "Samples used in this analysis")
```


# Feature pre-filtering

Feature pre-filtering is an important step of data analysis that aims to
reduce as much as possible the random error that occurs during the measurement
of an analyte: during this process, features with high noise and features that
were detected in a low number of samples are removed. As a side effect, by
reducing the number of features that are being tested later, the pre-filtering
reduces also the loss of power by the subsequent adjustment for multiple
hypothesis testing.

This step is fundamental, though one must be careful not to pre-filter for a
characteristic that will be tested later: the pre-filtering must be, as a matter
of fact, independent from later analyses.

The first step of pre-filtering consists of removing features with high
technical variance; several methods have been developed to determine which
signals must be removed, the most common of which relies on the **relative
standard deviation (RSD)**, which is defined as the ratio between the standard
deviation and the mean:

$RSD = \dfrac{s_{i,qc}}{\bar{m}_{i,qc}}$

This value is calculated for each feature found in the pooled QC samples: when
this is higher than 30%, the feature is removed from the dataset
[@broadhurstGuidelinesConsiderationsUse2018].

Another common approach is based on the **dispersion ratio (D-ratio)**: this is
defined as the ratio between the sample standard deviation for the pooled QC
samples and the sample standard deviation for the study samples (the former
expected to represent technical variance, the latter a combination of technical
and biological variance):

$D-ratio = \dfrac{s_{i,qc}}{s_{i,sample}}$

The interpretation of this value goes as follows: when the D-ratio is 0%, there
is no technical variance in the observed measurements, whereas a D-ratio of
100% represents only noise and no biological variance detected. A common cut-off
for the D-ratio is 0.5, aiming at keeping features whose variation in study
samples is twice as large as the one in QC samples
[@broadhurstGuidelinesConsiderationsUse2018]. 

For the present dataset calculate the D-ratio **separately** for each source
(sample matrix) because the variance between sample matrices is expected to be
very large. For each feature the mean D-ratio across the 3 sample matrices is
used for the filtering.


```{r filter-rsd, warning = FALSE}
rsds <- rowRsd(assay(res_qc, "normalized_filled"))

dratios_ven <- apply(
    log2(assay(res_qc, "normalized_filled")), 1, sd, na.rm = TRUE) /
    apply(log2(assay(res_pos[, res_pos$source == "venous"],
                     "normalized_filled")), 1, sd, na.rm = TRUE)

dratios_cap <- apply(
    log2(assay(res_qc, "normalized_filled")), 1, sd, na.rm = TRUE) /
    apply(log2(assay(res_pos[, res_pos$source == "capillary"],
                     "normalized_filled")), 1, sd, na.rm = TRUE)

dratios_pla <- apply(
    log2(assay(res_qc, "normalized_filled")), 1, sd, na.rm = TRUE) /
    apply(log2(assay(res_pos[, res_pos$source == "plasma"],
                     "normalized_filled")), 1, sd, na.rm = TRUE)

dratios <- apply(cbind(dratios_ven, dratios_cap, dratios_pla),
                 MARGIN = 1, mean, na.rm = TRUE)
dratios[is.infinite(dratios)] <- NA
```

The distribution of RSD values and D-ratio is shown in the plot below:

```{r filter-rsd-plot, fig.path = IMAGE_PATH, fig.width = 5 * phi, fig.height = 5, fig.cap = "Distribution of RSD values and D-ratios in the data set. The dashed vertical red line represents the cut-off value for the RSD and D-ratio, respectively.", echo = FALSE}
par(mfrow = c(1, 2))
plot(density(rsds, na.rm = TRUE), xlab = "RSD",
     main = "Distribution of RSD values")
abline(v = 0.3, col = "red", lty = 2)
plot(density(dratios, na.rm = TRUE), xlab = "D-ratio",
     main = "Distribution of D-ratios")
abline(v = 0.5, col = "red", lty = 2)
```

The plot below directly compares the RSD and D-ratio for each feature.

```{r filter-rsd-vs-dratio-plot, fig.path = IMAGE_PATH, fig.width = 5, fig.height = 5, fig.cap = "Direct comparison of RSD and D-ratios.", echo = FALSE}
plot(log2(rsds), log2(dratios), xlab = expression(log[2]~RSD),
     ylab = expression(log[2]~D-ratio), pch = 16, col = "#00000040")
abline(v = log2(0.3), col = "red", lty = 2)
abline(h = log2(0.5), col = "red", lty = 2)
```

The plot shows a correlation between RSD and D-ratios, though the two methods
are not interchangeable. Below we pre-filter the data using the D-ratio.

```{r do-filter}
res <- res_pos[which(dratios < 0.5), ]
```

This reduced the data set from `r length(dratios)` to `r nrow(res)` features.
Next, we discard the features that have not been identified in at least one
third of the samples in any of the sample groups.

```{r filter-proportion}
keep <- moreAreValidThan(assay(res, "raw"), f = res$source, prop = 1/3)
res <- res[keep, ]
```

The dataset has been reduced from `r length(rsds)` to `r nrow(res)` features:
this result shows that most features have been retained even after
pre-filtering, thus ensuring a dataset where features have a D-ratio lower than
0.5 and have less than 70% missing values.


```{r, echo = FALSE}
## Variability across data sets.
matrices <- vapply(strsplit(colnames(M), split = ".", fixed = TRUE),
                   function(z) z[1], character(1))
png(paste0(FIG_PATH, "FIG_8_POS_untargeted_diff_rep_A.png"),
    width = 10, height = 5, res = 600, pointsize = 4, units = "cm")
library(vioplot)
par(mfrow = c(1, 2))
m <- split(colMeans(M), matrices)
vioplot(m, main = "All features", las = 2,
        ylab = expression(log[2]~average~difference))
grid(nx = NA, ny = NULL)
##
m <- split(colMeans(M[rownames(res), ]), matrices)
vioplot(m, main = "Features passed pre-filtering", las = 2,
        ylab = expression(log[2]~average~difference))
grid(nx = NA, ny = NULL)
dev.off()

## The same plot, but average per feature.
png(paste0(FIG_PATH, "FIG_8_POS_untargeted_diff_rep_B.png"),
    width = 10, height = 5, res = 600, pointsize = 4, units = "cm")
par(mfrow = c(1, 2))
m <- t(M)
m <- split.data.frame(m, matrices)
m <- lapply(m, colMeans)
vioplot(m, main = "All features", las = 2, ylab = expression(log[2]~difference))
grid(nx = NA, ny = NULL)
m <- t(M[rownames(res), ])
m <- split.data.frame(m, matrices)
m <- lapply(m, colMeans)
vioplot(m, main = "Features passed pre-filtering", las = 2,
        ylab = expression(log[2]~difference))
grid(nx = NA, ny = NULL)
dev.off()

```


# Exploratory analysis: PCA

Next, we perform a PCA analysis: this allows us to gather information about any
possible similarities among the samples, based on the measured metabolite
intensities.

```{r pca-all}
pc <- prcomp(t(log2(assay(res, "normalized_filled_imputed"))),
             center = TRUE, scale. = FALSE)
```

```{r pca-plot, fig.path = IMAGE_PATH, fig.cap = "PCA of the samples based on feature intensities.", fig.width = 7 * phi, fig.height = 7, echo = FALSE}
par(mfrow = c(1, 2))
col <- "#00000080"
plot_pca(pc, bg = paste0(col_source[as.character(res$source)], 90),
         pc_x = 1, pc_y = 2, pch = 21, col = col)
legend("topright", col = col_source, legend = names(col_source),
       pch = 16, ncol = 2)
plot_pca(pc, bg = paste0(col_source[as.character(res$source)], 90),
         pc_x = 3, pc_y = 4, pch = 21, col = col)

```

```{r, echo = FALSE}
png(paste0(FIG_PATH, "FIG_2_POS_PCA_untargeted.png"), width = 5, height = 5,
    res = 600, pointsize = 4, units = "cm")
par(mfrow = c(1, 1), mar = c(4.2, 4.5, 1.5, 1), cex.lab = 1.5)
plot_pca(pc, bg = paste0(col_source[as.character(res$source)], 90),
         pc_x = 1, pc_y = 2, pch = 21, col = col, cex = 1.5)
legend("topright", pt.bg = col_source, col = "#00000080",
       legend = names(col_source), pch = 21, cex = 1.5)
## plot_pca(pc, bg = paste0(col_source[as.character(res$source)], 90),
##          pc_x = 3, pc_y = 4, pch = 21, col = col, cex = 1.5)
dev.off()

```

We see a clear separation by sample matrix for venous and plasma samples on PC1 
and PC2, whereas 3 samples of the capillary blood samples are clustered apart 
from the remaining cpillary blood samples on PC2.


# Present and absent features

We then wanted to check if there are compounds that are only present in some of
the three matrices. For a metabolite to be *present* we require that a signal
was detected by the MS instrument for that metabolite in a large fraction of
samples of a certain sample matrix. We thus call a metabolite to be present, if
a chromatographic peak was detected in 1/3 of the samples per matrix. Therefore,
we use below the `raw` data of our three sample sources (which only contains
intensity values for **detected** chromatographic peaks) and looked, which
compounds are present in more than 1/3 of the samples.

```{r Present and absent features, echo = FALSE}
plas <- assay(res, "raw")[, res$source == "plasma"]
plas_present <- moreAreValidThan(plas, prop = 1/3)

ven <- assay(res, "raw")[, res$source == "venous"]
ven_present <- moreAreValidThan(ven, prop = 1/3)

cap <- assay(res, "raw")[, res$source == "capillary"]
cap_present <- moreAreValidThan(cap, prop = 1/3)

```

We then create a plot an `UpSetR` plot to see the overlap of present features:

```{r present-feature-comparison-upset, fig.path = IMAGE_PATH, fig.cap = "Overlap of present features across the 3 sample matrices.", echo = FALSE, eval = TRUE}

#create an UpSetR plot
library(UpSetR)
listInput <- list(capillary = which(cap_present, useNames = TRUE),
                  venous = which(ven_present, useNames = TRUE),
                  plasma = which(plas_present, useNames = TRUE))
upset(fromList(listInput), order.by = "freq",
      sets.bar.color = col_source[c("venous", "capillary", "plasma")])
```

```{r, echo = FALSE}
png(paste0(FIG_PATH, "FIG_2_POS_upset_untargeted.png"), width = 7, height = 7,
    res = 600, pointsize = 4, units = "cm")
par(mfrow = c(1, 1), mar = c(4.2, 4.5, 1.5, 1), cex.lab = 1.5)
upset(fromList(listInput), keep.order = FALSE,
      sets.bar.color = col_source[c("capillary", "venous", "plasma")],
      order.by = "freq")
dev.off()

```

292 of the 705 features found to be significant in at least one of the matrices 
are shared between capillary and venous samples, whereas 206 features are
shared throughout all matrices. 62 features were found in venous and plasma 
samples, 13 features were found in capillary and plasma samples. 76 features
were only present in venous samples, 43 features in capillary samples and 13
were linked to plasma sample exclusively. 

# Differential abundance analysis

In this section, we perform a differential abundance analysis to identify
features that have significantly different abundances between male and female 
samples. The analysis is based on feature-wise multiple linear regression:
the aim of such analysis is to find the relationship between the independent
variables (age and sex) and the response variable (signal intensity).
In short, multiple linear regression is a form of linear regression that is used
when there are two or more predictors.
Multiple linear regression is preferred over separate simple linear regression
in order to avoid wrong predictions: this could happen because the input
variables may be correlated, which could lead to unsatisfactory results. The
formula for multiple regression model is:

$Y = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} \ldots, + \beta_px_{ip} + \epsilon$

where: \
- $Y =$ predicted value (dependent variable) \
- $\beta_0 =$ y intercept, constant term
- $\beta_1, \beta_2, \ldots, \beta_p =$ regression coefficients \
- $x_i =$ independent variables \
- $\epsilon =$ residuals.

The `limma` package contains the `lmFit` function, which calculates the linear
model that best describes the data. The results are stored in a `MArrayLM`
(Microarray Linear Model Fit) object.

This model, though, is not enough to accept whether or not there is actually a
relationship among the response and the independent variables, therefore we must
perform a hypothesis test: we define the *null hypothesis* as there not being
any differences in the abundances of metabolites in male and female samples. The
*alternative hypothesis* is therefore defined when there are differences in the
intensities of the signals coming from the metabolites in the two different
experimental groups.
To accept or reject the alternative hypothesis, it is necessary to calculate the
p-value: the function that allows us to add the results to the `MArrayLM` object
created before is `eBayes`. The `eBayes` function computes several statistics,
including the moderated t-test, which is defined as follows:

$\dfrac{d}{s + s_0}$

where \
- $d =$ difference in two group means ($m_1 - m_2$) \
- $s =$ pooled standard deviation \
- $s_0 =$ small constant (it depends on the variance within the group).

The constant is added to the denominator in order to avoid a division by an
extremely low number, which would of course increase the result of the statistic
falsely inducing us into rejecting the null hypothesis, thus considering the
difference as significant, when it is not. When performing multiple hypothesis
testing, though, there is a high chance of rejecting the null hypothesis when it
is true (type I error), thus a method to control the False Discovery Rate is
required: in this case we opted for Benjamini-Hochberg correction. In
conclusion, the alternative hypothesis is rejected when the adjusted p-value is
smaller than the confidence threshold that was set at the beginning of this
document to `r p.cut`. This means we accept `r p.cut * 100`% false positives
among the features called *significant*.

```{r analysis}
## Factor sample source, sex and age
source <- factor(res$source)
sex <- factor(res$sex)

## Fit the data to the desired design
dsgn <- model.matrix(~ 0 + source + sex)
fit <- lmFit(log2(assay(res, "normalized_filled_imputed")), design = dsgn)

## Fit the actual contrasts of interest
contr_mat <- makeContrasts(
  CapvsVen = sourcecapillary - sourcevenous,
  CapvsPlas = sourcecapillary - sourceplasma,
  VenvsPlas = sourcevenous - sourceplasma,
  levels = dsgn)
fit <- contrasts.fit(fit, contrasts = contr_mat)
fit <- eBayes(fit)
adjp <- apply(fit$p.value, 2, p.adjust, method = "BH")
tmp <- data.frame(
    coef = fit$coefficient,
    pvalue = fit$p.value,
    adjp = adjp,
    significant = adjp < p.cut & abs(fit$coefficient) > m.cut
)
tmp$RSD <- rsds[rownames(tmp)]
tmp$Dratio <- dratios[rownames(tmp)]
tmp$avg.Cap <- rowMeans(
    log2(assay(res, "normalized_filled_imputed")[, res$source == "capillary"]))
tmp$avg.Ven <- rowMeans(
    log2(assay(res, "normalized_filled_imputed")[, res$source == "venous"]))
tmp$avg.Plas <- rowMeans(
    log2(assay(res, "normalized_filled_imputed")[, res$source == "plasma"]))
rowData(res) <- cbind(rowData(res), tmp)
```

We plot then the distribution of p-values, both raw and adjusted:

```{r standards-p-value-histogram, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 7 * phi, fig.height = 7, fig.cap = "Distribution of raw (left) and adjusted p-values (right) for all 3 comparisons."}
par(mfrow = c(3, 2))
hist(rowData(res)$pvalue.CapvsPlas, breaks = 64, xlab = "p value",
     main = "Capillary vs plasma")
hist(rowData(res)$adjp.CapvsPlas, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "Capillary vs plasma")
hist(rowData(res)$pvalue.VenvsPlas, breaks = 64, xlab = "p value",
     main = "Venous vs plasma")
hist(rowData(res)$adjp.VenvsPlas, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "Venous vs plasma")
hist(rowData(res)$pvalue.CapvsVen, breaks = 64, xlab = "p value",
     main = "Capillary vs venous")
hist(rowData(res)$adjp.CapvsVen, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "Capillary vs venous")
```

Thus, most of the metabolites were identified to have significantly different
concentrations between the comparison. A table with the number of significant
metabolites is shown below.

```{r table-sig, echo = FALSE, results = "asis"}
tab <- colSums(as.matrix(rowData(res)[, grep("significant", 
                                             colnames(rowData(res)))]))
pandoc.table(tab, style = "rmarkdown",
             caption = paste0("Number of significant features of the in",
                              " total ", nrow(res), " analyzed features."))
```


### Volcano plots

We first evaluate features found to be significantly different between two 
matrices visually by generating volcano plots:

```{r volcano, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 10, fig.height = 10, fig.cap = "Volcano plot showing the analysis results for the comparison between capillary and venous samples (left), capillary and plasma samples (mid) and venous and plasma samples (right)."}
plot_volcano <- function(x, contrast = y, main = "", xlim, ylim) {
    X <- rowData(x)[, paste0("coef.", contrast)]
    Y <- rowData(x)[, paste0("adjp.", contrast)]
    minp <- min(Y[Y > 0])
    Y[Y == 0] <- minp / 100
    Y <- -log10(Y)
    if (missing(xlim))
        xlim <- range(X)
    if (missing(ylim))
        ylim <- range(Y)
    plot(X, Y,
         xlab = expression(log[2]~difference),
         ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
         xlim = xlim, ylim = ylim, main = main)
    rect(xleft = -100, ybottom = -log10(p.cut), xright = -m.cut, ytop = 100,
         border = NA, col = paste0(brewer.pal(3, "Set1")[2], 10))
    rect(xleft = m.cut, ybottom = -log10(p.cut), xright = 100, ytop = 100,
         border = NA, col = paste0(brewer.pal(3, "Set1")[2], 10))
        sigs <- rowData(x)[, paste0("significant.", contrast)]
    if (any(sigs))
        points(X[sigs], Y[sigs], col = paste0(brewer.pal(3, "Set1")[2], "ce"))
}
par(mfrow = c(1, 3))
XL <- c(-15, 15)
YL <- c(0, 60)
plot_volcano(res, "CapvsVen", xlim = XL, ylim = YL,
             main = "Capillary vs Venous")
grid()
plot_volcano(res, "CapvsPlas", xlim = XL, ylim = YL,
             main = "Capillary vs Plasma")
grid()
plot_volcano(res, "VenvsPlas", xlim = XL, ylim = YL,
             main = "Venous vs Plasma")
grid()
```

```{r, echo = FALSE}
png(paste0(FIG_PATH, "FIG_POS_volcano_untargeted.png"),
    width = 12, height = 4, units = "cm", res = 600, pointsize = 6)
par(mfrow = c(1, 3), mar = c(4, 4.2, 1.5, 0.5))
XL <- c(-15, 15)
YL <- c(0, 60)
plot_volcano(res, "CapvsVen", xlim = XL, ylim = YL,
             main = "Capillary vs Venous")
grid()
plot_volcano(res, "CapvsPlas", xlim = XL, ylim = YL,
             main = "Capillary vs Plasma")
grid()
plot_volcano(res, "VenvsPlas", xlim = XL, ylim = YL,
             main = "Venous vs Plasma")
grid()
dev.off()

```


### UpSetR plot of the significant features

We then compare the features found to be significant in the 1 vs 1 matrix 
comparison across all three analyses by creating an `UpSetR` plot:

```{r overlap-significant-ft-upset, fig.path = IMAGE_PATH, fig.cap = "Overlap of significant features for the comparison between capillary and venous samples, capillary and plasma samples and venous and plasma samples.", echo = FALSE, eval = TRUE}
listInput <- list(CapvsVen = which(rowData(res)$significant.CapvsVen, 
                                   useNames = TRUE),
                  CapvsPlas = which(rowData(res)$significant.CapvsPlas, 
                                    useNames = TRUE),
                  VenvsPlas = which(rowData(res)$significant.VenvsPlas, 
                                    useNames = TRUE))
upset(fromList(listInput), order.by = "freq")
```


### Heatmap

We want to plot a heatmap for the features found to be significant in at least 
one matrix comparison.

```{r heatmap-subset, echo = FALSE}

sig_CV <- which(rowData(res)$significant.CapvsVen, useNames = TRUE)
sig_CP <- which(rowData(res)$significant.CapvsPlas, useNames = TRUE)
sig_VP <- which(rowData(res)$significant.VenvsPlas, useNames = TRUE)

sig_ft <- unique(c(sig_CV, sig_CP, sig_VP))

```

Therefore, we generate a heatmap of individual log2 abundances for these 
features:

```{r heatmap-log2, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 10, fig.cap = "Heatmap of log2 abundances for significant features.", echo = FALSE}
## Create heatmap
tmp <- log2(assay(res, "normalized_filled_imputed")[sig_ft, , drop = FALSE])
tmp <- tmp - rowMeans(tmp, na.rm = TRUE)
ann <- data.frame(source = colData(res)[, c("source")])
rownames(ann) <- colnames(tmp)
pm <- pheatmap(tmp, annotation_col = ann, labels_col = res$sample,
               annotation_color = list(source = col_source),
               labels_row = rep("", nrow(tmp)))
```

```{r, echo = FALSE}
pheatmap(tmp, annotation_col = ann, labels_col = res$sample,
         annotation_color = list(source = col_source),
         labels_row = rep("", nrow(tmp)),
         filename = paste0(FIG_PATH, "FIG_POS_heatmap_untargeted.png"),
         width = 7, height = 10)

```


The data is separated into two main clusters, one containing all plasma samples
and the other containing all capillary and venous samples which indicates a
higher similarity between the latter ones.  In the heatmap we observe a cluster
of features that are highly abundant in capillary samples, but lower in plasma
and/or in venous blood in the upper part, whereas in the lower part of the map a
cluster of features, that are low in intensity in capillary blood and higher in
intensity in plasma and venous blood.


# Difference between male and female participants

```{r}
all_present <- plas_present & cap_present & ven_present
```

In order to evaluate whether the 3 matrices would yield similar biological
results, we compare the metabolic profiles of male against female participants
in each sample matrix and determine whether the differences are similar. To
ensure comparability, we restrict the analysis to features defined to be present
in each of the 3 sample matrices. This reduced the data set to 
`r sum(all_present)` features.

We next perform a differential abundance analysis (male *vs* female) separately
for each sample matrix.

```{r diff-abundance-capillary}
## Restrict to capillary samples
res_cap <- res[all_present, res$source == "capillary"]

## Define the linear model and perform the analysis
sex <- factor(res_cap$sex)
dsgn <- model.matrix(~ 0 + sex)
fit <- lmFit(log2(assay(res_cap, "normalized_filled_imputed")), design = dsgn)

## Fit the actual contrasts of interest
contr_mat <- makeContrasts(
  MvsF = sexM - sexF,
  levels = dsgn)
fit <- contrasts.fit(fit, contrasts = contr_mat)
fit <- eBayes(fit)
adjp <- apply(fit$p.value, 2, p.adjust, method = "BH")
tmp <- data.frame(
    coef.MvsF = fit$coefficient[, "MvsF"],
    pvalue.MvsF = fit$p.value[, "MvsF"],
    adjp.MvsF = p.adjust(fit$p.value[, "MvsF"], method = "BH")
)
tmp$significant.MvsF <- tmp$adjp.MvsF < p.cut & abs(tmp$coef.MvsF) > m.cut

tmp$avg.M <- rowMeans(
    log2(assay(res_cap, "normalized_filled_imputed")[, res_cap$sex == "M"]))
tmp$avg.F <- rowMeans(
    log2(assay(res_cap, "normalized_filled_imputed")[, res_cap$sex == "F"]))
rowData(res_cap) <- cbind(rowData(res_cap), tmp)
```

The same analysis is performed for venous and plasma samples.

```{r diff-abundance-others, echo = FALSE}
## venous samples
res_ven <- res[all_present, res$source == "venous"]

## Define the linear model and perform the analysis
sex <- factor(res_ven$sex)
dsgn <- model.matrix(~ 0 + sex)
fit <- lmFit(log2(assay(res_ven, "normalized_filled_imputed")), design = dsgn)

## Fit the actual contrasts of interest
contr_mat <- makeContrasts(
  MvsF = sexM - sexF,
  levels = dsgn)
fit <- contrasts.fit(fit, contrasts = contr_mat)
fit <- eBayes(fit)
adjp <- apply(fit$p.value, 2, p.adjust, method = "BH")
tmp <- data.frame(
    coef.MvsF = fit$coefficient[, "MvsF"],
    pvalue.MvsF = fit$p.value[, "MvsF"],
    adjp.MvsF = p.adjust(fit$p.value[, "MvsF"], method = "BH")
)
tmp$significant.MvsF <- tmp$adjp.MvsF < p.cut & abs(tmp$coef.MvsF) > m.cut

tmp$avg.M <- rowMeans(
    log2(assay(res_ven, "normalized_filled_imputed")[, res_ven$sex == "M"]))
tmp$avg.F <- rowMeans(
    log2(assay(res_ven, "normalized_filled_imputed")[, res_ven$sex == "F"]))
rowData(res_ven) <- cbind(rowData(res_ven), tmp)

## plasma samples
res_plas <- res[all_present, res$source == "plasma"]

## Define the linear model and perform the analysis
sex <- factor(res_plas$sex)
dsgn <- model.matrix(~ 0 + sex)
fit <- lmFit(log2(assay(res_plas, "normalized_filled_imputed")), design = dsgn)

## Fit the actual contrasts of interest
contr_mat <- makeContrasts(
  MvsF = sexM - sexF,
  levels = dsgn)
fit <- contrasts.fit(fit, contrasts = contr_mat)
fit <- eBayes(fit)
adjp <- apply(fit$p.value, 2, p.adjust, method = "BH")
tmp <- data.frame(
    coef.MvsF = fit$coefficient[, "MvsF"],
    pvalue.MvsF = fit$p.value[, "MvsF"],
    adjp.MvsF = p.adjust(fit$p.value[, "MvsF"], method = "BH")
)
tmp$significant.MvsF <- tmp$adjp.MvsF < p.cut & abs(tmp$coef.MvsF) > m.cut

tmp$avg.M <- rowMeans(
    log2(assay(res_plas, "normalized_filled_imputed")[, res_plas$sex == "M"]))
tmp$avg.F <- rowMeans(
    log2(assay(res_plas, "normalized_filled_imputed")[, res_plas$sex == "F"]))
rowData(res_plas) <- cbind(rowData(res_plas), tmp)
```

The results from the analyses are shown in the volcano plots below.

```{r volcano-m-vs-f, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 4, fig.cap = "Volcano plot showing the results for the comparison between male and female participants in capillary (left), venous (middle) and plasma (right) samples."}
par(mfrow = c(1, 3))
XL <- c(-3, 3)
YL <- c(0, 2)
plot_volcano(res_cap, "MvsF", xlim = XL, ylim = YL,
             main = "Capillary")
grid()
plot_volcano(res_ven, "MvsF", xlim = XL, ylim = YL,
             main = "Venous")
grid()
plot_volcano(res_plas, "MvsF", xlim = XL, ylim = YL,
             main = "Plasma")
grid()
```


```{r, echo = FALSE}
png(paste0(FIG_PATH, "FIG_POS_sex_volcano_untargeted.png"),
    width = 12, height = 4, units = "cm", res = 600, pointsize = 6)
par(mfrow = c(1, 3), mar = c(4, 4.2, 1.5, 0.5))
XL <- c(-3, 3)
YL <- c(0, 2)
plot_volcano(res_cap, "MvsF", xlim = XL, ylim = YL,
             main = "Capillary")
grid()
plot_volcano(res_ven, "MvsF", xlim = XL, ylim = YL,
             main = "Venous")
grid()
plot_volcano(res_plas, "MvsF", xlim = XL, ylim = YL,
             main = "Plasma")
grid()
tmp <- dev.off()
```

In addition we directly compare the log2 fold change (M) values to evaluate
whether results would be comparable.

```{r sex-comparison-M, fig.width = 8, fig.height = 4, fig.path = IMAGE_PATH, fig.cap = "Comparison of log2 fold change values representing differences of feature abundances in male against female participants between different sample matrices.", echo = FALSE}
cor_spear <- numeric()

par(mfrow = c(1, 2))
X <- rowData(res_cap)$coef.MvsF
Y <- rowData(res_ven)$coef.MvsF
L <- lm(Y ~ X)
lims <- range(c(X, Y))
plot(X, Y, xlab = expression(M[capillary]), ylab = expression(M[venous]),
     main = "", pch = 21, col = "#00000080",
     bg = "#00000040", cex = 1.3, xlim = lims, ylim = lims)
grid()
abline(0, 1, col = "#00000020", lty = 3)
cor_spear <- c(cor_spear, cor(X, Y, method = "spearman"))

X <- rowData(res_cap)$coef.MvsF
Y <- rowData(res_plas)$coef.MvsF
L <- lm(Y ~ X)
lims <- range(c(X, Y))
plot(X, Y, xlab = expression(M[capillary]), ylab = expression(M[plasma]),
     main = "", pch = 21, col = "#00000080",
     bg = "#00000040", cex = 1.3, xlim = lims, ylim = lims)
grid()
abline(0, 1, col = "#00000020", lty = 3)
cor_spear <- c(cor_spear, cor(X, Y, method = "spearman"))

png(paste0(FIG_PATH, "FIG_6_POS_comparison-sex-M.png"),
    width = 8, height = 4, units = "cm", res = 600, pointsize = 4)
par(mfrow = c(1, 2), mar = c(4, 4.4, 1, 0.6), cex.lab = 1.5)
X <- rowData(res_cap)$coef.MvsF
Y <- rowData(res_ven)$coef.MvsF
L <- lm(Y ~ X)
lims <- range(c(X, Y))
plot(X, Y, xlab = expression(M[capillary]), ylab = expression(M[venous]),
     main = "", pch = 21, col = "#00000080",
     bg = "#00000040", cex = 1.3, xlim = lims, ylim = lims,
     xaxt = "n", yaxt = "n")
xat <- axTicks(1, usr = par("usr")[1:2])
labs <- gsub("-", "\U2212", print.default(xat))
axis(1, at = xat, labels = labs)
yat <- axTicks(2, usr = par("usr")[1:2])
labs <- gsub("-", "\U2212", print.default(yat))
axis(2, at = yat, labels = labs)
grid()
abline(0, 1, col = "#00000020", lty = 3)

X <- rowData(res_cap)$coef.MvsF
Y <- rowData(res_plas)$coef.MvsF
L <- lm(Y ~ X)
lims <- range(c(X, Y))
plot(X, Y, xlab = expression(M[capillary]), ylab = expression(M[plasma]),
     main = "", pch = 21, col = "#00000080",
     bg = "#00000040", cex = 1.3, xlim = lims, ylim = lims,
     xaxt = "n", yaxt = "n")
xat <- axTicks(1, usr = par("usr")[1:2])
labs <- gsub("-", "\U2212", print.default(xat))
axis(1, at = xat, labels = labs)
yat <- axTicks(2, usr = par("usr")[1:2])
labs <- gsub("-", "\U2212", print.default(yat))
axis(2, at = yat, labels = labs)
grid()
abline(0, 1, col = "#00000020", lty = 3)
tmp <- dev.off()

```

The correlation of the log2 fold change values is shown below.

```{r, echo = FALSE, results = "asis"}
tab <- data.frame(spearman_rho = cor_spear)
rownames(tab) <- c("Capillary vs venous", "Capillary vs plasma")
pandoc.table(tab, style = "rmarkdown",
             caption = paste0("Correlation (Spearman's rho) of M values ",
                              "representing differences in abundances of male",
                              " against female participants between different",
                              " matrices."))
```

# Session information

```{r}
sessionInfo()
```
