---
title: "Matrix comparison - Pos"
author: "Søren Fjelstrup, Chiara Volani, Giuseppe Paglia and Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results='hide', message = FALSE}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_matrix_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
#' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

#' Define paths:
filename <- "vams_matrix_pos"
#' Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH)) 
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

#' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction

In this document we evaluate differences between the 4 matrices (sample sources)
*venous*, *capillary*, *plasma* and *RBC*. We are specifically interested in
identifying features which are specific to one matrix or common to all matrices
(qualitative analysis) as well as features with different abundances between
matrices (quantitative analyses). For the former analysis we use determine for
each feature whether it is *present* in a sample or not based on whether a
chromatographic peak was identified for it. The latter analysis is based on the
feature abundances.

The analysis uses the normalized feature abundances generated in
[vams_normalization.Rmd](vams_normalization.Rmd). The normalized data contains
abundance estimates for all features that are present in at least 30% of samples
of a matrix (QC samples excluded). The abundances from the two technical
replicates of each sample were combined into a single feature abundance. Also,
flags have been defined that provide information about the *quality* of a
normalized feature concentration. It is based on whether a chromatographic peak
was identified in a sample and how similar the reported abundances for the
feature in the two (technical) replicates for each sample are. Based on their
flag value, features can be classified into 6 different types:

- `1`: signal from 2 detected peaks present and the RSD of their abundances is <
  30%. The `mean` of the two is reported.
- `0.9`: 1 detected and one filled-in signal present and their RSD is < 30%. The
  `mean` of the two is reported.
- `0.8`: only filled-in signal, but their RSD is < 30%. The `mean` of the two is
  reported.
- `0.7`: signal for two detected peaks but their RSD is >= 30%. The `mean` of
  the two is reported.
- `0.5`: signal from 1 detected peak present (independently of whether a
  filled-in signal is present for the other replicate). Report only
  the value for the detected peak.
- `0.25`: only filled-in signal is present. Either mean or single value
  reported.

Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data , results='hide', message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
#' register(SerialParam())
library(SummarizedExperiment)
library(UpSetR)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization/vams_pos.RData")

```

# Data pre-filtering

The loaded data set consists of `r nrow(assay(vams_pos))` features that have
multiple peaks per samples in less than 10% of samples in which a peak was found
and that are present in at least 30% of samples from one matrix. We next
pre-filter this data set (further) by removing features with an RSD higher than
30% across QC samples (sample pools). The only concern with such a pre-filtering
on QC samples is that, in the present data set, the QC, which is in fact a
sample pool, seems to not be completely representative for every matrix.

Prior pre-filtering we plot the distribution of the RSD values across QC
samples.

```{r qc-rsd-distribution, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Distribution of RSD values across QC samples. The grey vertical line indicates the 30% cut-off.", fig.width = 6, fig.height = 6}
plot(density(rowData(vams_pos)$QC_rsd, na.rm = TRUE),
     xlab = expression(RSD[QC]))
abline(v = 0.3, col = "grey")

```

This 30% RSD filtering will remove a considerable amount of samples.

```{r pre-filter}
keep <- rowData(vams_pos)$QC_rsd < 0.3
vams_pos <- vams_pos[which(keep), ]
```

Pre-filtering reduced the data set from `r length(keep)` features to 
`r sum(keep, na.rm = TRUE)` (`r sum(is.na(keep))` had an RSD of `NA` and were 
also excluded.


# *Qualtitative* data analysis

## Identifying detected/present features for each matrix

We next aim to identify *present* (detected) features for each matrix which
allows to perform qualitative comparisons between the matrices. In contrast to
features with differences in abundances (identified in the next section) we do
not consider absolute intensities in this analysis but rather evaluate whether a
feature is detected in most samples of a matrix or not, i.e. whether a
chromatographic peak was identified or not. This information is encoded into th
*weight* of a measurement (see Introduction for the definition of the weights):
a chromatographic peak was detected for a feature in at least one of the two
replicates if the weight is either 1, 0.9, 0.7 or 0.5. Below we use these
weights to determine the proportion of samples for each matrix in which a
chromatographic peak was identified.

```{r proportion-detected, message = FALSE, warning = FALSE}
wght_detected <- c(1, 0.9, 0.7, 0.5)

#' Determine for each feature the proportion of samples (per matrix) in
#' which a peak was detected/
prop_detected <- apply(assay(vams_pos, "weights"), 1, function(x, grps) {
    vapply(split(x, grps), function(z) sum(z %in% wght_detected) / length(z),
    numeric(1))}, grps = vams_pos$source)

prop_detected <- t(prop_detected)
```

Defining absent/present calls is not trivial. Using a hard cut-off on the
proportion of samples with a detected peak can be problematic, as features can
not be expected to be present in all samples. Sex-specific features would for
example be expected to be present in only 50% of samples. Requiring features to
be present in 60% of samples to define them as *present* would thus flag all
sex-specific features to be *absent*.

We therefore use an approach that compares, separately for each feature, and in
each matrix the number of samples in which a chromatographic peak was detected
to the maximum such number in any of the matrices. Specifically, features are
called *absent* in a certain matrix, if a peak was detected in less than one
third of the maximal number of samples with a detected peak in any of the
matrices. As example, if a feature was detected in 10 samples of a matrix, but
only in 3 in another matrix, it will be called *absent* in the latter. On the
other hand, if a feature was detected at most in 4 samples in a matrix, it is
still considered to be present in matrices in which e.g. in 2 samples a peak was
detected.

Below we calculate first the ratio between the proportion of samples in which a
peak was detected and the maximal proportion for that feature and subsequently
define the absent/present calls for each feature in each matrix based on these
values.

Note that the present analysis bases on features with a detected peak in at
least 1/3 of samples.

```{r present-absent-call, message = FALSE, warning = FALSE}
#' Calculate ratio between proportion and max proportion
prop_ratio <- (1 / rowMax(prop_detected)) * prop_detected

#' Define present/absent calls
present_call <- prop_ratio > 1/3
```


## Comparison of numbers of features detected in each matrix

With the present/absent calls available we can next evaluate the number of
features being detected in each matrix. The results are represented with the
barplot below.

```{r barplot-features, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Number of features detected by each matrix", fig.width = 6, fig.height = 8}
par(mfrow = c(1, 1), mar = c(7, 4.5, 2, 0.5))
barplot(colSums(present_call), col = col_source[colnames(present_call)],
        main = "Detected features", ylab = "count", las = 2)
```

Not unexpectedly, most features are present in venous samples followed by
capillary and RBC samples. The fewest are present in plasma samples. The numbers
are shown also in the table below.

```{r feature-count-table, results = "asis", echo = FALSE}
cpt <- paste0("Number of features detected in each matrix")
pandoc.table(colSums(present_call), style = "rmarkdown",
             caption = cpt)
```


## Matrix-specific features and features present in all matrices

We next compare the overlaps and intersections of features being *present* in
the 4 matrices.

```{r upsetr-plot, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Visualisation of the feature intersects between the 4 matrices.", fig.width = 10, fig.height = 8}
ints <- apply(present_call, MARGIN = 2, which)
upset(fromList(ints), order.by = "freq",
      sets.bar.color = col_source[c("venous", "capillary", "RBC", "plasma")])
```

The largest overlap with 1,301 features is between venous, capillary and RBC
cells. About one fourth of the ~ 4,000 features is detectable in all 4
matrices. The overlap for all other combinations is much smaller. The proportion
of matrix-specific features is also very low (128, 112, 85 and 30 for venous,
plasma, capillary and RBC, respectively).

Next we define the features present in either all matrices, or are specific for
a single matrix.

```{r matrix-specific-features}
feats_common <- rownames(present_call)[rowSums(present_call) == 4]
feats_plasma <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "plasma"]]
feats_venous <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "venous"]]
feats_RBC <- rownames(present_call)[rowSums(present_call) == 1 &
                                    present_call[, "RBC"]]
feats_capillary <- rownames(present_call)[rowSums(present_call) == 1 &
                                          present_call[, "capillary"]]
```

```{r nacount, echo = FALSE}
na_count <- apply(assay(vams_pos), MARGIN = 1, function(z) sum(is.na(z)))
```

About one fourth of features still have (even after filling-in missing peaks)
missing values in some few samples (average number of missing values is 
`r mean(na_count)`. We replace these with half of the feature's smallest value 
in all samples adding a small standard deviation thus avoiding constant values.

```{r Imputation-of-Values}
assays(vams_pos)$imputed <- imputeRowMinRand(
                    assay(vams_pos), min_fraction = 1/2,
                    sd_fraction = 1, abs = TRUE)

```

We generate also extracted ion chromatograms for some randomly selected features
common to all samples or specific for one matrix.

```{r plot-common-features, echo = FALSE, warning = FALSE, eval = TRUE}
load("data/RData/vams_normalization/data_pos.RData")
col_sample <- col_source[data_pos$source]

plot_feats <- function(x, col) {
    col_bg <- paste0(col[chromPeaks(x)[, "sample"]], 20)
    col_fg <- paste0(col[chromPeaks(x)[, "sample"]], 60)
    col_bg[grep("^NA", col_bg)] <- NA
    col_fg[is.na(col_bg)] <- NA
    col_smpl <- paste0(col, 30)
    col_smpl[grep("^NA", col_smpl)] <- NA
    plot(x, col = col_smpl, peakBg = col_bg, peakCol = col_fg)
}

#' Present in all
dr <- paste0(IMAGE_PATH, "common_features/")
dir.create(dr, showWarnings = FALSE)
feats <- sample(feats_common, 20)
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- feats[1]

#' Plasma
dr <- paste0(IMAGE_PATH, "plasma_features/")
dir.create(dr, showWarnings = FALSE)
#' most specific:
feats <- feats_plasma[order(rowMeans(prop_ratio[feats_plasma, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

#' Venous
dr <- paste0(IMAGE_PATH, "venous_features/")
dir.create(dr, showWarnings = FALSE)
feats <- feats_venous[order(rowMeans(prop_ratio[feats_venous, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

#' RBC
dr <- paste0(IMAGE_PATH, "RBC_features/")
dir.create(dr, showWarnings = FALSE)
feats <- feats_RBC[order(rowMeans(prop_ratio[feats_RBC, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

#' Capillary
dr <- paste0(IMAGE_PATH, "capillary_features/")
dir.create(dr, showWarnings = FALSE)
feats <- feats_capillary[order(rowMeans(prop_ratio[feats_capillary, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

```

Some example data for features are shown below. The first shows an example for a
*common* feature, i.e. a feature that is present in all 4 matrices.

```{r out.width = "750px", echo = FALSE, fig.cap = "Feature present in all 4 matrices."}
knitr::include_graphics(paste0(IMAGE_PATH, "common_features/",
                               feats_example[1], ".png"))
```

Below an example for a plasma-only feature is shown. There seems to be some
signal also for some samples from other matrices, but that is mostly below
detection limit.

```{r out.width = "750px", echo = FALSE, fig.cap = "Plasma-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "plasma_features/",
                               feats_example[2], ".png"))
```

An example for a venous-specific feature is shown below.

```{r out.width = "750px", echo = FALSE, fig.cap = "Venous-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "venous_features/",
                               feats_example[3], ".png"))
```

An example for a RBC-only feature is shown below.

```{r out.width = "750px", echo = FALSE, fig.cap = "RBC-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "RBC_features/",
                               feats_example[4], ".png"))
```

An example for a capillary-specific feature is shown below.

```{r out.width = "750px", echo = FALSE, fig.cap = "Capillary-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "capillary_features/",
                               feats_example[5], ".png"))
```

What seems to be common to all matrix-specific features is that their abundance
is moderate. Below we thus evaluate whether the average abundance level of
matrix-specific features is different to those from common features.

```{r matrix-specific-abundance, message = FALSE, fig.path = IMAGE_PATH, fig.cap = "Distribution of average (across samples) abundances (only detected peaks) of common and matrix-specific features (i.e. features detected **only** in a specific matrix).", fig.width = 6, fig.height = 8}
ints <- list(
    common = rowMeans(assay(vams_pos)[feats_common, ]),
    plasma = rowMeans(assay(vams_pos)[feats_plasma,
                                      vams_pos$source == "plasma"]),
    venous = rowMeans(assay(vams_pos)[feats_venous,
                                      vams_pos$source == "venous"]),
    RBC = rowMeans(assay(vams_pos)[feats_RBC,
                                   vams_pos$source == "RBC"]),
    capillary = rowMeans(assay(vams_pos)[feats_capillary,
                                         vams_pos$source == "capillary"])
)
par(mar = c(7, 4.5, 1, 1))
boxplot(lapply(ints, log2), varwidth = TRUE, col = col_source[names(ints)],
        ylab = expression(log[2]~abundance), las = 2, main = "average abundance")
grid(nx = NA, ny = NULL)
```

Indeed, the average abundance of matrix-specific features seems to be about
two-fold lower than the abundance of *common* features. Thus many features might
appear to be matrix specific because their abundance is below the detection
limit in other matrices.

# *Quantitative* data analysis

## Grouping samples based on metabolite abundances

We perform a principal component analysis to evaluate the grouping of the
samples/matrices in our experiment.

```{r pca-rsd, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 10, fig.height = 5, fig.cap = "PCA grouping samples based on (centered) feature abundances. Circles represent samples from female, rectangles from male participants. Missing values were replaced with half of the feature's minimal abundance adding also a small random variance."}
tmp <- log2(assay(vams_pos, "imputed"))
pc <- prcomp(t(tmp), scale = FALSE, center = TRUE)
pchs <- ifelse(vams_pos$sex == "F", 21, 22)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[vams_pos$source], "ce"),
         bg = paste0(col_source[vams_pos$source], 60), pch = pchs, cex = 1.5)
legend("bottomright", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[vams_pos$source], 80),
         bg = paste0(col_source[vams_pos$source], 40), pch = pchs, cex = 1.5)

```

As expected, a clear separation of plasma samples from all other samples
(containing also blood cells) can be observed on PC1. Capillary and red blood
cell samples cluster close together, with some RBC samples being closest to the
cluster consisting of venous samples. Both show however a stronger variation
(especially in PC2) than plasma or venous samples. On PC3 we can see again a
separation by matrix, with plasma and venous samples being closest. The
separation on PC3 could reflect the different sampling source, i.e. sampling
from the finger tip with the Mitra device (capillary) *vs* sampling from
intravenous blood (venous, plasma and RBC).

Despite data normalization, it could be possible that differences seen in the
PCA are due to overall concentration differences. To evaluate this each of the
sample types were plotted to see if differences in overall concentrations are
present. In contrast to the previous plot comparing average abundances is that
here we consider all detected features in each matrix (before only
matrix-specific features were considered).

```{r BoxplotOfAbundances, fig.path = IMAGE_PATH, fig.cap = "Distribution of average abundances of common features and features detected within a certain matrix.", echo = FALSE}

#' All feautures present in the respective sample types are found. Note this is
#' different to when only the unique were found.
feats_plasma <- rownames(present_call)[present_call[, "plasma"]]
feats_venous <- rownames(present_call)[present_call[, "venous"]]
feats_RBC <- rownames(present_call)[present_call[, "RBC"]]
feats_capillary <- rownames(present_call)[present_call[, "capillary"]]

#' A list containing the abundances of the respective features were made. 
#' Note that they are imputed as it otherwise have too many missing values
intsMean <- list(
    common = rowMeans(assay(vams_pos, "imputed")[feats_common, ]),
    plasma = rowMeans(assay(vams_pos, "imputed")[feats_plasma,
                                                 vams_pos$source == "plasma"]),
    venous = rowMeans(assay(vams_pos, "imputed")[feats_venous,
                                                 vams_pos$source == "venous"]),
    RBC = rowMeans(assay(vams_pos, "imputed")[feats_RBC,
                                              vams_pos$source == "RBC"]),
    capillary = rowMeans(assay(vams_pos, "imputed")[feats_capillary,
                                                    vams_pos$source == "capillary"])
)
  
#' Boxplot showing the distributions of the different groups
par(mar = c(7, 4.5, 1, 1))
boxplot(lapply(intsMean, log2), varwidth = TRUE, col = col_source[names(intsMean)],
        ylab = expression(log[2]~abundance), las = 2, main = "average abundance")
```

With the exception of common features that show slightly higher average
abundances, the average abundances of features detected in a matrix are
comparable. In addition to boxplots we create also cumulative distribution
function to evaluate whether the distribution of average feature abundances are
comparable.

```{r CumulativeDistributionFunctionsOfAbundances, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Cumulative density function of each matrix. The distributions of the matrices are mostly identical, with the plasma having a sligthly higher overall abundance and the RBC a slightly higher proportion of low abundance features"}
#' Cumulative distribution functions are made for the different groups, 
#' log2 is necessary due to the large differences in abundances
plasma_ecdf <- ecdf(as.numeric(log2(intsMean$plasma)))
venous_ecdf <- ecdf(as.numeric(log2(intsMean$venous))) 
RBC_ecdf <- ecdf(as.numeric(log2(intsMean$RBC)))
capillary_ecdf <- ecdf(as.numeric(log2(intsMean$capillary)))

plot(plasma_ecdf, col = col_source["plasma"], main = "cdf of abundance",
     xlab = "log2 abundance", ylab = "Density" , xlim = c(0,20), ylim = (c(0,1)))
lines(venous_ecdf, col = col_source["venous"])
lines(RBC_ecdf, col = col_source["RBC"])
lines(capillary_ecdf, col = col_source["capillary"])

legend('left', c("plasma","venous","RBC","capillary"), fill = col_source[
  c("plasma","venous","RBC","capillary")], border=NA)
```

The cumulative density function shows that the abundances of the features for
the RBC is the lowest,and the plasma is the highest, the venous and capillary
are almost identical. Over and above, on average, abundances of all features are
however comparable between samples and matrix (which would also be expected
because the data was normalized to level between-sample differences).


## Differential abundance analysis

In this analysis we fit the regression model to log2 transformed abundances
(including imputed values) of each feature and extract the coefficients from the
fitted model representing the slope (or difference) of the average abundance of
samples from the respective matrix to the base line capillary samples. The
p-value represents the significance from the Wald test against the null
hypothesis of the coefficient being 0 (i.e. no differences between the average
abundances).

Capillary samples are used as baseline, comparing thus abundances of all other
matrices against these.

The p-values are subsequently adjusted for multiple hypothesis testing using the
method from Benjamini and Hochberg for a strong control of the false discovery
rate.

The employed model `conc ~ source + sex` adjusts for sex-specific differences. A
simple model without accounting for any confounding effects yielded however
similar results (not shown).

```{r ModelforlinearmodelVSplasmanoconfounding, warning = FALSE, echo = FALSE, eval = FALSE}
## Old version - kept for historical reasons
SourcesTypes <- c("sourceplasma","sourceRBC","sourcevenous")
pLimit <- 0.01
foldLimit <- 1

#' The function for the multiple linear regression is made. This takes a numeric
#' input along with a matrix containing the confounding factors, it returns the
#' means and the p value
fitlmFunction <- function(x, dataMat, model = conc ~ source) {
    dataMat$conc <- x
    dataMat$source <- factor(x = dataMat$source, 
                             levels = c("capillary", "plasma", "RBC", "venous"))
    mdl <- lm(model, data = dataMat)
    smry <- data.frame(summary(mdl)$coefficients)
    SourceTypes <- paste0("source", levels(dataMat$source)[-1])
    Output <- c(coef = smry[SourceTypes, 1], pvalue =  smry[SourceTypes, 4])
    names(Output) <- c(paste0("coef.", SourceTypes),
                       paste0("pvalue.", SourceTypes))
    Output
}

#' The data is log2 transformed to make it mostly gaussian in distribution  
fitMatrix <- t(apply(log2(assay(vams_pos, "imputed")), 1, fitlmFunction, 
                     dataMat = as.data.frame(colData(vams_pos)),
                     model = conc ~ source + sex))

#' Adjusting for multiple hypothesis testing
p.adjusted <- apply(fitMatrix[ , paste0("pvalue.", SourcesTypes)], 2, p.adjust, 
                    method = "BH")

#' Define "significant" features
sign <- p.adjusted < pLimit &
    abs(fitMatrix[, grep("^coef", colnames(fitMatrix))]) > foldLimit
colnames(sign) <- sub("pvalue.", "", colnames(sign), fixed = TRUE)

#' Compiling the final result table
res_matrix_pos <- cbind(rowData(vams_pos), fitMatrix,
                        adj = p.adjusted, significant = sign,
                        present = present_call)

```

Linear models are fit using the `limma` package and p-values are estimated with
the empirical Bayes-based moderated t-test.

```{r limma, message = FALSE}
library(limma)
SourcesTypes <- c("sourceplasma","sourceRBC","sourcevenous")
pLimit <- 0.01
foldLimit <- 1

source <- vams_pos$source
sex <- vams_pos$sex
dsgn <- model.matrix(~ source + sex)
fit <- lmFit(log2(assay(vams_pos, "imputed")), design = dsgn)
fit <- eBayes(fit)

res_matrix_pos <- data.frame(coef = fit$coefficients[, -1],
                             pvalue = fit$p.value[, -1])
adjp <- apply(fit$p.value[, -1], MARGIN = 2, p.adjust, method = "BH")
sign <- adjp < pLimit & abs(fit$coefficients[, -1]) > foldLimit
res_matrix_pos <- cbind(as.data.frame(rowData(vams_pos)),
                        res_matrix_pos,
                        adj.pvalue = adjp,
                        significant = sign,
                        present = present_call)

```

Histograms of each of the 3 matrixes against capillary is made.

```{r Histogramofpvalues, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values each of the linear models against capillary. All histograms show a smooth decrease over the range of pvalues, indicating only a small possible effect from confounding factors.", fig.width = 12, fig.height = 4}
TestedSourcenames <-  c("plasma", "RBC", "venous")
par(mfrow=c(1, 3))
for (i in 1:3) {
    hist(res_matrix_pos[, paste0("pvalue.", SourcesTypes[i])],
         main = paste0(TestedSourcenames[i], " vs capillary"),
         xlab = "p-value")
}


```

It looks like the p-values decrease gradually, this suggest that the effect of
any confounding effects are likely to be minor, compared to the overall
differences between the matrices. We do however have a very large number of
features with differences in their abundances.

Based on the models created in the previous section volcano plots are made for
each of the matrices compared to capillary.

```{r VolcanoPlotMultipleRegression, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Volcano plots of each of the linear models against capillary. Benjamini-Hochberg adjusted values, against the mean differences for each feature, show a large proportion of the features are statistically significantly different between matrixes", fig.width = 12, fig.height = 4}

#' A new color set is defined to better conform to what is need by the volcano plot
col_volcano <- brewer.pal(9, name = "Set1")[c(9, 9, 9, 3)]
names(col_volcano) <- c("all", "pLimit", "foldLimit", "significant")

#' The filters for pvalue and fold increase is defined
colorFilterLowP <- as.matrix(
    res_matrix_pos[, paste0("adj.pvalue.", SourcesTypes)]) < pLimit
colorFilterHighFold <- abs(as.matrix(
    res_matrix_pos[, paste0("coef.", SourcesTypes)])) > foldLimit

#' The filter for coloring the points is made
colorFilter <- 1 * colorFilterLowP
colorFilter[colorFilterHighFold] <- 2
colorFilter[colorFilterHighFold * colorFilterLowP == 1] <- 3
colorFilter <- colorFilter + 1

par(mfrow=c(1, 3), mar = c(4.3, 4.3, 1, 1))
x_range <- range(as.matrix(
    res_matrix_pos[, grep("^coef", colnames(res_matrix_pos))]), na.rm = TRUE)
y_range <- range(
    -log10(as.matrix(res_matrix_pos[, grep("^adj", colnames(res_matrix_pos))])),
    na.rm = TRUE)
for (i in 1:3){
    plot(3, 3, pch = NA, xlim = x_range, ylim = y_range,
         xlab = expression(log[2]~fold~change),
         ylab = expression(-log[10]~p[BH]~value),
         main = paste0(TestedSourcenames[i], " vs capillary"))
    grid(col = "#cecece20")
    points(res_matrix_pos[, paste0("coef.", SourcesTypes[i])],
           -log10(res_matrix_pos[, paste0("adj.pvalue.", SourcesTypes[i])]),
           pch = 16, col = paste0(col_volcano[colorFilter[, i]], 60),
           cex = 0.6)
    abline(h = -log10(pLimit))
    abline(v = c(-foldLimit, foldLimit))
}

```

A surprisingly large amount of features were detected to have statistically
different abundances between the compared matrices. This was expected for the
RBC and plasma, but not for the venous one. The number of significant features
for each comparison are listed below.

```{r count-table_linear_model, results = "asis", echo = FALSE}
cpt <- paste0("Number of features with a significant difference in abundances",
              " (more than two-fold difference in abundances at a 1% FDR.")
pandoc.table(colSums(as.matrix(
    res_matrix_pos[, grep("significant", colnames(res_matrix_pos))]),
    na.rm = TRUE), style = "rmarkdown", caption = cpt)

```

As expected, the largest number of features with differences in abundances was
identified between plasma and capillary samples. The number of significant
features is much smaller for the comparisons between RBC and capillary and for
venous and capillary samples.

The large differences may have many sources. To help identify these it was
decided to probe the underlying distributions of the features based on the two
two factors we know. The mass/charge ratio and the retention time. It would be
expected that the venous and capillary would be mostly the same with the plasma
and RBC being different.

```{r pValueDependencyOnmzNoconfounding, fig.cap = "Smoothed histograms showing the distribution of m/z values (left) or retention time (right) for detected features in a matrix.", echo = FALSE, fig.width = 12, fig.height = 6, fig.path = IMAGE_PATH}
#' Distributions of mz values for each matrix.
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
hsts <- apply(present_call, MARGIN = 2, function(z) {
    hist(res_matrix_pos$mzmed[z], plot = FALSE, breaks = 128)
})
ymax <- max(vapply(hsts, function(z) max(z$counts), numeric(1)))
plot(3, 3, pch = NA, ylab = "counts", xlab = "m/z", ylim = c(0, ymax),
     xlim = range(res_matrix_pos$mzmed),
     main = "Detected features in a matrix")
legend("topleft", col = col_source, legend = names(col_source), lwd = 2)
for (i in 1:ncol(present_call)) {
    col <- paste0(col_source[colnames(present_call)[i]], 80)
    ## points(hsts[[i]]$mids, hsts[[i]]$counts, type = "l", col = col)
    points(lowess(hsts[[i]]$mids, hsts[[i]]$counts, f = 0.1),
           type = "l", lwd = 2, col = col)
}
#' retention time
hsts <- apply(present_call, MARGIN = 2, function(z) {
    hist(res_matrix_pos$rtmed[z], plot = FALSE, breaks = 128)
})
ymax <- max(vapply(hsts, function(z) max(z$counts), numeric(1)))
plot(3, 3, pch = NA, ylab = "counts", xlab = "retention time",
     ylim = c(0, ymax), xlim = range(res_matrix_pos$rtmed),
     main = "Detected features in a matrix")
for (i in 1:ncol(present_call)) {
    col <- paste0(col_source[colnames(present_call)[i]], 80)
    ## points(hsts[[i]]$mids, hsts[[i]]$counts, type = "l", col = col)
    points(lowess(hsts[[i]]$mids, hsts[[i]]$counts, f = 0.05),
           type = "l", lwd = 2, col = col)
}

```

The distribution of m/z values for detected features is similar across all
matrices, with the exception of plasma samples that show a lower proportion of
features with high mass-to-charge ratios with an m/z between 790-100. Fatty
acids such as diacyl-glycerols could potentially fall into this category.

The distribution of retention times for detected features is also highly
comparable between the matrices, again with the exception of plasma samples,
that seem to completely miss features with a retention time between 200 and 210
seconds while the other 3 matrices have the same number of detected features in
that time window. Interestingly, no differences between matrices can be seen for
highly non-polar (hydrophobic) features (with retention times around 30
seconds). This argues against the theory of fatty acids being missing in plasma
samples. Plasma samples seem to lack relatively highly polar compounds with with
a large molecular weight. 

Next we evaluate the distribution of m/z and retention times for the features
with significant differences in abundances compared to capillary samples.

```{r mzSignificant, fig.cap = "Smoothed histograms showing the distributions of m/z values (left) and retention times (right) for features with significant different abundances compared to the levels in capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6, fig.path = IMAGE_PATH}
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
sources <- c("plasma", "RBC", "venous")
#' m/z
hsts <- lapply(res_matrix_pos[, paste0("significant.source", sources)],
               function(z) hist(res_matrix_pos$mzmed[z], plot = FALSE,
                                breaks = 128))
ymax <- max(vapply(hsts, function(z) max(z$counts), numeric(1)))
plot(3, 3, pch = NA, ylab = "counts", xlab = "m/z", ylim = c(0, ymax),
     xlim = range(res_matrix_pos$mzmed),
     main = "Significant features against capillary")
legend("topleft", col = col_source[sources], legend = sources, lwd = 2)
for (i in 1:length(sources)) {
    col <- paste0(col_source[sources[i]], 80)
    ## points(hsts[[i]]$mids, hsts[[i]]$counts, type = "l", col = col)
    points(lowess(hsts[[i]]$mids, hsts[[i]]$counts, f = 0.1),
           type = "l", lwd = 2, col = col)
}
#' retention time
hsts <- lapply(res_matrix_pos[, paste0("significant.source", sources)],
               function(z) hist(res_matrix_pos$rtmed[z], plot = FALSE,
                                breaks = 128))
ymax <- max(vapply(hsts, function(z) max(z$counts), numeric(1)))
plot(3, 3, pch = NA, ylab = "counts", xlab = "retention time",
     ylim = c(0, ymax), xlim = range(res_matrix_pos$rtmed),
     main = "Significant features against capillary")
for (i in 1:length(sources)) {
    col <- paste0(col_source[sources[i]], 80)
    ## points(hsts[[i]]$mids, hsts[[i]]$counts, type = "l", col = col)
    points(lowess(hsts[[i]]$mids, hsts[[i]]$counts, f = 0.04),
           type = "l", lwd = 2, col = col)
}

```

The plot above shows distributions of features that are significantly different
between a matrix and capillary samples. The shape of the distribution of the m/z
values for these features are comparable, while the magnitude differs
(reflecting the differences in numbers of significant features for the
individual comparisons). The most notable difference is that a higher proportion
of features found significantly different between plasma and capillary samples
seems to be present in the m/z range from 600 to 800.

The distribution of significant features along the retention time dimension
differs considerably. A higher number of significant features was found for the
comparison between RBC and capillary samples at a retention time of about 160
seconds. Features with significant abundances in the comparison between plasma
and capillary samples are enriched in the higher retention time range,
especially between 200 and 220 seconds.

Next we evaluate also the direction of the difference in the matrix comparison,
i.e. consider feature with higher or lower abundances in a matrix compared to
plasma samples separately.

```{r mzCountsSignificant, fig.cap = "Smoothed histograms showing the distribution of m/z values (left) and retention time (right) for significant features with a higher or lower abundance compared to capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6, fig.path = IMAGE_PATH}

par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1.5, 1))
plot(x = 1, y = 1, pch = NA, xlim = c(50, 1100), ylim = c(-30, 30),
     xlab = "m/z", ylab = "count", main = "Significant features")
text(x = c(50, 50), y = c(-25, 25), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] < 0), "mzmed"]
    yLow <- hist(yLow, breaks = seq(50, 1100, 5), plot = FALSE)   
    yHigh <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] > 0), "mzmed"]
    yHigh <- hist(yHigh, breaks = seq(50, 1100, 5), plot = FALSE)
    ## lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    ## lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
    points(lowess(yHigh$mids, yHigh$counts, f = 0.05),
           col = paste0(col_source[src], 80), lwd = 2, type = "l")
    points(lowess(yLow$mids, -yLow$counts, f = 0.05),
           col = paste0(col_source[src], 80), lwd = 2, type = "l")
}
grid()
abline(h = 0)
legend("topright", legend = TestedSourcenames, lwd = 2, lty = 1,
       col = col_source[TestedSourcenames], inset = c(0, 0))
plot(x = 1, y = 1, pch = NA, xlim = c(0, 250), ylim = c(-400, 400),
     xlab = "rt", ylab = "count", main = "Significant features")
text(x = c(0, 0), y = c(-350, 350), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] < 0), "rtmed"]
    yLow <- hist(yLow, breaks = seq(0, 250, 5), plot = FALSE)   
    yHigh <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] > 0), "rtmed"]
    yHigh <- hist(yHigh, breaks = seq(0, 250, 5), plot = FALSE)
    ## lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    ## lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
    points(lowess(yHigh$mids, yHigh$counts, f = 0.1),
           col = paste0(col_source[src], 80), lwd = 2, type = "l")
    points(lowess(yLow$mids, -yLow$counts, f = 0.1),
           col = paste0(col_source[src], 80), lwd = 2, type = "l")
}
grid()
abline(h = 0)

```
 
As expected the plasma sample is characterized by having lower abundance for
features specifically in the higher m/z range.

The distribution of retention times, representing the polarity, for significant
features is very informative too (right plot in the figure above). The largest
differences can again be spotted in the high retention time range where we would
expect compounds with a higher polarity. At a retention time of about 160
seconds we find a large proportion of features with significantly lower
abundances in RBC samples, while only fewer significant features were found for
the comparisons of plasma and venous samples against capillary samples. For
these, an enrichment of significant features with higher abundances can be seen
at a retention time of about 170 seconds. Most of the significant features
between plasma and capillary samples have lower abundances in plasma samples and
retention times between 160 and 220 seconds.

# Result object

The results of this analysis are exported to the *vams_matrix_pos_results.txt*
file.

```{r writexl}
library(writexl)
write_xlsx(as.data.frame(res_matrix_pos),
           path = "data/xls/vams_matrix_pos_results.xlsx")

save(res_matrix_pos, file = paste0(RDATA_PATH, "res_matrix_pos.RData"))
```

The most important columns in the result table are:
- `mzmed`, `mzmin`, `mzmax`: m/z of the feature.
- `rtmed`, `rtmin`, `rtmax`: retention time window of the feature.
- `QC_rsd`: RSD over QC samples.
- `coef.source<matrix>`: log2 fold change of feature abundance between
  `<matrix>` and capillary samples.
- `pvalue.source<matrix>`: p-value representing the significance of the
  difference in abundances between `<matrix>` and capillary samples. These are
  raw (unadjusted) p-values and should only considered for sorting of the
  features.
- `adj.pvalue.source<matrix>`: p-values adjusted for multiple hypothesis
  testing representing the significance of differential abundance.
- `significant.source<matrix>`: logical indicating whether the feature was
  identified to have a significant difference in abundances.
- `present.<matrix>`: *present* call of the feature from the qualitative
  analysis.


# Session information

```{r sessionInfo}
devtools::session_info()
```

# References
