---
title: "Matrix comparison"
author: "Johannes Rainer, Chiara Volani and Giuseppe Paglia"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results = "asis"}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_matrix_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE}
##' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

##' Define paths:
filename <- "vams_matrix_pos"
##' Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
##' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

##' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction

In this document we evaluate differences between the 4 matrices (sample sources)
*venous*, *capillary*, *plasma* and *RBC*. We are specifically interested in
identifying features which are specific to one matrix or common to all matrices
(qualitative analysis) as well as features with different abundances between
matrices (quantitative analyses). For the former analysis we use determine for
each feature whether it is *present* in a sample or not based on whether a
chromatographic peak was identified for it. The latter analysis is based on the
feature abundances.

The analysis uses the normalized feature abundances generated in
[vams_normalization.Rmd](vams_normalization.Rmd). The normalized data contains
abundance estimates for all features that are present in at least 30% of samples
of a matrix (QC samples excluded). The abundances from the two technical
replicates of each sample were combined into a single feature abundance. Also,
weights have been defined that provide information about the *quality* of a
normalized feature concentration. It is based on whether a chromatographic peak
was identified in a sample and how similar the reported abundances for the
feature in the two (technical) replicates for each sample are. Based on their
weight, features can be classified into 6 different types:

- `1`: signal from 2 detected peaks present and the RSD of their abundances is <
  30%. The `mean` of the two is reported.
- `0.9`: 1 detected and one filled-in signal present and their RSD is < 30%. The
  `mean` of the two is reported.
- `0.8`: only filled-in signal, but their RSD is < 30%. The `mean` of the two is
  reported.
- `0.7`: signal for two detected peaks but their RSD is >= 30%. The `mean` of
  the two is reported.
- `0.5`: signal from 1 detected peak present (independently of whether a
  filled-in signal is present for the other replicate). Report only
  the value for the detected peak.
- `0.25`: only filled-in signal is present. Either mean or single value
  reported.

Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
library(SummarizedExperiment)
library(UpSetR)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization/vams_pos.RData")

#' Remove rows/features that have only missing values
vams_pos <- vams_pos[rowNaProp(assay(vams_pos, "abundances")) < 1, ]
```


## Grouping samples based on metabolite abundances

We perform a principal component analysis to evaluate the grouping of the
samples/matrices in our experiment.

```{r pca-rsd, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 10, fig.height = 10, fig.cap = "PCA grouping samples based on (centered) feature abundances. Only features with an RSD < 30% in QC samples were used. Missing values were replaced with half of the feature's minimal abundance adding also a small random variance."}
#' RSD < 30% in QC samples
tmp <- log2(assay(vams_pos, "abundances")[which(rowData(vams_pos)$QC_rsd < 0.3), ])
na_prop <- rowNaProp(tmp)
tmp <- imputeRowMinRand(tmp, min_fraction = 1/2)
pc <- prcomp(t(tmp), scale = FALSE, center = TRUE)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[vams_pos$source], "ce"),
         bg = paste0(col_source[vams_pos$source], 60), pch = 21, cex = 1.5)
legend("bottomright", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[vams_pos$source], 80),
         bg = paste0(col_source[vams_pos$source], 40), pch = 21, cex = 1.5)

## plot_pca(pc, pc_x = 2, pc_y = 3, col = paste0(col_source[vams_pos$source], 80),
##          bg = paste0(col_source[vams_pos$source], 40), pch = 21, cex = 1.5)

```

The PCA above was performed on all features that have an RSD < 30% in QC samples
(about 4000). Missing values were replaced with half of the feature's smallest
abundance adding also a small random value based on the features' standard
deviations. Note that excluding features with `NA` values (~ 1000) or performing
the PCA on all over 10,000 features yielded the same result (data not shown).

As expected, a clear separation of plasma samples from all other samples
(containing also blood cells) can be observed on PC1. Capillary and red blood
cell samples cluster close together, with some RBC samples being closest to the
cluster consisting of venous samples. Both show however a stronger variation
(especially in PC2) than plasma or venous samples. On PC3 we can see again a
separation by matrix, with plasma and venous samples being closest. The
separation on PC3 could reflect the different sampling source, i.e. sampling
from the finger tip with the Mitra device (capillary) *vs* sampling from
intravenous blood (venous, plasma and RBC).

## Identifying detected/present features for each matrix

We next aim to identify *present* (detected) features for each matrix which
allows to perform qualitative comparisons between the matrices. In contrast to
features with differences in abundances (identified in the next section) we do
not consider absolute intensities in this analysis but rather evaluate whether a
feature is detected in most samples of a matrix or not, i.e. whether a
chromatographic peak was identified or not. This information is encoded into the
*weight* of a measurement (see Introduction for the definition of the weights):
a chromatographic peak was detected for a feature in at least one of the two
replicates if the weight is either 1, 0.9, 0.7 or 0.5. Below we use these
weights to determine the proportion of samples for each matrix in which a
chromatographic peak was identified.

```{r proportion-detected, message = FALSE, warning = FALSE}
wght_detected <- c(1, 0.9, 0.7, 0.5)

#' Determine for each feature the proportion of samples (per matrix) in
#' which a peak was detected/
prop_detected <- apply(assay(vams_pos, "weights"), 1, function(x, grps) {
    vapply(split(x, grps), function(z) sum(z %in% wght_detected) / length(z),
           numeric(1))
}, grps = vams_pos$source)
prop_detected <- t(prop_detected)
```

Defining absent/present calls is not trivial. Using a hard cut-off on the
proportion of samples with a detected peak can be problematic, as features can
not be expected to be present in all samples. Sex-specific features would for
example be expected to be present in only 50% of samples. Requiring features to
be present in 60% of samples to define them as *present* would thus flag all
sex-specific features to be *absent*.

We therefore use an approach that compares, separately for each feature, and in
each matrix the number of samples in which a chromatographic peak was detected
to the maximum such number in any of the matrices. Specifically, features are
called *absent* in a certain matrix, if a peak was detected in less than one
third of the maximal number of samples with a detected peak in any of the
matrices. As example, if a feature was detected in 10 samples of a matrix, but
only in 3 in another matrix, it will be called *absent* in the latter. On the
other hand, if a feature was detected at most in 4 samples in a matrix, it is
still considered to be present in matrices in which e.g. in 2 samples a peak was
detected.

Below we calculate first the ratio between the proportion of samples in which a
peak was detected and the maximal proportion for that feature and subsequently
define the absent/present calls for each feature in each matrix based on these
values.

Note that the present analysis bases on features with a detected peak in at
least 1/3 of samples.

```{r present-absent-call, message = FALSE, warning = FALSE}
#' Calculate ratio between proportion and max proportion
prop_ratio <- (1 / rowMax(prop_detected)) * prop_detected

#' Define present/absent calls
present_call <- prop_ratio > 1/3
```


## Matrix with the largest number of detected features

With the present/absent calls available we can next evaluate the number of
features being detected in each matrix. The results are represented with the
barplot below.

```{r barplot-features, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Number of features detected by each matrix", fig.width = 6, fig.height = 8}
par(mfrow = c(1, 1), mar = c(7, 4.5, 2, 0.5))
barplot(colSums(present_call), col = col_source[colnames(present_call)],
        main = "Detected features", ylab = "count", las = 2)
```

Not unexpectedly, most features are present in venous samples followed by
capillary and RBC samples. The fewest are present in plasma samples. The numbers
are shown also in the table below.

```{r feature-count-table, results = "asis", echo = FALSE}
cpt <- paste0("Number of features detected in each matrix")
pandoc.table(colSums(present_call), style = "rmarkdown",
             caption = cpt)
```


## Matrix-specific features and features present in all matrices

We next compare the overlaps and intersects of features being *present* in the 4
matrices.

```{r upsetr-plot, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Visualisation of the feature intersects between the 4 matrices.", fig.width = 10, fig.height = 8}
ints <- apply(present_call, MARGIN = 2, which)
upset(fromList(ints), order.by = "freq",
      sets.bar.color = col_source[c("venous", "capillary", "RBC", "plasma")])
```

About one third of the ~ 10,000 features is detectable in all 4 matrices. The
next largest overlap with ~ 2,000 features is between venous, capillary and RBC
samples followed by about 1,200 features that are specific to plasma and venous
samples. Over and above the results are as expected, with the largest overlap of
matrices being between samples containing also blood cells.


## Identifying features with differential abundance between the matrices

- Quantitative analysis.
- Use paired statistics.
- Use plasma as baseline.


## Comparison between venous and capillary blood

- 

# Session information

```{r sessionInfo}
devtools::session_info()
```

# References
