---
title: "Matrix comparison - Pos"
author: "Søren Fjelstrup, Chiara Volani, Giuseppe Paglia and Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results='hide', message = FALSE}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_matrix_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
#' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

#' Define paths:
filename <- "vams_matrix_pos"
#' Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH)) 
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

#' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction

In this document we evaluate differences between the 4 matrices (sample sources)
*venous*, *capillary*, *plasma* and *RBC*. We are specifically interested in
identifying features which are specific to one matrix or common to all matrices
(qualitative analysis) as well as features with different abundances between
matrices (quantitative analyses). For the former analysis we use determine for
each feature whether it is *present* in a sample or not based on whether a
chromatographic peak was identified for it. The latter analysis is based on the
feature abundances.

The analysis uses the normalized feature abundances generated in
[vams_normalization.Rmd](vams_normalization.Rmd). The normalized data contains
abundance estimates for all features that are present in at least 30% of samples
of a matrix (QC samples excluded). The abundances from the two technical
replicates of each sample were combined into a single feature abundance. Also,
flags have been defined that provide information about the *quality* of a
normalized feature concentration. It is based on whether a chromatographic peak
was identified in a sample and how similar the reported abundances for the
feature in the two (technical) replicates for each sample are. Based on their
flag value, features can be classified into 6 different types:

- `1`: signal from 2 detected peaks present and the RSD of their abundances is <
  30%. The `mean` of the two is reported.
- `0.9`: 1 detected and one filled-in signal present and their RSD is < 30%. The
  `mean` of the two is reported.
- `0.8`: only filled-in signal, but their RSD is < 30%. The `mean` of the two is
  reported.
- `0.7`: signal for two detected peaks but their RSD is >= 30%. The `mean` of
  the two is reported.
- `0.5`: signal from 1 detected peak present (independently of whether a
  filled-in signal is present for the other replicate). Report only
  the value for the detected peak.
- `0.25`: only filled-in signal is present. Either mean or single value
  reported.

Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data , results='hide', message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
#' register(SerialParam())
library(SummarizedExperiment)
library(UpSetR)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization/vams_pos.RData")

```


## Identifying detected/present features for each matrix

We next aim to identify *present* (detected) features for each matrix which
allows to perform qualitative comparisons between the matrices. In contrast to
features with differences in abundances (identified in the next section) we do
not consider absolute intensities in this analysis but rather evaluate whether a
feature is detected in most samples of a matrix or not, i.e. whether a
chromatographic peak was identified or not. This information is encoded into th
*weight* of a measurement (see Introduction for the definition of the weights):
a chromatographic peak was detected for a feature in at least one of the two
replicates if the weight is either 1, 0.9, 0.7 or 0.5. Below we use these
weights to determine the proportion of samples for each matrix in which a
chromatographic peak was identified.

```{r proportion-detected, message = FALSE, warning = FALSE}
wght_detected <- c(1, 0.9, 0.7, 0.5)

#' Determine for each feature the proportion of samples (per matrix) in
#' which a peak was detected/
prop_detected <- apply(assay(vams_pos, "weights"), 1, function(x, grps) {
    vapply(split(x, grps), function(z) sum(z %in% wght_detected) / length(z),
    numeric(1))}, grps = vams_pos$source)

prop_detected <- t(prop_detected)
```

Defining absent/present calls is not trivial. Using a hard cut-off on the
proportion of samples with a detected peak can be problematic, as features can
not be expected to be present in all samples. Sex-specific features would for
example be expected to be present in only 50% of samples. Requiring features to
be present in 60% of samples to define them as *present* would thus flag all
sex-specific features to be *absent*.

We therefore use an approach that compares, separately for each feature, and in
each matrix the number of samples in which a chromatographic peak was detected
to the maximum such number in any of the matrices. Specifically, features are
called *absent* in a certain matrix, if a peak was detected in less than one
third of the maximal number of samples with a detected peak in any of the
matrices. As example, if a feature was detected in 10 samples of a matrix, but
only in 3 in another matrix, it will be called *absent* in the latter. On the
other hand, if a feature was detected at most in 4 samples in a matrix, it is
still considered to be present in matrices in which e.g. in 2 samples a peak was
detected.

Below we calculate first the ratio between the proportion of samples in which a
peak was detected and the maximal proportion for that feature and subsequently
define the absent/present calls for each feature in each matrix based on these
values.

Note that the present analysis bases on features with a detected peak in at
least 1/3 of samples.

```{r present-absent-call, message = FALSE, warning = FALSE}
#' Calculate ratio between proportion and max proportion
prop_ratio <- (1 / rowMax(prop_detected)) * prop_detected

#' Define present/absent calls
present_call <- prop_ratio > 1/3
```


## Comparison of numbers of features detected in each matrix

With the present/absent calls available we can next evaluate the number of
features being detected in each matrix. The results are represented with the
barplot below.

```{r barplot-features, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Number of features detected by each matrix", fig.width = 6, fig.height = 8}
par(mfrow = c(1, 1), mar = c(7, 4.5, 2, 0.5))
barplot(colSums(present_call), col = col_source[colnames(present_call)],
        main = "Detected features", ylab = "count", las = 2)
```

Not unexpectedly, most features are present in venous samples followed by
capillary and RBC samples. The fewest are present in plasma samples. The numbers
are shown also in the table below.

```{r feature-count-table, results = "asis", echo = FALSE}
cpt <- paste0("Number of features detected in each matrix")
pandoc.table(colSums(present_call), style = "rmarkdown",
             caption = cpt)
```


## Matrix-specific features and features present in all matrices

We next compare the overlaps and intersects of features being *present* in the 4
matrices.

```{r upsetr-plot, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Visualisation of the feature intersects between the 4 matrices.", fig.width = 10, fig.height = 8}
ints <- apply(present_call, MARGIN = 2, which)
upset(fromList(ints), order.by = "freq",
      sets.bar.color = col_source[c("venous", "capillary", "RBC", "plasma")])
```

About one third of the ~ 10,000 features is detectable in all 4 matrices. The
next largest overlap with ~ 2,000 features is between venous, capillary and RBC
samples followed by about 1,200 features that are specific to plasma and venous
samples. Over and above the results are as expected, with the largest overlap of
matrices being between samples containing also blood cells. The numbers of
matrix-specific features is with ~ 200 - 600 relatively small, with plasma
samples having the largest number of unique features. Note that the overlap
of features is higher between capillary and RBC samples (657) compared to
capillary and venous samples (309).


The features present in either all matrices, or all of a specific matrix is
defined.

```{r matrix-specific-features}
feats_common <- rownames(present_call)[rowSums(present_call) == 4]
feats_plasma <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "plasma"]]
feats_venous <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "venous"]]
feats_RBC <- rownames(present_call)[rowSums(present_call) == 1 &
                                    present_call[, "RBC"]]
feats_capillary <- rownames(present_call)[rowSums(present_call) == 1 &
                                          present_call[, "capillary"]]
```

Below we impute missing feature intensities. These are replaced with half of the
feature's smallest value in all samples adding a small standard deviation
thus avoiding constant values.

```{r Imputation-of-Values}
assays(vams_pos)$Imputed <- imputeRowMinRand(
                    assay(vams_pos), min_fraction = 1/2,
                    sd_fraction = 1, abs = TRUE)

```

We generate also extracted ion chromatograms for some randomly selected features
common to all samples or specific for one matrix.

```{r plot-common-features, echo = FALSE, warning = FALSE, eval = FALSE}
load("data/RData/vams_normalization/data_pos.RData")
col_sample <- col_source[data_pos$source]

plot_feats <- function(x, col) {
    col_bg <- paste0(col[chromPeaks(x)[, "sample"]], 20)
    col_fg <- paste0(col[chromPeaks(x)[, "sample"]], 60)
    col_bg[grep("^NA", col_bg)] <- NA
    col_fg[is.na(col_bg)] <- NA
    col_smpl <- paste0(col, 30)
    col_smpl[grep("^NA", col_smpl)] <- NA
    plot(x, col = col_smpl, peakBg = col_bg, peakCol = col_fg)
}

#' Present in all
dr <- paste0(IMAGE_PATH, "common_features/")
dir.create(dr, showWarnings = FALSE)
feats <- sample(feats_common, 20)
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- feats[1]

#' Plasma
dr <- paste0(IMAGE_PATH, "plasma_features/")
dir.create(dr, showWarnings = FALSE)
#' most specific:
feats <- feats_plasma[order(rowMeans(prop_ratio[feats_plasma, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

#' Venous
dr <- paste0(IMAGE_PATH, "venous_features/")
dir.create(dr, showWarnings = FALSE)
feats <- feats_venous[order(rowMeans(prop_ratio[feats_venous, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

#' RBC
dr <- paste0(IMAGE_PATH, "RBC_features/")
dir.create(dr, showWarnings = FALSE)
feats <- feats_RBC[order(rowMeans(prop_ratio[feats_RBC, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

#' Capillary
dr <- paste0(IMAGE_PATH, "capillary_features/")
dir.create(dr, showWarnings = FALSE)
feats <- feats_capillary[order(rowMeans(prop_ratio[feats_capillary, ]))][1:20]
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
feats_example <- c(feats_example, feats[1])

```

Some example data for features are shown below. The first shows an example for a
*common* feature, i.e. a feature that is present (although at difference
abundance) in all 4 matrices.

```{r out.width = "750px", echo = FALSE, fig.cap = "Feature present in all 4 matrices."}
knitr::include_graphics(paste0(IMAGE_PATH, "common_features/",
                               feats_example[1], ".png"))
```

Below an example for a plasma-only feature is shown. There seems to be some
signal also for some samples from other matrices, but that is mostly below
detection limit.

```{r out.width = "750px", echo = FALSE, fig.cap = "Plasma-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "plasma_features/",
                               feats_example[2], ".png"))
```

An example for a venous-specific feature is shown below.

```{r out.width = "750px", echo = FALSE, fig.cap = "Venous-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "venous_features/",
                               feats_example[3], ".png"))
```

An example for a RBC-only feature is shown below.

```{r out.width = "750px", echo = FALSE, fig.cap = "RBC-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "RBC_features/",
                               feats_example[4], ".png"))
```

An example for a capillary-specific feature is shown below.

```{r out.width = "750px", echo = FALSE, fig.cap = "Capillary-specific feature."}
knitr::include_graphics(paste0(IMAGE_PATH, "capillary_features/",
                               feats_example[5], ".png"))
```

What seems to be common to all matrix-specific features is that their abundance
is moderate. Below we thus evaluate whether the average abundance level of
matrix-specific features is different to those from common features.

```{r matrix-specific-abundance, message = FALSE, fig.path = IMAGE_PATH, fig.cap = "Distribution of average (across samples) abundances of common and matrix-specific features.", fig.width = 6, fig.height = 8}
ints <- list(
    common = rowMeans(assay(vams_pos)[feats_common, ]),
    plasma = rowMeans(assay(vams_pos)[feats_plasma,
                                      vams_pos$source == "plasma"]),
    venous = rowMeans(assay(vams_pos)[feats_venous,
                                      vams_pos$source == "venous"]),
    RBC = rowMeans(assay(vams_pos)[feats_RBC,
                                   vams_pos$source == "RBC"]),
    capillary = rowMeans(assay(vams_pos)[feats_capillary,
                                         vams_pos$source == "capillary"])
)
par(mar = c(7, 4.5, 1, 1))
boxplot(lapply(ints, log2), varwidth = TRUE, col = col_source[names(ints)],
        ylab = expression(log[2]~abundance), las = 2, main = "average abundance")
grid(nx = NA, ny = NULL)
```

Indeed, the average abundance of matrix-specific features seems to be about
two-fold lower than the abundance of *common* features. Thus many features might
appear to be matrix specific because their abundance is below the detection
limit in other matrices.


## Grouping samples based on metabolite abundances

We perform a principal component analysis to evaluate the grouping of the
samples/matrices in our experiment.

```{r pca-rsd, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 10, fig.height = 5, fig.cap = "PCA grouping samples based on (centered) feature abundances. Only features with an RSD < 30% in QC samples were used. Circles represent samples from female, rectangles from male participants. Missing values were replaced with half of the feature's minimal abundance adding also a small random variance."}
#' RSD < 30% in QC samples
tmp <- log2(assay(vams_pos, "Imputed")[which(rowData(vams_pos)$QC_rsd < 0.3), ])
pc <- prcomp(t(tmp), scale = FALSE, center = TRUE)
pchs <- ifelse(vams_pos$sex == "F", 21, 22)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[vams_pos$source], "ce"),
         bg = paste0(col_source[vams_pos$source], 60), pch = pchs, cex = 1.5)
legend("bottomright", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[vams_pos$source], 80),
         bg = paste0(col_source[vams_pos$source], 40), pch = pchs, cex = 1.5)

```

The PCA above was performed on all features that have an RSD < 30% in QC samples
(about 4000). Missing values were replaced with half of the feature's smallest
abundance adding also a small random value based on the features' standard
deviations. Note that excluding features with `NA` values (~ 1000) or performing
the PCA on all over 10,000 features yielded the same result (data not shown).

As expected, a clear separation of plasma samples from all other samples
(containing also blood cells) can be observed on PC1. Capillary and red blood
cell samples cluster close together, with some RBC samples being closest to the
cluster consisting of venous samples. Both show however a stronger variation
(especially in PC2) than plasma or venous samples. On PC3 we can see again a
separation by matrix, with plasma and venous samples being closest. The
separation on PC3 could reflect the different sampling source, i.e. sampling
from the finger tip with the Mitra device (capillary) *vs* sampling from
intravenous blood (venous, plasma and RBC).


## Identifying features with differential abundance between the matrices

It is possible that the differences seen in the previous PCA plots are due to
overall concentration differences. Thus each of the sample types were plotted to 
see if significant differences in overall concentration is found.

```{r BoxplotOfAbundances, fig.path = IMAGE_PATH, fig.cap = "Distribution of average abundances of common features and features detected within a certain matrix.", echo = FALSE}

#' All feautures present in the respective sample types are found. Note this is
#' different to when only the unique were found.
feats_plasma <- rownames(present_call)[present_call[, "plasma"]]
feats_venous <- rownames(present_call)[present_call[, "venous"]]
feats_RBC <- rownames(present_call)[present_call[, "RBC"]]
feats_capillary <- rownames(present_call)[present_call[, "capillary"]]

#' A list containing the abundances of the respective features were made. 
#' Note that they are imputed as it otherwise have too many missing values
intsMean <- list(
    common =  rowMeans(assay(vams_pos, "Imputed")[feats_common, ]),
    plasma = rowMeans(assay(vams_pos, "Imputed")[feats_plasma,
                                                 vams_pos$source == "plasma"]),
    venous = rowMeans(assay(vams_pos, "Imputed")[feats_venous,
                                                 vams_pos$source == "venous"]),
    RBC = rowMeans(assay(vams_pos, "Imputed")[feats_RBC,
                                              vams_pos$source == "RBC"]),
    capillary = rowMeans(assay(vams_pos, "Imputed")[feats_capillary,
                                                    vams_pos$source == "capillary"])
)
  
#' Boxplot showing the distributions of the different groups
par(mar = c(7, 4.5, 1, 1))
boxplot(lapply(intsMean, log2), varwidth = TRUE, col = col_source[names(intsMean)],
        ylab = expression(log[2]~abundance), las = 2, main = "average abundance")
```

With the exception of common features that show slightly higher average
abundances, the average abundances of features detected in a matrix are
comparable. In addition to boxplots we create also cumulative distribution
function to evaluate whether the distribution of average feature abundances are
comparable.

```{r CumulativeDistributionFunctionsOfAbundances, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Cumulative density function of each matrix. The distributions of the matrices are mostly identical, with the plasma having a sligthly higher overall abundance and the RBC a slightly higher proportion of low abundance features"}
#' Cumulative distribution functions are made for the different groups, 
#' log2 is necessary due to the large differences in abundances
plasma_ecdf <- ecdf(as.numeric(log2(intsMean$plasma)))
venous_ecdf <- ecdf(as.numeric(log2(intsMean$venous))) 
RBC_ecdf <- ecdf(as.numeric(log2(intsMean$RBC)))
capillary_ecdf <- ecdf(as.numeric(log2(intsMean$capillary)))

plot(plasma_ecdf, col = col_source["plasma"], main = "cdf of abundance",
     xlab = "log2 abundance", ylab = "Density" , xlim = c(0,20), ylim = (c(0,1)))
lines(venous_ecdf, col = col_source["venous"])
lines(RBC_ecdf, col = col_source["RBC"])
lines(capillary_ecdf, col = col_source["capillary"])

legend('left', c("plasma","venous","RBC","capillary"), fill = col_source[
  c("plasma","venous","RBC","capillary")], border=NA)
```

The cumulative density function shows that the abundances of the features for
the RBC is the lowest,and the plasma is the highest, the venous and capillary
are almost identical.



# Unsupervised modelling

A PCA of the samples, based on the features which for all the sources are found
in at least 30% of the samples within that source. This means that all features 
specific for a matrix is excluded. Thus it is expected that the seperation will 
be much smaller than the earlier PCA plot, where matrix specific features would 
naturally be a large part of the variation seen.

Note also that this PCA is performed on the actual concentrations without log2
transformation or scaling. PCA separation will thus mostly driven by features
showing large differences.

```{r PCAOfAbundances, fig.cap = "PCA of all common features by abundance. In plot showing the 1st and 2nd dimension, 6 outliers are found. Some difference by martrix is seen as well, with the capillary being separated, this is also seen in the 3rd dimension.", echo = FALSE}

#' Note that this is done using data for which log10 has not been taken, if this is done the outliers are not found
#' A PCA is performed of the samples.
pc <- prcomp(t(assay(vams_pos, "Imputed")[feats_common, ]),
             scale = FALSE, center = TRUE)
pchs <- ifelse(vams_pos$sex == "F", 21, 22)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[vams_pos$source], "ce"),
         bg = paste0(col_source[vams_pos$source], 60), pch = pchs, cex = 1.5)
legend("topleft", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[vams_pos$source], 80),
         bg = paste0(col_source[vams_pos$source], 40), pch = pchs, cex = 1.5)


```

The first and second dimension is dominated by 6 samples.
To identify these, the samples were filtered for low values in the 1st and 2nd
dimension.

Then the features were ranked according to loading in the first dimension. 

```{r Outlier, fig.cap = "The 10  features with the most negative values in the 1st dimension of the PCA. These were chosen based on their loadings in the first dimension. Two features FT08876 and FT08384 have a much higher weigth than the rest of the features", echo = FALSE}
#' The Outliers are found and the loadings in PC1 is sorted to find 
#'  the features responsible.
OutlierFilter <- (pc$x[, 1]<(-1e+5) & pc$x[, 2] < 0)

sortedLoading <- sort(pc$rotation[,1])
plot(sortedLoading[1:10], xlab = "Rank in 1st dimension loading", 
     ylab =  "1st dimension loading weight")
#' The two outlier features are defined
feats <- c(names(sortedLoading[1:2]))
```
The samples were identified as:
 `r names(OutlierFilter)[OutlierFilter]` these seem to be clear outlier samples,
 indeed if the loadings in the first and second dimension is found it is seen
 that they have just 2 outlier features  `r feats`.


It was then decided to look at the chromatograms of the two features to
determine whether it could be found why they are so different.

```{r Chromatogram, eval = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
dr <- paste0(IMAGE_PATH, "Outlier_features/")
dir.create(dr, showWarnings = FALSE)


feats <- c("FT08876","FT08384")
feats_chrs <- featureChromatograms(data_pos, expandRt = 4, features = feats)
for (i in seq_along(feats)) {
    fn <- paste0(dr, feats[i], ".png")
    png(fn, width = 10, height = 8, units = "cm", res = 200, pointsize = 6)
    plot_feats(feats_chrs[i, ], col = col_sample)
    dev.off()
}
```

Here it is evident that for the outlier samples these features have values much
higher than the rest of the samples. It is unclear if the smaller peaks at the
lower retention times should be included here as well.

Knowing this, a new PCA is performed excluding these features.

```{r OutlierFilteredPCA, echo = FALSE, fig.cap = "PCA of abundances of all common features, with FT08876 and FT08384 removed. Only the plasma samples separate from the other samples, in the 2nd dimension"}

#' PCA without the outliers are made
commonImputedValues <- assays(vams_pos)$Imputed[feats_common,]
intsCommonNoOutliers <- commonImputedValues[!(rownames(commonImputedValues)
                                              %in% feats), ] 

pc <- prcomp(t(intsCommonNoOutliers), scale = FALSE, center = TRUE)
pchs <- ifelse(vams_pos$sex == "F", 21, 22)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[vams_pos$source], "ce"),
         bg = paste0(col_source[vams_pos$source], 60), pch = pchs, cex = 1.5)
legend("topright", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[vams_pos$source], 80),
         bg = paste0(col_source[vams_pos$source], 40), pch = pchs, cex = 1.5)



```

Removing the outliers resulted in a plot without any real discernable
differences between the groups, apart from the plasma which seperates from all
other sources, in the second dimension. No seperation were seen in other
dimensions (data not shown)


# Supervised modelling

## Linear Model

The multiple linear regression of the features is made without imputation.
The models were made using the capillary as a baseline. Since the RBC and the
plasma were made from the venous it is expected to see large differences in all 
the models made. Although it would be expected that the pure venous would be the
most alike to the capillary.

In this analysis we fit the regression model to log2 transformed abundances of
each feature and extract the coefficients from the fitted model representing the
slope (or difference) of the average abundance of samples from the respective
matrix to the base line capillary samples. The p-value represents the
significance from the Wald test against the null hypothesis of the coefficient
being 0 (i.e. no differences between the average abundances).

The p-values are subsequently adjusted for multiple hypothesis testing using the
method from Benjamini and Hochberg for a strong control of the false discovery
rate.

The employed model `conc ~ source + sex` adjusts for sex-specific differences. A
simple model without accounting for any confounding effects yielded however
similar results (not shown).

```{r ModelforlinearmodelVSplasmanoconfounding, warning = FALSE}
SourcesTypes <- c("sourceplasma","sourceRBC","sourcevenous")

#' The function for the multiple linear regression is made. This takes a numeric
#' input along with a matrix containing the confounding factors, it returns the
#' means and the p value
fitlmFunction <- function(x, dataMat, model = conc ~ source) {
    dataMat$conc <- x
    dataMat$source <- factor(x = dataMat$source, 
                             levels = c("capillary", "plasma", "RBC", "venous"))
    mdl <- lm(model, data = dataMat)
    smry <- data.frame(summary(mdl)$coefficients)
    SourceTypes <- paste0("source", levels(dataMat$source)[-1])
    Output <- c(coef = smry[SourceTypes, 1], pvalue =  smry[SourceTypes, 4])
    names(Output) <- c(paste0("coef.", SourceTypes),
                       paste0("pvalue.", SourceTypes))
    Output
}

#' The data is log2 transformed to make it mostly gaussian in distribution  
Data <- log2(assay(vams_pos, "abundances"))
fitMatrix <- t(apply(Data, 1, fitlmFunction, 
                     dataMat = as.data.frame(colData(vams_pos)),
                     model = conc ~ source + sex))

#' Adjusting for multiple hypothesis testing
p.adjusted <- apply(fitMatrix[ ,paste0("pvalue.", SourcesTypes)], 2, p.adjust, 
                    method = "BH")
#' The fitMatrix is appended to the rest of the data in vams_pos, making the
#' new dataframe res_matrix_pos
res_matrix_pos<- cbind(rowData(vams_pos), fitMatrix, adj = p.adjusted)
```

Histograms of each of the 3 matrixes against capillary is made.

```{r Histogramofpvalues, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values each of the linear models against capillary. All histograms show a smooth decrease over the range of pvalues, indicating only a small possible effect from confounding factors"}
TestedSourcenames <-  c("plasma", "RBC", "venous")
par(mfrow=c(1, 3))
for (i in 1:3) {
    hist(res_matrix_pos[, paste0("pvalue.", SourcesTypes[i])],
         main = paste0(TestedSourcenames[i], " vs capillary"),
         xlab = "p-value")
}


```

It looks like the p-values decrease gradually, this suggest that the effect of
any confounding effects are likely to be minor, compared to the overall
differences from the matrices. We do however have a very large number of
features with differences in their abundances.

Based on the models created in the previous section volcano plots are made for
each of the matrices compared to capillary.

```{r VolcanoPlotMultipleRegression, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Volcano plots of each of the linear models against capillary. Benjamini-Hochberg adjusted values, against the mean differences for each feature, show a large proportion of the features are statistically significantly different between matrixes", fig.width = 12, fig.height = 4}
#' A new color set is defined to better conform to what is need by the volcano plot
col_volcano <- brewer.pal(4, name = "Set1")
col_volcano[1] <- "#DDDDDD" #' Grey
col_volcano_Trans <- paste0(col_volcano, 80)

pLimit <- 0.01
foldLimit <- 1

#' The filters for pvalue and fold increase is defined
colorFilterLowP <- as.matrix(
    res_matrix_pos[, paste0("adj.pvalue.", SourcesTypes)]) < pLimit
colorFilterHighFold <- abs(as.matrix(
    res_matrix_pos[, paste0("coef.", SourcesTypes)])) > foldLimit

#' The filter for coloring the points is made
colorFilter <- 1 * colorFilterLowP
colorFilter[colorFilterHighFold] <- 2
colorFilter[colorFilterHighFold * colorFilterLowP == 1] <- 3
colorFilter <- colorFilter + 1

#' The significant samples (both high fold difference and low p-value) are
#' defined
sign <- colorFilterLowP & colorFilterHighFold
#' replace the "coef" in the column names with "significant"
colnames(sign) <- sub("adj.pvalue", "significant", colnames(sign))
res_matrix_pos <- cbind(res_matrix_pos, sign)

par(mfrow=c(1, 3), mar = c(4.3, 4.3, 1, 1))
for (i in 1:3){
    plot(res_matrix_pos[, paste0("coef.", SourcesTypes[i])],  cex = 0.6,
         -log10(res_matrix_pos[, paste0("adj.pvalue.", SourcesTypes[i])]),
         pch = 16, main = paste0(TestedSourcenames[i], " vs capillary"), 
         col = col_volcano_Trans[colorFilter[, i]], 
         ylab = expression(log[10]~p[BH]-value), 
         xlab = expression(log[2]~fold~change))
    abline(h = -log10(pLimit))
    abline(v = c(-foldLimit, foldLimit))
    legend("top", c("non-Significant", "lowP", "Highfold", "Significant"),
           col = col_volcano, pch = 16, bg = "white")
}


```

A surprisingly large amount of features were detected to have statistically
different abundances between the compared matrices. This was expected for the
RBC and plasma, but not for the venous one. The number of significant features
for each comparison are listed below.

```{r count-table_linear_model, results = "asis", echo = FALSE}
cpt <- paste0("Number of features with a significant difference in abundances",
              " (more than two-fold difference in abundances at a 1% FDR.")
pandoc.table(colSums(as.matrix(
    res_matrix_pos[, grep("significant", colnames(res_matrix_pos))]),
    na.rm = TRUE), style = "rmarkdown", caption = cpt)
```

As expected, the largest number of features with differences in abundances was
identified between plasma and capillary samples. The number of significant
features is much smaller for the comparisons between RBC and capillary and for
venous and capillary samples.

The large differences may have many sources. To help identify these it was
decided to probe the underlying distributions of the features based on the two
two factors we know. The mass/charge ratio and the retention time. It would be
expected that the venous and capillary would be mostly the same with the plasma
and RBC being different.

Plots showing the distributions of the matrixes based on m/z were made.

```{r pValueDependencyOnmzNoconfounding, fig.cap = "Distribution of m/z values (left) or retention time (right) for detected features in a matrix.", echo = FALSE, fig.width = 12, fig.height = 6}

#' Distributions of mz values for each matrix.
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
plot(density(res_matrix_pos$mzmed[present_call[, 1]], na.rm = TRUE),
     col = paste0(col_source[colnames(present_call)[1]], 80),
     main = "Detected features in a matrix",
     xlab = "m/z", ylim = c(0, 0.002), lwd = 2)
for (i in 2:4) {
    lines(density(res_matrix_pos$mzmed[present_call[, i]], na.rm = TRUE), 
          col = paste0(col_source[colnames(present_call)[i]], 80), lwd = 2)
}
legend("topleft", legend = c(colnames(present_call)),
       col = col_source[colnames(present_call)], pch = 16)
#' retention time
plot(density(res_matrix_pos$rtmed[present_call[, 1]], na.rm = TRUE),
     col = paste0(col_source[colnames(present_call)[1]], 80),
     main = "Detected features in a matrix",
     xlab = "rt", ylim = c(0, 0.015), lwd = 2)
for (i in 2:4) {
    lines(density(res_matrix_pos$rtmed[present_call[, i]], na.rm = TRUE), 
          col = paste0(col_source[colnames(present_call)[i]], 80), lwd = 2)
}

```

Looking at the distribution of detected features as a function of their m/z,
only the plasma is clearly different from the others. Here a higher proportion
of medium-m/z, and much fewer high-m/z are found. This is not surprising. When
making plasma high-weight compounds are removed from the matrix. The m/z
roughly corresponds to what we would expect from diacyl-glycerols. (Fatty acids)

The distributions of retention times for detected features (right plot in the
figure above) are mostly similar with some notable differences especially in the
higher retention time range: venous samples have more features in the 140-170
seconds range compared to all other matrices, but much fewer in the 170-230
seconds range. This goes against the theory that the differences seen for the
m/z is due to fatty acids. If this was the case we would expect to see a low
peak for the low retention time. The opposite is seen, features with a long
retention time, i.e. being hydrophillic, have a lower presence. Thus whatever is
missing in the plasma sample must be both large and hydrophillic.

The distributions for capillary and RBC samples are highly similar, while venous
samples are more like plasma samples in the retention time range around 160
seconds.

Next we evaluate the distribution of m/z and retention times for the features
with significant differences in abundances compared to capillary samples.

```{r mzSignificant, fig.cap = "Distributions of m/z values (left) and retention times (right) for features with significant different abundances compared to the levels in capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6}
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
#' m/z
plot(density(res_matrix_pos$mzmed, na.rm = TRUE),
     col = "#cecece", main = "Significant features",
     xlab = "m/z", ylim = c(0, 0.002), lwd = 2)
for (i in 1:3) {
    lines(density(
        res_matrix_pos$mzmed[res_matrix_pos[, paste0("significant.",
                                                     SourcesTypes[i])]],
        na.rm = TRUE), col = paste0(col_source[TestedSourcenames[i]], 80),
        lwd = 2)
}
legend("topleft", legend=c("All features", TestedSourcenames), 
       col = c("#cecece", col_source[TestedSourcenames]), pch = 16)
#' rt
plot(density(res_matrix_pos$rtmed, na.rm = TRUE),
     col = "#cecece", main = "Significant features",
     xlab = "rt", ylim = c(0, 0.04), lwd = 2)
for (i in 1:3) {
    lines(density(
        res_matrix_pos$rtmed[res_matrix_pos[, paste0("significant.",
                                                     SourcesTypes[i])]],
        na.rm = TRUE), col = paste0(col_source[TestedSourcenames[i]], 80),
        lwd = 2)
}
```

This distribution includes only the features shown to be significantly different
in the previous model. The overall shape of the density lines is mostly similar
with the one for all features with the most notable differences in the m/z range
from 600 to 800.

Features with significant different abundances between plasma and capillary
samples are enriched in this m/z region, while capillary and venous samples seem
to be similar in that range. Also interestingly, many significant features
between venous and capillary samples were found for the m/z range of ~ 250 to
450. This range corresponds roughly to where sugars and steroids are expected to
be found.

The distribution of significant features along the retention time dimension
differs considerably. The largest differences are in the high retention time
range. Specifically venous samples show a high proportion of significantly
different features (against capillary samples) with a retention of ~ 160 seconds
and, to a lower degree also of, 180 and 200 seconds. The plasma is not
suprisingly mostly different in the high range (200-220 seconds).

Next we evaluate over- and under-represented features (compared to capillary
samples) separately.

```{r mzCountsSignificant, fig.cap = "Distribution of m/z values (left) and retention time (right) for significant features with a higher or lower abundance compared to capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6}

par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1.5, 1))
plot(x = 1, y = 1, pch = NA, xlim = c(50, 1100), ylim = c(-100, 100),
     xlab = "m/z", ylab = "count", main = "Significant features")
text(x = c(50, 50), y = c(-90, 90), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] < 0), "mzmed"]
    yLow <- hist(yLow, breaks = seq(50, 1100, 5), plot = FALSE)   
    yHigh <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] > 0), "mzmed"]
    yHigh <- hist(yHigh, breaks = seq(50, 1100, 5), plot = FALSE)
    lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
}
grid()
abline(h = 0)
legend("topright", legend = TestedSourcenames, lwd = 2, lty = 1,
       col = col_source[TestedSourcenames], inset = c(0, 0))
plot(x = 1, y = 1, pch = NA, xlim = c(0, 250), ylim = c(-800, 800),
     xlab = "rt", ylab = "count", main = "Significant features")
text(x = c(0, 0), y = c(-750, 750), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] < 0), "rtmed"]
    yLow <- hist(yLow, breaks = seq(0, 250, 5), plot = FALSE)   
    yHigh <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] > 0), "rtmed"]
    yHigh <- hist(yHigh, breaks = seq(0, 250, 5), plot = FALSE)
    lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
}
grid()
abline(h = 0)

```
 
As expected the plasma sample is characterized by having lower abundance for
features in the higher m/z range. Interestingly this seems to be periodic with a
distance of about 24 Da, corresponding to about 2 residues in the acyl chain of
a fatty acid.

The distribution of retention times, representing the polarity, for significant
features is very informative too (right plot in the figure above). For the low
retention time range (from 30-50 seconds), capillary and venous samples seem to
be very much alike since only few features have significantly different
abundances between venous and capillary samples in that range. A considerable
number of features have however significantly different abundances in that range
between RBC and capillary, as well as between plasma and capillary samples. The
largest differences can again be spotted in the high retention time range. Most
features with significant differences between plasma and capillary have larger
abundances in plasma samples, in the retention time range from 160-220 with the
largest number at a retention time of about 210 seconds. Also very
interestingly, about the same number of significant features was found to have
significantly higher abundances in plasma and in venous samples compared to the
capillary samples at a retention time of about 160 (between 150 and
165). Capillary, but also RBC samples seem to have a lower sensitivity to detect
these features.


# Session information

```{r sessionInfo}
devtools::session_info()
```

# References
