---
title: "Sex comparison, negative polarity"
author: "Søren Fjelstrup, Chiara Volani, Giuseppe Paglia and Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results='hide', message = FALSE}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_sex_neg.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
#' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

#' Define paths:
filename <- "vams_sex_neg"
#' Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH)) 
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

#' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction

In this document we aim to identify compounds with significantly different
abundances between male and female participants of the study for each of the
matrices. The analysis is performed analogous to the one described in
[vams_sex_pos.Rmd](vams_sex_pos.Rmd) for positive polarity data hence we do not
repeat methodological descriptions here.

Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data , results='hide', message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
#' register(SerialParam())
library(SummarizedExperiment)
library(UpSetR)
library(limma)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization_neg/vams_neg.RData")

```

Below we impute missing values replacing them with half of the feature's
smallest reported abundance adding also a small random value. The analysis will
be performed on this imputed data set.

```{r Imputation-of-Values}
assays(vams_neg)$imputed <- imputeRowMinRand(
                    assay(vams_neg), min_fraction = 1/2,
                    sd_fraction = 1, abs = TRUE)

TestedSourcenames <-  c("capillary","plasma", "RBC", "venous")

```

# Data filtering

Optionally we could filter the data, e.g. exclude low intensity features or
similar. At present we do not pre-filter the data prior to the differential
abundance analysis.


# Differential abundance analysis

## Separate analysis for each matrix


```{r volcanoplot-function, echo = FALSE}
col_volcano <- brewer.pal(4, name = "Set1")
col_volcano[1] <- "#DDDDDD" #' Grey
col_volcano_Trans <- paste0(col_volcano, 80)

volcanoplotFunction <- function(coef, pval, pLimit, foldLimit, Title,
                                xlim = range(coef), ylim = range(-log10(pval)),
                                col_volcano = paste0(
                                    brewer.pal(9, "Set1")[c(9, 9, 9, 3)], 60)) {
    
    colorFilterLowP <- as.matrix(pval) < pLimit
    colorFilterHighFold <- abs(coef) > foldLimit
    
    colorFilter <- 1 * colorFilterLowP
    colorFilter[colorFilterHighFold] <- 2
    colorFilter[colorFilterHighFold * colorFilterLowP == 1] <- 3
    colorFilter <- colorFilter + 1

    plot(coef, -log10(pval), cex = 0.6,
         pch = 16, main = Title, 
         col = col_volcano[colorFilter], 
         ylab = expression(log[2] ~ p[BH]-value), 
         xlab = expression(log[2] ~ fold ~ change),
         xlim = xlim, ylim = ylim)
    abline(h = -log10(pLimit), col = "grey", lty = 3)
    abline(v = c(-foldLimit, foldLimit), col = "grey", lty = 3)
}

```

We use a linear model-based approach with empirical Bayes-based p-value
estimation (with Bioconductor's `limma` packages). 

```{r sex-fit-function-limma, message = FALSE, warning = FALSE}
pLimit <- 0.05
foldLimit <- 1
res_sex_neg_sep <- rowData(vams_neg)

for (src in TestedSourcenames) {
    sex <- vams_neg$sex[vams_neg$source == src]
    dsgn <- model.matrix(~ sex)
    fit <- lmFit(log2(assay(vams_neg, "imputed")[, vams_neg$source == src]),
                 design = dsgn)
    fit <- eBayes(fit)
    fitm <- cbind(fit$p.value[, "sexM"], fit$coefficients[, "sexM"],
                  p.adjust(fit$p.value[, "sexM"], method = "BH"))
    fitm <- cbind(fitm, fitm[, 3] < pLimit & abs(fitm[, 2]) > foldLimit)
    colnames(fitm) <- paste0(c("pvalue.", "coef.", "adj.pvalue.",
                               "significant."), src, ".sexM")
    res_sex_neg_sep <- cbind(res_sex_neg_sep, fitm)
}

```

Next we create histograms of the raw p-values.

```{r Histogramofpvalues, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values in each of the matrices of male vs female. The histograms generally show somewhat few significant features, especially for RBC", fig.width = 12, fig.height = 4}
par(mfrow = c(1, 4))
for (src in TestedSourcenames) {
    hist(res_sex_neg_sep[, paste0("pvalue.", src, ".sexM")],
         main = paste0("p-value of Male vs Female in ", src),
         xlab = "p-value")
}

```

Similar to the positive polarity data we see an enrichment for small p-values
only for capillary, plasma and venous samples. Given the relatively high
variance in RBC samples (also seen e.g. in PCA plots in
[vams_matrix_neg.html](vams_matrix_neg.html)) this is however not unexpected.

```{r sex-sep-volcano, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Volcano plots representing the results for the comparison between male and female individuals. Analysis was performed separately for each matrix. Horizontal and vertical dotted lines indicate the cut-off criteria for significance.", fig.width = 12, fig.height = 4}
par(mfrow = c(1, 4))
xlim <- range(as.matrix(res_sex_neg_sep[, grep("^coef", colnames(res_sex_neg_sep))]))
ylim <- range(as.matrix(res_sex_neg_sep[, grep("^adj", colnames(res_sex_neg_sep))]), pLimit)
for (src in TestedSourcenames) {
    volcanoplotFunction(
        res_sex_neg_sep[, paste0("coef.", src, ".sexM")],
        res_sex_neg_sep[, paste0("adj.pvalue.", src, ".sexM")],
        pLimit = pLimit, foldLimit = foldLimit, xlim = xlim,
        ylim = range(-log10(ylim)), Title = paste0("M vs F ", src))
}

```

Only in plasma samples we could identify features with significant differences
in abundances.

At last we compare the top 50 features with the most significant differences
between male and female for each matrix to evaluate whether they are shared
between matrices.

```{r sex-sep-top-50-pval, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Overlap of top 50 features with smallest p-values (male vs female)."}
fts_capillary <- rownames(res_sex_neg_sep)[
    order(res_sex_neg_sep$pvalue.capillary.sexM)][1:50]
fts_venous <- rownames(res_sex_neg_sep)[
    order(res_sex_neg_sep$pvalue.venous.sexM)][1:50]
fts_plasma <- rownames(res_sex_neg_sep)[
    order(res_sex_neg_sep$pvalue.plasma.sexM)][1:50]
fts_RBC <- rownames(res_sex_neg_sep)[
    order(res_sex_neg_sep$pvalue.RBC.sexM)][1:50]

fts_any <- unique(c(fts_capillary, fts_venous, fts_plasma, fts_RBC))
ovlp <- matrix(nrow = length(fts_any), ncol = 4, 0,
               dimnames = list(fts_any, c("capillary", "plasma", "RBC",
                                          "venous")))
ovlp[fts_capillary, "capillary"] <- 1
ovlp[fts_plasma, "plasma"] <- 1
ovlp[fts_RBC, "RBC"] <- 1
ovlp[fts_venous, "venous"] <- 1

upset(data.frame(ovlp))

```

The overlap is surprisingly low.

```{r sex-sep-export, echo = FALSE}
library(writexl)
write_xlsx(as.data.frame(res_sex_neg_sep),
           path = "data/xls/vams_sex_neg_sep_results.xlsx")

save(res_sex_neg_sep, file = paste0(RDATA_PATH, "res_sex_neg_sep.RData"))

```

Things we could do:
- check if the plasma-specific features are very low in the other matrices and
  *vice versa*.



## Multiple regression analysis

We next use multiple linear regression to identify features with (potentially
significant) differences between male and female participants on the full data
set.

```{r Linear-fit-combined, message = FALSE, warning = FALSE}
sex <- vams_neg$sex
source <- vams_neg$source
age <- vams_neg$age
#' Define the design of the study
dsgn <- model.matrix(~ 0 + sex : source)
colnames(dsgn) <- sub(":", ".", colnames(dsgn), fixed = TRUE)

#' Fit the linear model
fit <- lmFit(log2(assay(vams_neg, "imputed")), design = dsgn)

#' Define and fit the contrasts
contr <- makeContrasts(
    capillary.sexM = sexM.sourcecapillary - sexF.sourcecapillary,
    plasma.sexM = sexM.sourceplasma - sexF.sourceplasma,
    RBC.sexM = sexM.sourceRBC - sexF.sourceRBC,
    venous.sexM = sexM.sourcevenous - sexF.sourcevenous,
    levels = dsgn)
fit <- contrasts.fit(fit, contr)
#' Calculate p-values
fit <- eBayes(fit)

#' Adjust for multiple hypothesis testing
adjp <- apply(fit$p.value, 2, p.adjust, method = "BH")

sign <- adjp < pLimit & abs(fit$coefficients) > foldLimit

res_sex_neg <- cbind(rowData(vams_neg),
                     pvalue = fit$p.value,
                     coef = fit$coefficients,
                     adj.pvalue = adjp,
                     significant = sign)

```

As before, histograms of the unadjusted p-values are made.

```{r sex-neg-pvalue-hist, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p-values for male vs female in each of the matrices. Analysis was performed on the full data.", fig.width = 12, fig.height = 4}
par(mfrow = c(1, 4))
for (src in TestedSourcenames) {
    hist(res_sex_neg[, paste0("pvalue.", src, ".sexM")],
         main = paste0("p-value of Male vs Female in ", src),
         xlab = "p-value")
}

```

The enrichment for small p-values seems smaller than in the analysis performed
separately for each matrix. We even see an enrichment of large p-values. By
including the participant's age into the linear model this enrichment in large
p-values disappears, but the results are similar (data not shown). Next we
create volcano plots representing the results for the comparison between male
and female individuals.

```{r sex-neg-volcano, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Volcano plots representing the results for the comparison between male and female individuals. Analysis was performed on the full data.", fig.width = 12, fig.height = 4}
par(mfrow = c(1, 4))
xlim <- range(as.matrix(res_sex_neg[, grep("^coef", colnames(res_sex_neg))]))
ylim <- range(as.matrix(res_sex_neg[, grep("^adj", colnames(res_sex_neg))]))
for (src in TestedSourcenames) {
    volcanoplotFunction(
        res_sex_neg[, paste0("coef.", src, ".sexM")],
        res_sex_neg[, paste0("adj.pvalue.", src, ".sexM")],
        pLimit = pLimit, foldLimit = foldLimit, xlim = xlim,
        ylim = range(-log10(ylim)), Title = paste0("M vs F ", src))
}

```

We do now find (at least) one significant feature in all matrices.

The table below lists all features that are found to be significant in at least
one of the matrices.

```{r sex-neg-sig-table, results = "asis", echo = FALSE}
sig_count <- rowSums(
    as.matrix(res_sex_neg[, grep("significant", colnames(res_sex_neg))]))
sig_fts <- rownames(res_sex_neg)[sig_count > 0]
sig_fts <- sig_fts[order(sig_count[sig_fts], decreasing = TRUE)]

T <- res_sex_neg[sig_fts, c("mzmed", "rtmed", "coef.capillary.sexM",
                            "coef.plasma.sexM", "coef.RBC.sexM",
                            "coef.venous.sexM", "adj.pvalue.capillary.sexM",
                            "adj.pvalue.plasma.sexM", "adj.pvalue.RBC.sexM",
                            "adj.pvalue.venous.sexM")]
colnames(T) <- sub(".sexM", "", colnames(T), fixed = TRUE)
cap <- paste0("Features with significant difference in abundances between ",
              "male and female individuals in at least one matrix.")
pandoc.table(as.data.frame(T), style = "rmarkdown", caption = cap)

```

There is indeed no overlap of significant features. With the exception of some
(rows 3, 4, 12) there is also no agreement in the extent and direction of
differential abundance.

```{r sex-neg-sig-ann, echo = FALSE, eval = FALSE}
library(CompoundDb)
cdb <- CompDb("~/Projects/compounds/hmdb/4.1/CompDb.Hsapiens.HMDB.4.1.sqlite")

annotateMz(T$mzmed, cdb, adduct = c("[M-H]-", "[M+Cl]-"), ppm = 10)

#' it's a mess - either no annotation or multiple.
```

At last we compare the top 50 features with the most significant differences
between male and female for each matrix to evaluate whether they are shared
between matrices.

```{r sex-neg-top-50-pval, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Overlap of top 50 features with smallest p-values (male vs female)."}
fts_capillary <- rownames(res_sex_neg)[
    order(res_sex_neg$pvalue.capillary.sexM)][1:50]
fts_venous <- rownames(res_sex_neg)[
    order(res_sex_neg$pvalue.venous.sexM)][1:50]
fts_plasma <- rownames(res_sex_neg)[
    order(res_sex_neg$pvalue.plasma.sexM)][1:50]
fts_RBC <- rownames(res_sex_neg)[
    order(res_sex_neg$pvalue.RBC.sexM)][1:50]

fts_any <- unique(c(fts_capillary, fts_venous, fts_plasma, fts_RBC))
ovlp <- matrix(nrow = length(fts_any), ncol = 4, 0,
               dimnames = list(fts_any, c("capillary", "plasma", "RBC",
                                          "venous")))
ovlp[fts_capillary, "capillary"] <- 1
ovlp[fts_plasma, "plasma"] <- 1
ovlp[fts_RBC, "RBC"] <- 1
ovlp[fts_venous, "venous"] <- 1

upset(data.frame(ovlp))

```

The overlap is again surprisingly low.

```{r sex-neg-export, echo = FALSE}
write_xlsx(as.data.frame(res_sex_neg),
           path = "data/xls/vams_sex_neg_results.xlsx")

save(res_sex_neg, file = paste0(RDATA_PATH, "res_sex_neg.RData"))

```

# Session information

```{r sessionInfo}
devtools::session_info()
```

# References
