---
title: "Sex comparison"
author: "Johannes Rainer, Søren Fjelstrup, Chiara Volani and Giuseppe Paglia"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results='hide', message = FALSE}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_sex_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
#' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

#' Define paths:
filename <- "vams_sex_pos"
#' Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH)) 
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

#' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction
Having found the differences between each matrix in [vams_matrix_neg.Rmd](vams_matrix_neg.Rmd).
We move to a different way of comparing the matrices. Here we focus on the 
differences in the blood able to distinguish between male and female. 
We would expect that certain hormones (such as testosterone and estradiol) will 
be clear differences between genders, as well as derived differences such as 
creatine, as a consequences of higher average muscle mass in the males etc.


Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data , results='hide', message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
#' register(SerialParam())
library(SummarizedExperiment)
library(UpSetR)
library(pls)
library(caret)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization_neg/vams_neg.RData")


```

The values are imputed to allow their later usage in the linear model.

```{r Imputation-of-Values}
assays(vams_neg)$Imputed <- imputeRowMinRand(
                    assay(vams_neg), min_fraction = 1/2,
                    sd_fraction = 1, abs = TRUE)

TestedSourcenames <-  c("capillary","plasma", "RBC", "venous")

```


# Supervised modelling

## Linear Model - no confounding effects
A function for making the volcano plot later used for the linear model is made.


```{r volcanoplot-function}

col_volcano <- brewer.pal(4, name = "Set1")
col_volcano[1] <- "#DDDDDD" #' Grey
col_volcano_Trans <- paste0(col_volcano, 80)



volcanoplotFunction <- function(pval,coef,pLimit,foldLimit,Title){

#' The filters for pvalue and fold increase is defined
colorFilterLowP <- as.matrix(pval) < pLimit
colorFilterHighFold <- abs(coef) > foldLimit

#' The filter for coloring the points is made
colorFilter <- 1 * colorFilterLowP
colorFilter[colorFilterHighFold] <- 2
colorFilter[colorFilterHighFold * colorFilterLowP == 1] <- 3
colorFilter <- colorFilter + 1

    plot(coef, -log10(pval), cex = 0.6,
         pch = 16, main = Title, 
         col = col_volcano_Trans[colorFilter], 
         ylab = expression(log[2]~p[BH]-value), 
         xlab = expression(log[2]~fold~change))
    abline(h = -log10(pLimit))
    abline(v = c(-foldLimit, foldLimit))
    legend("bottomleft", c("non-Significant", "lowP", "Highfold", "Significant"),
           col = col_volcano, pch = 16, bg = "white")
     
}



```


The differences between male and female are found within each of the matrices.
Using a multiple linear model. No confounding factors are included.


```{r sex-fit-function}

#The new fit function is defined
fitlmSexFunction <- function(x, dataMat, model = conc ~ source, weights = NULL){
    dataMat <- cbind(conc = x,as.data.frame(dataMat))
    mdl <- lm(model, data = dataMat)
    smry <- data.frame(summary(mdl)$coefficients)
        Output <- c(coef = smry["sexM", 1], pvalue =  smry["sexM", 4])
    names(Output) <- c("coef.sexM","pvalue.sexM")
    Output
}

#The rowdata of vamspos is put into the result matrix. Likewise the imputed data
#is redefined into the container Data
res_matrix_neg<- rowData(vams_neg)
Data <- log2(assays(vams_neg)$Imputed)[,]

#A for loop in which each of the matrices are are passed to the above function
for (src in TestedSourcenames) {

fitMatrix <- t(apply(Data[, which(vams_neg$source==src)], 1, fitlmSexFunction, 
  dataMat = colData(vams_neg)[which(vams_neg$source==src),], model = conc ~ sex))

#The p-value is adjusted
adj.pvalue.sexM <- p.adjust(fitMatrix[, "pvalue.sexM"], method="BH")
fitMatrix <- cbind(fitMatrix, adj.pvalue.sexM)
#The column names are changed from sexM, to the current matrix being investigated.
colnames(fitMatrix) <- sub("sexM",src,colnames(fitMatrix))
#The new fit is bound to the result matrix
res_matrix_neg<- cbind(res_matrix_neg, fitMatrix)

#For reasons of aesthetics, the pvalue and coeffient is defined as variables 
#before being passed to the volcano plot function
pval <- res_matrix_neg[, paste0("adj.pvalue.", src)]
coef <- res_matrix_neg[, paste0("coef.", src)]
volcanoplotFunction(pval, coef, pLimit = 0.01, foldLimit = 2, Title=paste0("Volcano Plot, of Male vs Female in ", src))
}




```

Surprisingly not a single feature are found to be significant. Numerous have 
fold change above 2, but none have an adjusted p-value below 0.01.
The most extreme matrix is the RBC. Here almost all the features have a p-value
effectively equivalent to 1.

Histograms of the nonadjusted values are made as well.

```{r Histogramofpvalues, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values in each of the matrices of male vs female. The histograms generally show somewhat few significant features, especially for RBC"}
#Histograms of each of the sources are made in a for loop
for (src in TestedSourcenames) {
    hist(res_matrix_neg[, paste0("pvalue.", src)],
         main = paste0("pvalue. of Male vs Female in ", src),
         xlab = "p-value")
}

```

Looking at the histograms for the unadjusted pvalues, it is also evident that, 
compared to the inter-matrix changes, few significant features are found. Furthermore
confounding effects may be present, as the pvalues are not uniformly distributed.

```{r feature-count-table-no-confounding}

#The sum of unadjusted pvalues above 0.01 are found made put into a table
tableSums <- c()
for (src in TestedSourcenames) {
tableSums <- cbind(tableSums,sum(res_matrix_neg[,paste0("pvalue.",src)]<0.01))
}
names(tableSums) <- TestedSourcenames
cpt <- paste0("Number of features with a nonadjusted pvalue below 0.01 detected
              in each matrix")
pandoc.table(tableSums, style = "rmarkdown",
             caption = cpt)


```
As seen here most significant features are found in the three sources containing
plasma with the most in the plasma sample itself. This is not surprising as the 
plasma is where we would expect to see such differences.


The same analysis is done again, this time with the only other known variable,
age, as a confounding factor.

```{r sex-fit-function-confounding}


#A new result matrix for the analysis done with confounding variables are made
res_matrix_neg_Conf<- rowData(vams_neg)
Data <- log2(assays(vams_neg)$Imputed)[,]

#As before #A for loop in which each of the matrices are are passed to the above function
for (src in TestedSourcenames) {
fitMatrix <- t(apply(Data[, which(vams_neg$source==src)], 1, fitlmSexFunction, 
  dataMat = colData(vams_neg)[which(vams_neg$source==src),], model = conc ~ sex + age))

#The p-value is adjusted
adj.pvalue.sexM <- p.adjust(fitMatrix[, "pvalue.sexM"], method="BH")
fitMatrix <- cbind(fitMatrix, adj.pvalue.sexM)

#The column names are changed from sexM, to the current matrix being investigated.
colnames(fitMatrix) <- sub("sexM",src,colnames(fitMatrix))
#The new fit is bound to the result matrix
res_matrix_neg_Conf<- cbind(res_matrix_neg_Conf, fitMatrix)

#For reasons of aesthetics, the pvalue and coeffient is defined as variables 
#before being passed to the volcano plot function
pval <- res_matrix_neg_Conf[, paste0("adj.pvalue.", src)]
coef <- res_matrix_neg_Conf[, paste0("coef.", src)]
volcanoplotFunction(pval,coef,pLimit=0.01,foldLimit=2,Title=paste0("Volcano Plot, of Male vs Female in ", src))
}




```
Since the only other possible confounding factor available is the age, which is 
somewhat evenly distrubted among genders it is not surprising that we don't find
a big effect of including it.

A histrogram of the non-adjusted values are made as well.
```{r HistogramofpvaluesConfounding, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values in each of the matrices of male vs female. The histograms generally show somewhat few significant features, especially for RBC"}

#Histograms for the pvalues of the different matrices are made in a for loop.
for (src in TestedSourcenames) {
    hist(res_matrix_neg_Conf[, paste0("pvalue.", src)],
         main = paste0("pvalue. of Male vs Female in ", src),
         xlab = "p-value")
}

```

The distributions mostly look the same, with the sole exception of the capillary.
Here more low p-value features are found

These are counted in the below table

```{r feature-count-table-confounding}
#The sum of unadjusted pvalues above 0.01 are found made put into a table
tableSums <- c()
for (src in TestedSourcenames) {
tableSums <- cbind(tableSums,sum(res_matrix_neg_Conf[,paste0("pvalue.",src)]<0.01))
}
names(tableSums) <- TestedSourcenames
cpt <- paste0("Number of features with a nonadjusted pvalue below 0.01 detected
              in each matrix")
pandoc.table(tableSums, style = "rmarkdown",
             caption = cpt)


```

Surprisingly only the capillary has a higher amount of low-pvalue features.
The others are mostly the same whereas the capillary changes from about 300 to
about 400, plasma falls from 400 to 330 

The adjusted and unadjusted pvalue, and the coefficients in terms of 
differentiating between male and female along with the retention time and
the mass/charge ratio for each feature in each matrix is sent to "Output" and
exported to "Output/Res-sex-pos.txt"


```{r Export}

#each of the subtypes in the fit is defined as a seperate variable and then exported
OutputAdjPval <- res_matrix_neg[, paste0("adj.pvalue.",c(TestedSourcenames))]
OutputPval <- res_matrix_neg[, paste0("pvalue.",c(TestedSourcenames))]
OutputCoef <- res_matrix_neg[, paste0("coef.",c(TestedSourcenames))]
OutputMzRt <- cbind(res_matrix_neg$rtmed,res_matrix_neg$mzmed)
colnames(OutputMzRt) <- c("mass/charge","Retention time")

Output <- as.data.frame(c(OutputAdjPval,OutputPval,OutputCoef,OutputMzRt))
Output <- cbind(row.names(res_matrix_neg),Output)
names(Output)[1] <- "FeatureID"

write.table(Output,"Output/Res-sex-pos.txt",sep="\t",row.names = FALSE)

```


# Session information

```{r sessionInfo}
devtools::session_info()
```

# References
