---
title: "Matrix comparison - Pos"
author: "Søren Fjelstrup, Chiara Volani, Giuseppe Paglia and Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results='hide', message = FALSE}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_matrix_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
#' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

#' Define paths:
filename <- "vams_matrix_pos"
#' Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH)) 
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

#' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction

In this document we evaluate differences between the 4 matrices (sample sources)
*venous*, *capillary*, *plasma* and *RBC*. We are specifically interested in
identifying features which are specific to one matrix or common to all matrices
(qualitative analysis) as well as features with different abundances between
matrices (quantitative analyses). For the former analysis we use determine for
each feature whether it is *present* in a sample or not based on whether a
chromatographic peak was identified for it. The latter analysis is based on the
feature abundances.

The analysis uses the normalized feature abundances generated in
[vams_normalization.Rmd](vams_normalization.Rmd). The normalized data contains
abundance estimates for all features that are present in at least 30% of samples
of a matrix (QC samples excluded). The abundances from the two technical
replicates of each sample were combined into a single feature abundance. Also,
flags have been defined that provide information about the *quality* of a
normalized feature concentration. It is based on whether a chromatographic peak
was identified in a sample and how similar the reported abundances for the
feature in the two (technical) replicates for each sample are. Based on their
flag value, features can be classified into 6 different types:

- `1`: signal from 2 detected peaks present and the RSD of their abundances is <
  30%. The `mean` of the two is reported.
- `0.9`: 1 detected and one filled-in signal present and their RSD is < 30%. The
  `mean` of the two is reported.
- `0.8`: only filled-in signal, but their RSD is < 30%. The `mean` of the two is
  reported.
- `0.7`: signal for two detected peaks but their RSD is >= 30%. The `mean` of
  the two is reported.
- `0.5`: signal from 1 detected peak present (independently of whether a
  filled-in signal is present for the other replicate). Report only
  the value for the detected peak.
- `0.25`: only filled-in signal is present. Either mean or single value
  reported.

Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data , results='hide', message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
#' register(SerialParam())
library(SummarizedExperiment)
library(UpSetR)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization/vams_pos.RData")
#' find path
load("Output/CAMERA/ExportTable.RData")

```

## Aggregating data Based on Groups from CAMERA Package
In this analysis the data from vams_pos is not used directly. As an attempt to 
simplify the data the possible adducts were found using the CAMERA package. This was done in [CAMERA package.Rmd](CAMERA package.Rmd). This was done on the original xcms object data_pos. In the [vams_normalization.Rmd](vams_normalization.Rmd) some features were found to be too scarce to be included and these will have to be removed from the grouped data.

```{r Subsetting-of-ExportTable}

presentFeatures <- rownames(ExportTable)
CAMERAData <- ExportTable[rownames(vams_pos),]

```

It is now possible to do the actual aggregation. For each of the groups
the abundance is extracted, in addition the max weight for a group, as well as
the rows in which the highest abundance is found is returned as well.

```{r aggregation-function}
aggregationFunction <- function(Pattern, y, FUN = colSums) {
  Output <- c()
  logicalPattern <-
  which(vapply(y, function(z)
  any(z == Pattern), logical(1)))
  if (length(logicalPattern) > 1) {
  Output$Abundance <-
  FUN(assay(vams_pos, "abundances")[logicalPattern,]
  , na.rm = TRUE)
  Output$Weight <-
  colMaxs(assay(vams_pos, "weights")[logicalPattern,]
  , na.rm = TRUE)
  Output$maxRow <- which.max(assay(vams_pos, "abundances")
  [logicalPattern,])
  
  }
  else {
  Output$Abundance <- assay(vams_pos, "abundances")[logicalPattern, ]
  Output$Weight <- assay(vams_pos, "weights")[logicalPattern, ]
  Output$maxRow <- logicalPattern
  }
  Output
}


```

First the list of adduct groups are unlisted and only the unique adducts are
retained. These are then passed to the function. The list is then split into
each of it's constituents, and finally made into a SummarizedExperiemnt object.
Note that due to the way sums treat dropped columns with missing values, 0' are
introduced, these are replaced with NA.


```{r aggregation-on-adduct}
library(tictoc)


#' The unique compound groups are identified. These are what will be cycled over

uniqueGroups <- unique(unlist(CAMERAData$adductGroup))
uniqueGroups <- uniqueGroups[!is.na(uniqueGroups)]



#' The features are aggregated based on their adduct group
tic("a")
aggData <-
lapply(uniqueGroups, aggregationFunction, y = CAMERAData$adductGroup)
aggData <- do.call(rbind, aggData)

#' The data is reformated
Grouped <- c()
Grouped$Abundances <- do.call(rbind, aggData[, 1])
Grouped$Weight <- do.call(rbind, aggData[, 2])
Grouped$maxRow <- do.call(rbind, aggData[, 3])
rownames(Grouped$Abundances) <- uniqueGroups
rownames(Grouped$Weight) <- uniqueGroups
toc()

#' The grouped data is appended with the data in which no groups were found.
groupedData <-
rbind(Grouped$Abundances, assay(vams_pos, "abundances")
[is.na(CAMERAData$adductGroup), ])
groupedDataWeights <-
rbind(Grouped$Weight, assay(vams_pos, "weights")
[is.na(CAMERAData$adductGroup), ])
#' any 0's in the intensity data is replaced with a NA
groupedData[groupedData == 0] <- NA

#' The rowdata from the most abundant feature in each group is appended
#' with the ungrouped rowData.
#' Rownames are removed to reduce errors
rowDataGroup <-
rbind(rowData(vams_pos)[as.vector(Grouped$maxRow), ], rowData(vams_pos)
[is.na(CAMERAData$adductGroup), ])
rownames(rowDataGroup) <- NULL
#' The SummarizedExperiment object is made.
grpDta_pos <-
SummarizedExperiment(
assays = list(count = groupedData,
weights = groupedDataWeights),
colData = colData(vams_pos),
rowData = rowDataGroup
)


# boxplot(log2(rowMeans(assay(vams_pos, "abundances")[!is.na(CAMERAData$adductGroup),]))
#         , log2(rowMeans(assay(vams_pos, "abundances")[is.na(CAMERAData$adductGroup),]))
#         , ylab = expression(log[2]~abundance)
#         , names = (c("Adduct Group","No Adduct group")), las = 1
#         , main = "Abundance of Isotopes")

#' The SumarizedExperiment is saved for later use in the sex difference analysis.
save(grpDta_pos,
file = paste0("data/RData/vams_normalization/", "grpDta_pos.RData"))

```


The Data is imputed to remove NA's as these will cause problems in the
downstream models.

```{r Imputation-of-Values}
assays(grpDta_pos)$Imputed <- imputeRowMinRand(
                    assay(grpDta_pos), min_fraction = 1/2,
                    sd_fraction = 1, abs = TRUE)

```




## Grouping samples based on metabolite abundances

We perform a principal component analysis to evaluate the grouping of the
samples/matrices in our experiment.

```{r pca-rsd, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 10, fig.height = 5, fig.cap = "PCA grouping samples based on (centered) feature abundances. Only features with an RSD < 30% in QC samples were used. Circles represent samples from female, rectangles from male participants. Missing values were replaced with half of the feature's minimal abundance adding also a small random variance."}
#' RSD < 30% in QC samples
tmp <- log2(assay(grpDta_pos, "Imputed"))
pc <- prcomp(t(tmp), scale = FALSE, center = TRUE)
pchs <- ifelse(grpDta_pos$sex == "F", 21, 22)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[grpDta_pos$source], "ce"),
         bg = paste0(col_source[grpDta_pos$source], 60), pch = pchs, cex = 1.5)
legend("topright", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[grpDta_pos$source], 80),
         bg = paste0(col_source[grpDta_pos$source], 40), pch = pchs, cex = 1.5)

```

The PCA above was performed on all features that have an RSD < 30% in QC samples
(about 4000). Missing values were replaced with half of the feature's smallest
abundance adding also a small random value based on the features' standard
deviations. These PCA-plots are mostly identical to the ones using the entire
dataset, As ony 800 features were removed this is not very surprising. What is
more surprising is that only using the aggregated data (about 5000 features)
gives the same results.

## Identifying detected/present features for each matrix

We next aim to identify *present* (detected) features for each matrix which
allows to perform qualitative comparisons between the matrices. we evaluate
whether a feature is detected in most samples of a matrix or not, i.e. whether a
chromatographic peak was identified or not. This information is encoded into the
*weight* of a measurement (see Introduction for the definition of the weights):
a chromatographic peak was detected for a feature in at least one of the two
replicates if the weight is either 1, 0.9, 0.7 or 0.5. Below we use these
weights to determine the proportion of samples for each matrix in which a
chromatographic peak was identified.


```{r proportion-detected, message = FALSE, warning = FALSE}
wght_detected <- c(1, 0.9, 0.7, 0.5)

#' Determine for each feature the proportion of samples (per matrix) in
#' which a peak was detected/
prop_detected <- apply(assay(grpDta_pos, "weights"), 1, function(x, grps) {
  vapply(split(x, grps), function(z) sum(z %in% wght_detected) / length(z),
           numeric(1)) }, grps = vams_pos$source)
prop_detected <- t(prop_detected)
```

Below we calculate first the ratio between the proportion of samples in which a
peak was detected and the maximal proportion for that feature and subsequently
define the absent/present calls for each feature in each matrix based on these
values.

Note that the present analysis bases on features with a detected peak in at
least 1/3 of samples.

```{r present-absent-call, message = FALSE, warning = FALSE}
#' Calculate ratio between proportion and max proportion
prop_ratio <- (1 / rowMax(prop_detected)) * prop_detected

#' Define present/absent calls
present_call <- prop_ratio > 1/3
```

## Comparison of numbers of features detected in each matrix

With the present/absent calls available we can next evaluate the number of
features being detected in each matrix. The results are represented with the
barplot below.

```{r barplot-features, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Number of features detected by each matrix", fig.width = 6, fig.height = 8}
par(mfrow = c(1, 1), mar = c(7, 4.5, 2, 0.5))
barplot(colSums(present_call), col = col_source[colnames(present_call)],
        main = "Detected features", ylab = "count", las = 2)
```
Not unexpectedly, Since most features are retianed the distribtions are mostly
identical tothe previous analysis. Most features are present in venous samples
followed by capillary and RBC samples. The fewest are present in plasma samples.
The numbers are shown also in the table below.

```{r feature-count-table, results = "asis", echo = FALSE}
cpt <- paste0("Number of features detected in each matrix")
pandoc.table(colSums(present_call), style = "rmarkdown",
             caption = cpt)
```
## Matrix-specific features and features present in all matrices

We next compare the overlaps and intersects of features being *present* in the 4
matrices.
```{r upsetr-plot, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Visualisation of the feature intersects between the 4 matrices.", fig.width = 10, fig.height = 8}
ints <- apply(present_call, MARGIN = 2, which)
upset(fromList(ints), order.by = "freq",
      sets.bar.color = col_source[c("venous", "capillary", "RBC", "plasma")])
```

About one third of the ~ 10,000 features is detectable in all 4 matrices. The
next largest overlap with ~ 2,000 features is between venous, capillary and RBC
samples followed by about 1,100 features that are specific to plasma and venous
samples. Over and above the results are as expected, with the largest overlap of
matrices being between samples containing also blood cells. The numbers of
matrix-specific features is with ~ 200 - 500 relatively small, with plasma
samples having the largest number of unique features. Note that the overlap
of features is higher between capillary and RBC samples (570) compared to
capillary and venous samples (280).


The features present in either all matrices, or all of a specific matrix is
defined.

```{r matrix-specific-features}
feats_common <- rownames(present_call)[rowSums(present_call) == 4]
feats_plasma <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "plasma"]]
feats_venous <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "venous"]]
feats_RBC <- rownames(present_call)[rowSums(present_call) == 1 &
                                    present_call[, "RBC"]]
feats_capillary <- rownames(present_call)[rowSums(present_call) == 1 &
                                          present_call[, "capillary"]]
```



```{r matrix-specific-abundance, message = FALSE, fig.path = IMAGE_PATH, fig.cap = "Distribution of average (across samples) abundances of common and matrix-specific features.", fig.width = 6, fig.height = 8}
ints <- list(
    common = rowMeans(assay(grpDta_pos)[feats_common, ]),
    plasma = rowMeans(assay(grpDta_pos)[feats_plasma,
                                      grpDta_pos$source == "plasma"]),
    venous = rowMeans(assay(grpDta_pos)[feats_venous,
                                      grpDta_pos$source == "venous"]),
    RBC = rowMeans(assay(grpDta_pos)[feats_RBC,
                                   grpDta_pos$source == "RBC"]),
    capillary = rowMeans(assay(grpDta_pos)[feats_capillary,
                                         grpDta_pos$source == "capillary"])
)
par(mar = c(7, 4.5, 1, 1))
boxplot(lapply(ints, log2), varwidth = TRUE, col = col_source[names(ints)],
        ylab = expression(log[2]~abundance), las = 2, main = "average abundance")
grid(nx = NA, ny = NULL)
```
Indeed, the average abundance of matrix-specific features seems to be about
two-fold lower than the abundance of *common* features. Thus many features might
appear to be matrix specific because their abundance is below the detection
limit in other matrices.


# Supervised modelling

## Linear Model

The multiple linear regression of the features is made without imputation.
The models were made using the capillary as a baseline. Since the RBC and the
plasma were made from the venous it is expected to see large differences in all 
the models made. Although it would be expected that the pure venous would be the
most alike to the capillary.

In this analysis we fit the regression model to log2 transformed abundances of
each feature and extract the coefficients from the fitted model representing the
slope (or difference) of the average abundance of samples from the respective
matrix to the base line capillary samples. The p-value represents the
significance from the Wald test against the null hypothesis of the coefficient
being 0 (i.e. no differences between the average abundances).

The p-values are subsequently adjusted for multiple hypothesis testing using the
method from Benjamini and Hochberg for a strong control of the false discovery
rate.

The employed model `conc ~ source + sex` adjusts for sex-specific differences. A
simple model without accounting for any confounding effects yielded however
similar results (not shown).

```{r ModelforlinearmodelVSplasmanoconfounding, warning = FALSE}
SourcesTypes <- c("sourceplasma","sourceRBC","sourcevenous")

#' The function for the multiple linear regression is made. This takes a numeric
#' input along with a matrix containing the confounding factors, it returns the
#' means and the p value
fitlmFunction <- function(x, dataMat, model = conc ~ source) {
    dataMat$conc <- x
    dataMat$source <- factor(x = dataMat$source, 
                             levels = c("capillary", "plasma", "RBC", "venous"))
    mdl <- lm(model, data = dataMat)
    smry <- data.frame(summary(mdl)$coefficients)
    SourceTypes <- paste0("source", levels(dataMat$source)[-1])
    Output <- c(coef = smry[SourceTypes, 1], pvalue =  smry[SourceTypes, 4])
    names(Output) <- c(paste0("coef.", SourceTypes),
                       paste0("pvalue.", SourceTypes))
    Output
}

#' The data is log2 transformed to make it mostly gaussian in distribution  
Data <- log2(assay(grpDta_pos))
fitMatrix <- t(apply(Data, 1, fitlmFunction, 
                     dataMat = as.data.frame(colData(grpDta_pos)),
                     model = conc ~ source + sex))

#' Adjusting for multiple hypothesis testing
p.adjusted <- apply(fitMatrix[ ,paste0("pvalue.", SourcesTypes)], 2, p.adjust, 
                    method = "BH")
#' The fitMatrix is appended to the rest of the data in vams_pos, making the
#' new dataframe res_matrix_pos
res_matrix_pos<- cbind(rowData(grpDta_pos), fitMatrix, adj = p.adjusted)
```

Histograms of each of the 3 matrixes against capillary is made.

```{r Histogramofpvalues, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values each of the linear models against capillary. All histograms show a smooth decrease over the range of pvalues, indicating only a small possible effect from confounding factors"}
TestedSourcenames <-  c("plasma", "RBC", "venous")
par(mfrow=c(1, 3))
for (i in 1:3) {
    hist(res_matrix_pos[, paste0("pvalue.", SourcesTypes[i])],
         main = paste0(TestedSourcenames[i], " vs capillary"),
         xlab = "p-value")
}


```

It looks like the p-values decrease gradually, this suggest that the effect of
any confounding effects are likely to be minor, compared to the overall
differences from the matrices. We do however have a very large number of
features with differences in their abundances.

Based on the models created in the previous section volcano plots are made for
each of the matrices compared to capillary.

```{r VolcanoPlotMultipleRegression, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Volcano plots of each of the linear models against capillary. Benjamini-Hochberg adjusted values, against the mean differences for each feature, show a large proportion of the features are statistically significantly different between matrixes", fig.width = 12, fig.height = 4}
#' A new color set is defined to better conform to what is need by the volcano plot
col_volcano <- brewer.pal(4, name = "Set1")
col_volcano[1] <- "#DDDDDD" #' Grey
col_volcano_Trans <- paste0(col_volcano, 80)

pLimit <- 0.01
foldLimit <- 1

#' The filters for pvalue and fold increase is defined
colorFilterLowP <- as.matrix(
    res_matrix_pos[, paste0("adj.pvalue.", SourcesTypes)]) < pLimit
colorFilterHighFold <- abs(as.matrix(
    res_matrix_pos[, paste0("coef.", SourcesTypes)])) > foldLimit

#' The filter for coloring the points is made
colorFilter <- 1 * colorFilterLowP
colorFilter[colorFilterHighFold] <- 2
colorFilter[colorFilterHighFold * colorFilterLowP == 1] <- 3
colorFilter <- colorFilter + 1

#' The significant samples (both high fold difference and low p-value) are
#' defined
sign <- colorFilterLowP & colorFilterHighFold
#' replace the "coef" in the column names with "significant"
colnames(sign) <- sub("adj.pvalue", "significant", colnames(sign))
res_matrix_pos <- cbind(res_matrix_pos, sign)

par(mfrow=c(1, 3), mar = c(4.3, 4.3, 1, 1))
for (i in 1:3){
    plot(res_matrix_pos[, paste0("coef.", SourcesTypes[i])],  cex = 0.6,
         -log10(res_matrix_pos[, paste0("adj.pvalue.", SourcesTypes[i])]),
         pch = 16, main = paste0(TestedSourcenames[i], " vs capillary"), 
         col = col_volcano_Trans[colorFilter[, i]], 
         ylab = expression(log[10]~p[BH]-value), 
         xlab = expression(log[2]~fold~change))
    abline(h = -log10(pLimit))
    abline(v = c(-foldLimit, foldLimit))
    legend("top", c("non-Significant", "lowP", "Highfold", "Significant"),
           col = col_volcano, pch = 16, bg = "white")
}


```

As the data is mostly the same as the original data the results are comparable A
large amount of features were detected to have statistically different
abundances between the compared matrices. The number of significant features for
each comparison are listed below.

```{r count-table_linear_model, results = "asis", echo = FALSE}
cpt <- paste0("Number of features with a significant difference in abundances",
              " (more than two-fold difference in abundances at a 1% FDR.")
pandoc.table(colSums(as.matrix(
    res_matrix_pos[, grep("significant", colnames(res_matrix_pos))]),
    na.rm = TRUE), style = "rmarkdown", caption = cpt)
```

As expected, The results match what was seen in the previous analysis the
largest number of features with differences in abundances was identified between
plasma and capillary samples. The number of significant features is much smaller
for the comparisons between RBC and capillary and for venous and capillary
samples.
The large differences may have many sources. To help identify these it was
decided to probe the underlying distributions of the features based on the two
two factors we know. The mass/charge ratio and the retention time. It would be
expected that the venous and capillary would be mostly the same with the plasma
and RBC being different.
Plots showing the distributions of the matrixes based on m/z were made.

```{r pValueDependencyOnmzNoconfounding, fig.cap = "Distribution of m/z values (left) or retention time (right) for detected features in a matrix.", echo = FALSE, fig.width = 12, fig.height = 6}

#' Distributions of mz values for each matrix.
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
plot(density(res_matrix_pos$mzmed[present_call[, 1]], na.rm = TRUE),
     col = paste0(col_source[colnames(present_call)[1]], 80),
     main = "Detected features in a matrix",
     xlab = "m/z", ylim = c(0, 0.002), lwd = 2)
for (i in 2:4) {
    lines(density(res_matrix_pos$mzmed[present_call[, i]], na.rm = TRUE), 
          col = paste0(col_source[colnames(present_call)[i]], 80), lwd = 2)
}
legend("topleft", legend = c(colnames(present_call)),
       col = col_source[colnames(present_call)], pch = 16)
#' retention time
plot(density(res_matrix_pos$rtmed[present_call[, 1]], na.rm = TRUE),
     col = paste0(col_source[colnames(present_call)[1]], 80),
     main = "Detected features in a matrix",
     xlab = "rt", ylim = c(0, 0.015), lwd = 2)
for (i in 2:4) {
    lines(density(res_matrix_pos$rtmed[present_call[, i]], na.rm = TRUE), 
          col = paste0(col_source[colnames(present_call)[i]], 80), lwd = 2)
}

```

Looking at the distribution of detected features as a function of their m/z,
only the plasma is clearly different from the others. Here a higher proportion
of medium-m/z, and much fewer high-m/z are found. This is not surprising. When
making plasma high-weight compounds are removed from the matrix. The m/z
roughly corresponds to what we would expect from diacyl-glycerols. (Fatty acids)

The distributions of retention times for detected features (right plot in the
figure above) are mostly similar with some notable differences especially in the
higher retention time range: venous samples have more features in the 140-170
seconds range compared to all other matrices, but much fewer in the 170-230
seconds range. This goes against the theory that the differences seen for the
m/z is due to fatty acids. If this was the case we would expect to see a low
peak for the low retention time. The opposite is seen, features with a long
retention time, i.e. being hydrophillic, have a lower presence. Thus whatever is
missing in the plasma sample must be both large and hydrophillic.

The distributions for capillary and RBC samples are highly similar, while venous
samples are more like plasma samples in the retention time range around 160
seconds.

Next we evaluate the distribution of m/z and retention times for the features
with significant differences in abundances compared to capillary samples.

```{r mzSignificant, fig.cap = "Distributions of m/z values (left) and retention times (right) for features with significant different abundances compared to the levels in capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6}
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
#' m/z
plot(density(res_matrix_pos$mzmed, na.rm = TRUE),
     col = "#cecece", main = "Significant features",
     xlab = "m/z", ylim = c(0, 0.002), lwd = 2)
for (i in 1:3) {
    lines(density(
        res_matrix_pos$mzmed[res_matrix_pos[, paste0("significant.",
                                                     SourcesTypes[i])]],
        na.rm = TRUE), col = paste0(col_source[TestedSourcenames[i]], 80),
        lwd = 2)
}
legend("topleft", legend=c("All features", TestedSourcenames), 
       col = c("#cecece", col_source[TestedSourcenames]), pch = 16)
#' rt
plot(density(res_matrix_pos$rtmed, na.rm = TRUE),
     col = "#cecece", main = "Significant features",
     xlab = "rt", ylim = c(0, 0.04), lwd = 2)
for (i in 1:3) {
    lines(density(
        res_matrix_pos$rtmed[res_matrix_pos[, paste0("significant.",
                                                     SourcesTypes[i])]],
        na.rm = TRUE), col = paste0(col_source[TestedSourcenames[i]], 80),
        lwd = 2)
}
```

This distribution includes only the features shown to be significantly different
in the previous model. The overall shape of the density lines is mostly similar
with the one for all features with the most notable differences in the m/z range
from 600 to 800.

Features with significant different abundances between plasma and capillary
samples are enriched in this m/z region, while capillary and venous samples seem
to be similar in that range. Also interestingly, many significant features
between venous and capillary samples were found for the m/z range of ~ 250 to
450. This range corresponds roughly to where sugars and steroids are expected to
be found.

The distribution of significant features along the retention time dimension
differs considerably. The largest differences are in the high retention time
range. Specifically venous samples show a high proportion of significantly
different features (against capillary samples) with a retention of ~ 160 seconds
and, to a lower degree also of, 180 and 200 seconds. The plasma is not
suprisingly mostly different in the high range (200-220 seconds).

Next we evaluate over- and under-represented features (compared to capillary
samples) separately.

```{r mzCountsSignificant, fig.cap = "Distribution of m/z values (left) and retention time (right) for significant features with a higher or lower abundance compared to capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6}

par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1.5, 1))
plot(x = 1, y = 1, pch = NA, xlim = c(50, 1100), ylim = c(-100, 100),
     xlab = "m/z", ylab = "count", main = "Significant features")
text(x = c(50, 50), y = c(-90, 90), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] < 0), "mzmed"]
    yLow <- hist(yLow, breaks = seq(50, 1100, 5), plot = FALSE)   
    yHigh <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] > 0), "mzmed"]
    yHigh <- hist(yHigh, breaks = seq(50, 1100, 5), plot = FALSE)
    lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
}
grid()
abline(h = 0)
legend("topright", legend = TestedSourcenames, lwd = 2, lty = 1,
       col = col_source[TestedSourcenames], inset = c(0, 0))
plot(x = 1, y = 1, pch = NA, xlim = c(0, 250), ylim = c(-800, 800),
     xlab = "rt", ylab = "count", main = "Significant features")
text(x = c(0, 0), y = c(-750, 750), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] < 0), "rtmed"]
    yLow <- hist(yLow, breaks = seq(0, 250, 5), plot = FALSE)   
    yHigh <- res_matrix_pos[
        which(res_matrix_pos[, paste0("significant.source", src)] & 
              res_matrix_pos[, paste0("coef.source", src)] > 0), "rtmed"]
    yHigh <- hist(yHigh, breaks = seq(0, 250, 5), plot = FALSE)
    lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
}
grid()
abline(h = 0)

```
 
The distributions are mostly the same as with the entire data set, with one
notable exception. In the original dataset almost no significant features were
found in the low Mz range. In constrast this is now where most of the features
are found, the reason for this is not clear



# Session information

```{r sessionInfo}
devtools::session_info()
```

# References
