---
title: "Sex comparison"
author: "Johannes Rainer, Søren Fjelstrup, Chiara Volani and Giuseppe Paglia"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results='hide', message = FALSE}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_sex_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
#' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

#' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction
Having found the differences between each matrix in [vams_matrix_neg_semtar.Rmd](vams_matrix_neg_semtar.Rmd).
We move to a different way of comparing the matrices. Here we focus on the 
differences in the blood able to distinguish between male and female. 
We would expect that certain hormones (such as testosterone and estradiol) will 
be clear differences between genders, as well as derived differences such as 
creatine, as a consequences of higher average muscle mass in the males etc.

Note that only a limited, already identified features are used.

Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data , results='hide', message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
#' register(SerialParam())
library(SummarizedExperiment)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization/SemiTargeted_neg.RData")


```

The values are imputed to allow their later usage in the linear model.
This is needed as the later models are run on one matrix at a time.
This means that if the imputation is not done, rows will be dropped, and the
final matrix will not fit.

```{r Imputation-of-Values}
assays(SemiTargeted)$Imputed <- imputeRowMinRand(
                    assay(SemiTargeted), min_fraction = 1/2,
                    sd_fraction = 1, abs = TRUE)

TestedSourcenames <-  c("capillary","plasma", "RBC", "venous")

```


# Supervised modelling

## Linear Model - no confounding effects
A function for making the volcano plot later used for the linear model is made.

```{r volcanoplot-function}

col_volcano <- brewer.pal(4, name = "Set1")
col_volcano[1] <- "#DDDDDD" #' Grey
col_volcano_Trans <- paste0(col_volcano, 80)



volcanoplotFunction <- function(pval, coef, pLimit, foldLimit, Title){

#' The filters for pvalue and fold increase is defined
colorFilterLowP <- as.matrix(pval) < pLimit
colorFilterHighFold <- abs(coef) > foldLimit

#' The filter for coloring the points is made
colorFilter <- 1 * colorFilterLowP
colorFilter[colorFilterHighFold] <- 2
colorFilter[colorFilterHighFold * colorFilterLowP == 1] <- 3
colorFilter <- colorFilter + 1

plot(coef, -log10(pval), cex = 0.6, pch = 16, main = Title, 
  col = col_volcano_Trans[colorFilter], ylab = expression(log[2] ~ p[BH]-value),
  xlab = expression(log[2] ~ fold ~ change))
    
abline(h = -log10(pLimit))
abline(v = c(-foldLimit, foldLimit))
legend("bottomleft", c("non-Significant", "lowP", "Highfold", "Significant"),
  col = col_volcano, pch = 16, bg = "white")
     
}



```


The differences between male and female are found within each of the matrices.
Using a multiple linear model. No confounding factors are included.


```{r sex-fit-function}

#' The new fit function is defined
fitlmSexFunction <- function(x, dataMat, model = conc ~ source, weights = NULL){
  dataMat <- cbind(conc = x, as.data.frame(dataMat))
  mdl <- lm(model, data = dataMat)
  smry <- data.frame(summary(mdl)$coefficients)
  Output <- c(coef = smry["sexM", 1], pvalue =  smry["sexM", 4])
  names(Output) <- c("coef.sexM", "pvalue.sexM")
  Output
}

#' The rowdata of vamspos is put into the result matrix. Likewise the imputed data
#' is redefined into the container Data
res_matrix_neg <- rowData(SemiTargeted)
Data <- log2(assays(SemiTargeted)$Imputed)[, ]

#' A for loop in which each of the matrices are are passed to the above function
for (src in TestedSourcenames) {

  fitMatrix <- t(apply(Data[, which(SemiTargeted$source == src)], 1,
    fitlmSexFunction, dataMat = colData(SemiTargeted)
    [which(SemiTargeted$source == src), ], model = conc ~ sex))

#' The p-value is adjusted
  adj.pvalue.sexM <- p.adjust(fitMatrix[, "pvalue.sexM"], method = "BH")
  fitMatrix <- cbind(fitMatrix, adj.pvalue.sexM)
#' The column names are changed from sexM, to the current matrix being prossesed
  colnames(fitMatrix) <- sub("sexM", src, colnames(fitMatrix))
#' The new fit is bound to the result matrix.
  res_matrix_neg <- cbind(res_matrix_neg, fitMatrix)

#' For reasons of aesthetics, the pvalue and coeffient is defined as variables 
#' before being passed to the volcano plot function.
  pval <- res_matrix_neg[, paste0("adj.pvalue.", src)]
  coef <- res_matrix_neg[, paste0("coef.", src)]
  volcanoplotFunction(pval, coef, pLimit = 0.01, foldLimit = 2, 
    Title = paste0("Volcano Plot, of Male vs Female in ", src))
}




```

For capillary one feature were found to have an adjusted p-value below 0.01 and
one feature a fold increase over 2, but none were both. Venous had a single
feature with a fold increase above 2. The most extreme matrix is the RBC. Here
almost all the features have a p-value effectively equivalent to 1.

Histograms of the nonadjusted values are made as well.

```{r Histogramofpvalues, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values in each of the matrices of male vs female. The histograms generally show somewhat few significant features, especially for RBC"}
#' Histograms of each of the sources are made in a for loop
for (src in TestedSourcenames) {
    hist(res_matrix_neg[, paste0("pvalue.", src)],
         main = paste0("pvalue. of Male vs Female in ", src), xlab = "p-value")
}

```

Looking at the histograms for the unadjusted pvalues, it is also evident that,
compared to the inter-matrix changes, few significant features are found.
Furthermore confounding effects may be present, as the pvalues are not uniformly
distributed.

```{r feature-count-table-no-confounding}

#' The sum of unadjusted pvalues above 0.01 are found made put into a table
tableSums <- c()
for (src in TestedSourcenames) {
  tableSums <- cbind(tableSums, sum(res_matrix_neg
    [, paste0("pvalue.", src)] < 0.01))
}
names(tableSums) <- TestedSourcenames
cpt <- paste0("Number of features with a nonadjusted pvalue below 0.01 detected
              in each matrix")
pandoc.table(tableSums, style = "rmarkdown", caption = cpt)


```
As seen here, most significant features are found in the three sources
containing plasma with the most in the plasma sample itself. This is not
surprising as the plasma is where we would expect to see differences between
genders.
The same analysis is done again, this time with the only other known variable,
age, as a confounding factor.

```{r sex-fit-function-confounding}


#' A new result matrix for the analysis done with confounding variables are made
res_matrix_neg_Conf <- rowData(SemiTargeted)
Data <- log2(assays(SemiTargeted)$Imputed)[, ]

#' As before #A for loop in which each of the matrices are are passed to
#' the above function
for (src in TestedSourcenames) {
fitMatrix <- t(apply(Data[, which(SemiTargeted$source == src)], 1,
  fitlmSexFunction, dataMat = colData(SemiTargeted)
  [which(SemiTargeted$source == src), ], model = conc ~ sex + age))

#' The p-value is adjusted
adj.pvalue.sexM <- p.adjust(fitMatrix[, "pvalue.sexM"], method="BH")
fitMatrix <- cbind(fitMatrix, adj.pvalue.sexM)

#' The column names are changed from sexM, to the current matrix being investigated.
colnames(fitMatrix) <- sub("sexM", src,colnames(fitMatrix))
#' The new fit is bound to the result matrix
res_matrix_neg_Conf<- cbind(res_matrix_neg_Conf, fitMatrix)

#' For reasons of aesthetics, the pvalue and coeffient is defined as variables 
#' before being passed to the volcano plot function
pval <- res_matrix_neg_Conf[, paste0("adj.pvalue.", src)]
coef <- res_matrix_neg_Conf[, paste0("coef.", src)]
volcanoplotFunction(pval, coef, pLimit = 0.01, foldLimit = 2, 
  Title = paste0("Volcano Plot, of Male vs Female in ", src))
}




```
Since the only other possible confounding factor available is the age, which is 
somewhat evenly distrubted among genders it is not surprising that we don't find
a big effect of including it.

A histrogram of the non-adjusted values are made as well.
```{r HistogramofpvaluesConfounding, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values in each of the matrices of male vs female. The histograms generally show somewhat few significant features, especially for RBC"}

#' Histograms for the pvalues of the different matrices are made in a for loop.
for (src in TestedSourcenames) {
    hist(res_matrix_neg_Conf[, paste0("pvalue.", src)],
         main = paste0("pvalue. of Male vs Female in ", src), xlab = "p-value")
}

```

The distributions mostly look the same. The features are counted in the below
table.

```{r feature-count-table-confounding}
#' The sum of unadjusted pvalues above 0.01 are found made put into a table
tableSums <- c()
for (src in TestedSourcenames) {
  tableSums <- cbind(tableSums,
    sum(res_matrix_neg_Conf[, paste0("pvalue.", src)] < 0.01))
}
names(tableSums) <- TestedSourcenames
cpt <- paste0("Number of features with a nonadjusted pvalue below 0.01 detected
  in each matrix")
pandoc.table(tableSums, style = "rmarkdown", caption = cpt)


```

Capillary and plasma are unhchanged, so is RBC practically. Venous is almost
halfed going from 12 to 7
The adjusted and unadjusted pvalue, and the coefficients in terms of
differentiating between male and female along with the retention time and the
mass/charge ratio for each feature in each matrix is sent to "Output" and
exported to "Output/Res-sex-neg-semtar.txt"


```{r Export}

#' Each of the subtypes in the fit is defined as a seperate variable 
#' and then exported
OutputAdjPval <- res_matrix_neg[, paste0("adj.pvalue.", c(TestedSourcenames))]
OutputPval <- res_matrix_neg[, paste0("pvalue.", c(TestedSourcenames))]
OutputCoef <- res_matrix_neg[, paste0("coef.", c(TestedSourcenames))]
OutputMzRt <- cbind(res_matrix_neg$rtmed, res_matrix_neg$mzmed)
colnames(OutputMzRt) <- c("Retention time","mass/charge")

Output <- as.data.frame(c(OutputAdjPval, OutputPval, OutputCoef, OutputMzRt))
Output <- cbind(row.names(res_matrix_neg), Output)
names(Output)[1] <- "FeatureID"


Output <- cbind(Output, rowData(SemiTargeted)[,c("name", "mass", "mz_diff_ppm"
        ,"rt","rt_diff", "adduct" )])

write.table(Output,"Output/Res-sex-neg-semtar.txt", sep = "\t", row.names = FALSE)

```

## All Matrices Combined

Since the differences within each matrix were not clear enough to give a
definitive answer. It was decided to test the gender differences across all
matrices, but adjusted for the matrix source and age.


```{r Linear-fit-combined}

#' The rowdata of vamspos is put into the result matrix. Likewise the imputed data
#' is redefined into the container Data
res_matrix_neg_All <- rowData(SemiTargeted)
Data <- log2(assays(SemiTargeted)$Imputed)[, ]

fitMatrix <- t(apply(Data[, ], 1, fitlmSexFunction, 
  dataMat = colData(SemiTargeted)[, ], model = conc ~ sex + source + age))

#' The p-value is adjusted
adj.pvalue.sexM <- p.adjust(fitMatrix[, "pvalue.sexM"], method = "BH")
fitMatrix <- cbind(fitMatrix, adj.pvalue.sexM)
#' The new fit is bound to the result matrix
res_matrix_neg_sexM<- cbind(res_matrix_neg, fitMatrix)

#' For reasons of aesthetics, the pvalue and coeffient is defined as variables 
#' before being passed to the volcano plot function
pval <- res_matrix_neg_sexM[, "adj.pvalue.sexM"]
coef <- res_matrix_neg_sexM[, "coef.sexM"]
volcanoplotFunction(pval, coef, pLimit = 0.01, foldLimit = 2, 
  Title = paste0("Volcano Plot, of Male vs Female"))

```

The model made from the combined matrices look different from the model made
on the individua matrices. Interestingly, multiple samples with an adjusted 
p-values below 0.01 were found, but only a single feature with a fold change
above 2. 

As before, a histogram of the unadjusted p-values are made.

```{r HistogramofpvaluesNoConfounding, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values in each of the matrices of male vs female. The histograms generally show somewhat few significant features, especially for RBC"}

#' Histograms for the pvalues the combined matrices are made in a for loop.

hist(res_matrix_neg_sexM[, "pvalue.sexM"], 
  main = paste0("pvalue. of Male vs Female"), xlab = "p-value")


```

The histogram resembles the one seen for the capillary. With a peak of around
40, and a very steep decrease. This indicates that the effect of additional
confounding values are likely to be minor.
The features with an unadjusted pvalue below 0.01 are summed in the table below.

```{r feature-count-table-all}
#' The sum of unadjusted pvalues above 0.01 are found made put into a table
tableSums <- sum(res_matrix_neg_sexM[, "pvalue.sexM"] < 0.01)
names(tableSums) <- "Features"

cpt <- paste0("Number of features with a nonadjusted pvalue below 0.01")
pandoc.table(tableSums, style = "rmarkdown", caption = cpt)


```

18 features were found to have unadjusted p-values below 0.01. This is more 
than any of the individual matrices by themselves.

# Session information

```{r sessionInfo}
devtools::session_info()
```

# References
