---
title: "Matrix comparison - Neg"
author: "Søren Fjelstrup, Chiara Volani, Giuseppe Paglia and Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results='hide', message = FALSE}
library(BiocStyle)
BiocStyle::markdown()
```

**Modified**: `r file.info("vams_matrix_pos.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
#' Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

#' Define paths:
filename <- "vams_matrix_pos"
#' Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH)) 
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

#' Get the number of cpus allocated or fall back to 3
ncores <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 3))
```

# Introduction

In this document we evaluate differences between the 4 matrices (sample sources)
*venous*, *capillary*, *plasma* and *RBC*. We are specifically interested in
identifying features which are specific to one matrix or common to all matrices
(qualitative analysis) as well as features with different abundances between
matrices (quantitative analyses). For the former analysis we use determine for
each feature whether it is *present* in a sample or not based on whether a
chromatographic peak was identified for it. The latter analysis is based on the
feature abundances.

The analysis uses the normalized feature abundances generated in
[vams_normalization.Rmd](vams_normalization.Rmd). The normalized data contains
abundance estimates for all features that are present in at least 30% of samples
of a matrix (QC samples excluded). The abundances from the two technical
replicates of each sample were combined into a single feature abundance. Also,
flags have been defined that provide information about the *quality* of a
normalized feature concentration. It is based on whether a chromatographic peak
was identified in a sample and how similar the reported abundances for the
feature in the two (technical) replicates for each sample are. Based on their
flag value, features can be classified into 6 different types:

- `1`: signal from 2 detected peaks present and the RSD of their abundances is <
  30%. The `mean` of the two is reported.
- `0.9`: 1 detected and one filled-in signal present and their RSD is < 30%. The
  `mean` of the two is reported.
- `0.8`: only filled-in signal, but their RSD is < 30%. The `mean` of the two is
  reported.
- `0.7`: signal for two detected peaks but their RSD is >= 30%. The `mean` of
  the two is reported.
- `0.5`: signal from 1 detected peak present (independently of whether a
  filled-in signal is present for the other replicate). Report only
  the value for the detected peak.
- `0.25`: only filled-in signal is present. Either mean or single value
  reported.

Below we load all required libraries, define a color for each matrix and load
the normalized data.

```{r load-data , results='hide', message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(ncores)
register(DoparParam(), default = TRUE)
#' register(SerialParam())
library(SummarizedExperiment)
library(UpSetR)

#' Load utility functions for the analysis
source("util-functions.R")

#' Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")[c(1, 2, 4, 5)]
names(col_source) <- c("RBC",           #' red
                       "plasma",        #' blue
                       "capillary",     #' purple
                       "venous")        #' orange

load("data/RData/vams_normalization_neg/vams_neg.RData")

```

## Identifying detected/present features for each matrix

We next aim to identify *present* (detected) features for each matrix which
allows to perform qualitative comparisons between the matrices. In contrast to
features with differences in abundances (identified in the next section) we do
not consider absolute intensities in this analysis but rather evaluate whether a
feature is detected in most samples of a matrix or not, i.e. whether a
chromatographic peak was identified or not. This information is encoded into th
*weight* of a measurement (see Introduction for the definition of the weights):
a chromatographic peak was detected for a feature in at least one of the two
replicates if the weight is either 1, 0.9, 0.7 or 0.5. Below we use these
weights to determine the proportion of samples for each matrix in which a
chromatographic peak was identified.

```{r proportion-detected, message = FALSE, warning = FALSE}
wght_detected <- c(1, 0.9, 0.7, 0.5)

#' Determine for each feature the proportion of samples (per matrix) in
#' which a peak was detected/
prop_detected <- apply(assay(vams_neg, "weights"), 1, function(x, grps) {
    vapply(split(x, grps), function(z) sum(z %in% wght_detected) / length(z),
           numeric(1))
}, grps = vams_neg$source)
prop_detected <- t(prop_detected)
```

Defining absent/present calls is not trivial. Using a hard cut-off on the
proportion of samples with a detected peak can be problematic, as features can
not be expected to be present in all samples. Sex-specific features would for
example be expected to be present in only 50% of samples. Requiring features to
be present in 60% of samples to define them as *present* would thus flag all
sex-specific features to be *absent*.

We therefore use an approach that compares, separately for each feature, and in
each matrix the number of samples in which a chromatographic peak was detected
to the maximum such number in any of the matrices. Specifically, features are
called *absent* in a certain matrix, if a peak was detected in less than one
third of the maximal number of samples with a detected peak in any of the
matrices. As example, if a feature was detected in 10 samples of a matrix, but
only in 3 in another matrix, it will be called *absent* in the latter. On the
other hand, if a feature was detected at most in 4 samples in a matrix, it is
still considered to be present in matrices in which e.g. in 2 samples a peak was
detected.

Below we calculate first the ratio between the proportion of samples in which a
peak was detected and the maximal proportion for that feature and subsequently
define the absent/present calls for each feature in each matrix based on these
values.

Note that the present analysis bases on features with a detected peak in at
least 1/3 of samples.

```{r present-absent-call, message = FALSE, warning = FALSE}
#' Calculate ratio between proportion and max proportion
prop_ratio <- (1 / rowMax(prop_detected)) * prop_detected

#' Define present/absent calls
present_call <- prop_ratio > 1/3
```



## Comparison of numbers of features detected in each matrix

With the present/absent calls available we can next evaluate the number of
features being detected in each matrix. The results are represented with the
barplot below.

```{r barplot-features, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Number of features detected by each matrix", fig.width = 6, fig.height = 8}
par(mfrow = c(1, 1), mar = c(7, 4.5, 2, 0.5))
barplot(colSums(present_call), col = col_source[colnames(present_call)],
        main = "Detected features", ylab = "count", las = 2)
```

Most features are present in venous samples followed by plasma and then 
capillary. The fewest are present in RBC samples. The numbers
are also shown in the table below.
This is in contrast to postive mode, where the order was:
Venous, Capillary, RBC, Plasma.

```{r feature-count-table, results = "asis", echo = FALSE}
cpt <- paste0("Number of features detected in each matrix")
pandoc.table(colSums(present_call), style = "rmarkdown",
             caption = cpt)
```


## Matrix-specific features and features present in all matrices

We next compare the overlaps and intersects of features being *present* in the 4
matrices.

```{r upsetr-plot, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.cap = "Visualisation of the feature intersects between the 4 matrices.", fig.width = 10, fig.height = 8}
ints <- apply(present_call, MARGIN = 2, which)
upset(fromList(ints), order.by = "freq",
      sets.bar.color = col_source[c("venous", "plasma", "capillary", "RBC")])
```



About one third of the ~ 16,000 features is detectable in all 4 matrices. The
next largest overlap with ~ 2,000 features is between venous, and plasma
samples followed by about 1,700 features that are shared between venous, 
capillary and RBC. The plasma is the largest unique with about 1400, the others 
have about 500 unique features. About 1300 are shared between plasma,
capillary and venous.
It would seem that in negative mode the Plasma has very large differences 
compared to the others, with some overlap with the venous.


The features present in either all matrices, or all of a specific matrix is
defined.

```{r matrix-specific-features}
feats_common <- rownames(present_call)[rowSums(present_call) == 4]
feats_plasma <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "plasma"]]
feats_venous <- rownames(present_call)[rowSums(present_call) == 1 &
                                       present_call[, "venous"]]
feats_RBC <- rownames(present_call)[rowSums(present_call) == 1 &
                                    present_call[, "RBC"]]
feats_capillary <- rownames(present_call)[rowSums(present_call) == 1 &
                                          present_call[, "capillary"]]
```

Below we impute missing feature intensities. These are replaced with half of the
feature's smallest value in all samples adding a small standard deviation
to avoid constant values.

```{r Imputation-of-Values}
assays(vams_neg)$Imputed <- imputeRowMinRand(
                    assay(vams_neg), min_fraction = 1/2,
                    sd_fraction = 1, abs = TRUE)

```

What seems to be common to all matrix-specific features is that their abundance
is moderate. Below we thus evaluate whether the average abundance level of
matrix-specific features is different to those from common features.

```{r matrix-specific-abundance, message = FALSE, fig.path = IMAGE_PATH, fig.cap = "Distribution of average (across samples) abundances of common and matrix-specific features.", fig.width = 6, fig.height = 8}
ints <- list(
    common = rowMeans(assay(vams_neg)[feats_common, ]),
    plasma = rowMeans(assay(vams_neg)[feats_plasma,
                                      vams_neg$source == "plasma"]),
    venous = rowMeans(assay(vams_neg)[feats_venous,
                                      vams_neg$source == "venous"]),
    RBC = rowMeans(assay(vams_neg)[feats_RBC,
                                   vams_neg$source == "RBC"]),
    capillary = rowMeans(assay(vams_neg)[feats_capillary,
                                         vams_neg$source == "capillary"])
)
par(mar = c(7, 4.5, 1, 1))
boxplot(lapply(ints, log2), varwidth = TRUE, col = col_source[names(ints)],
        ylab = expression(log[2]~abundance), las = 2, main = "average abundance")
grid(nx = NA, ny = NULL)
```

Indeed, the average abundance of matrix-specific features seems to be about
two to three-fold lower than the abundance of *common* features. Thus many features might
appear to be matrix specific because their abundance is below the detection
limit in other matrices.


## Grouping samples based on metabolite abundances

We perform a principal component analysis to evaluate the grouping of the
samples/matrices in our experiment.

```{r pca-rsd, message = FALSE, warning = FALSE, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 10, fig.height = 5, fig.cap = "PCA grouping samples based on (centered) feature abundances. Only features with an RSD < 30% in QC samples were used. Circles represent samples from female, rectangles from male participants. Missing values were replaced with half of the feature's minimal abundance adding also a small random variance."}
#' RSD < 30% in QC samples
tmp <- log2(assay(vams_neg, "Imputed")[which(rowData(vams_neg)$QC_rsd < 0.3), ])
pc <- prcomp(t(tmp), scale = FALSE, center = TRUE)
pchs <- ifelse(vams_neg$sex == "F", 21, 22)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[vams_neg$source], "ce"),
         bg = paste0(col_source[vams_neg$source], 60), pch = pchs, cex = 1.5)
legend("bottomright", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[vams_neg$source], 80),
         bg = paste0(col_source[vams_neg$source], 40), pch = pchs, cex = 1.5)

```
The first dimension very strongly separate the plasma, venous and capillary 
samples. The RBC are spilt, with most being close to the capillary and some 
being close to the venous. The RBC has a very large spread, whereas all other 
groups in all dimension are compact clusters.
The second dimension separate the venous from the capillary and plasma.
The RBS is spread evenly across the second dimension.
The third dimension separate the plasma from the venous and capillary.
all but four (likely the same 4 being close to venous in PC1) of the RBC
sample have lower values in PC3 than all the other samples in the other matrices.
The fourth dimension does not seem to have a specific seperation, apart from
internal seperation of the RBC samples.


## Identifying features with differential abundance between the matrices

It is possible that the differences seen in the previous PCA plots are due to
overall concentration differences. Thus each of the sample types were plotted to 
see if significant differences in overall concentration is found.

```{r BoxplotOfAbundances, fig.path = IMAGE_PATH, fig.cap = "Distribution of average abundances of common features and features detected within a certain matrix.", echo = FALSE}

#' All feautures present in the respective sample types are found. Note this is
#' different to when only the unique were found.
feats_plasma <- rownames(present_call)[present_call[, "plasma"]]
feats_venous <- rownames(present_call)[present_call[, "venous"]]
feats_RBC <- rownames(present_call)[present_call[, "RBC"]]
feats_capillary <- rownames(present_call)[present_call[, "capillary"]]

#' A list containing the abundances of the respective features were made. 
#' Note that they are imputed as it otherwise have too many missing values
intsMean <- list(
    common =  rowMeans(assay(vams_neg, "Imputed")[feats_common, ]),
    plasma = rowMeans(assay(vams_neg, "Imputed")[feats_plasma,
                                                 vams_neg$source == "plasma"]),
    venous = rowMeans(assay(vams_neg, "Imputed")[feats_venous,
                                                 vams_neg$source == "venous"]),
    RBC = rowMeans(assay(vams_neg, "Imputed")[feats_RBC,
                                              vams_neg$source == "RBC"]),
    capillary = rowMeans(assay(vams_neg, "Imputed")[feats_capillary,
                                                    vams_neg$source == "capillary"])
)
  
#' Boxplot showing the distributions of the different groups
par(mar = c(7, 4.5, 1, 1))
boxplot(lapply(intsMean, log2), varwidth = TRUE, col = col_source[names(intsMean)],
        ylab = expression(log[2]~abundance), las = 2, main = "average abundance")
```

With the exception of common features that show slightly higher average
abundances, the average abundances of features detected in a matrix are
comparable. In addition to boxplots we create also cumulative distribution
functions to evaluate whether the distribution of average feature abundances are
comparable.

```{r CumulativeDistributionFunctionsOfAbundances, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Cumulative density function of each matrix. The distributions of the matrices are mostly identical, with the plasma having a sligthly higher overall abundance and the RBC a slightly higher proportion of low abundance features"}
#' Cumulative distribution functions are made for the different groups, 
#' log2 is necessary due to the large differences in abundances
plasma_ecdf <- ecdf(as.numeric(log2(intsMean$plasma)))
venous_ecdf <- ecdf(as.numeric(log2(intsMean$venous))) 
RBC_ecdf <- ecdf(as.numeric(log2(intsMean$RBC)))
capillary_ecdf <- ecdf(as.numeric(log2(intsMean$capillary)))

plot(plasma_ecdf, col = col_source["plasma"], main = "cdf of abundance",
     xlab = "log2 abundance", ylab = "Density" , xlim = c(0,20), ylim = (c(0,1)))
lines(venous_ecdf, col = col_source["venous"])
lines(RBC_ecdf, col = col_source["RBC"])
lines(capillary_ecdf, col = col_source["capillary"])

legend('left', c("plasma","venous","RBC","capillary"), fill = col_source[
  c("plasma","venous","RBC","capillary")], border=NA)
```


The cumulative density function shows that the abundances of the features for
the RBC is the lowest,and the plasma is the highest, the venous and capillary
are almost identical. The differences are however very minor, so it is unlikely
to have had much of an effect.



# Unsupervised modelling

A PCA of the samples, based on the features which for all the sources are found
in at least 30% of the samples within that source. This means that all features 
specific for a matrix is excluded. Thus it is expected that the seperation will 
be much smaller than the earlier PCA plot, where matrix specific features would 
naturally be a large part of the variation seen.

Note also that this PCA is performed on the actual concentrations without log2
transformation or scaling. PCA separation will thus mostly be driven by features
showing large differences.

```{r PCAOfAbundances, fig.cap = "PCA of all common features by abundance. In plot showing the 1st and 2nd dimension, 6 outliers are found. Some difference by martrix is seen as well, with the capillary being separated, this is also seen in the 3rd dimension.", echo = FALSE}

#' Note that this is done using data for which log10 has not been taken, if this is done the outliers are not found
#' A PCA is performed of the samples.
pc <- prcomp(t(assay(vams_neg, "Imputed")[feats_common, ]),
             scale = FALSE, center = TRUE)
pchs <- ifelse(vams_neg$sex == "F", 21, 22)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1))
plot_pca(pc, pc_x = 1, pc_y = 2, col = paste0(col_source[vams_neg$source], "ce"),
         bg = paste0(col_source[vams_neg$source], 60), pch = pchs, cex = 1.5)
legend("topleft", col = col_source, legend = names(col_source), pch = 16)
plot_pca(pc, pc_x = 3, pc_y = 4, col = paste0(col_source[vams_neg$source], 80),
         bg = paste0(col_source[vams_neg$source], 40), pch = pchs, cex = 1.5)


```

As with the positive mode, the plasma easily separate from the other samples, 
The seperation is however stronger in this case, and in the first dimension as 
opposed to the second. Likewise the venous separate to a lesser extent in the 
first dimension. The capillary and the RBC separate to some extent from one 
another in the second dimension.
The thrid and fourth dimension does not seem to hold any information.

# Supervised modelling

## Linear Model

The multiple linear regression of the features is made without imputation.
The models were made using the capillary as a baseline. Since the RBC and the
plasma were made from the venous it is expected to see large differences in all 
the models made. Although it would be expected from the earlier PCA plots, that
the capillary is more akin to the RBC.

In this analysis we fit the regression model to log2 transformed abundances of
each feature and extract the coefficients from the fitted model representing the
slope (or difference) of the average abundance of samples from the respective
matrix to the base line capillary samples. The p-value represents the
significance from the Wald test against the null hypothesis of the coefficient
being 0 (i.e. no differences between the average abundances).

The p-values are subsequently adjusted for multiple hypothesis testing using the
method from Benjamini and Hochberg for a strong control of the false discovery
rate.

The employed model `conc ~ source + sex` adjusts for sex-specific differences. A
simple model without accounting for any confounding effects yielded however
similar results (not shown).

```{r ModelforlinearmodelVSplasmanoconfounding, warning = FALSE}
SourcesTypes <- c("sourceplasma","sourceRBC","sourcevenous")

#' The function for the multiple linear regression is made. This takes a numeric
#' input along with a matrix containing the confounding factors, it returns the
#' means and the p value
fitlmFunction <- function(x, dataMat, model = conc ~ source) {
    dataMat$conc <- x
    dataMat$source <- factor(x = dataMat$source, 
                             levels = c("capillary", "plasma", "RBC", "venous"))
    mdl <- lm(model, data = dataMat)
    smry <- data.frame(summary(mdl)$coefficients)
    SourceTypes <- paste0("source", levels(dataMat$source)[-1])
    Output <- c(coef = smry[SourceTypes, 1], pvalue =  smry[SourceTypes, 4])
    names(Output) <- c(paste0("coef.", SourceTypes),
                       paste0("pvalue.", SourceTypes))
    Output
}

#' The data is log2 transformed to make it mostly gaussian in distribution  
Data <- log2(assay(vams_neg, "Imputed"))
fitMatrix <- t(apply(Data, 1, fitlmFunction, 
                     dataMat = as.data.frame(colData(vams_neg)),
                     model = conc ~ source + sex))

#' Adjusting for multiple hypothesis testing
p.adjusted <- apply(fitMatrix[ ,paste0("pvalue.", SourcesTypes)], 2, p.adjust, 
                    method = "BH")
#' The fitMatrix is appended to the rest of the data in vams_neg, making the
#' new dataframe res_matrix_neg
res_matrix_neg<- cbind(rowData(vams_neg), fitMatrix, adj = p.adjusted)
```

Histograms of each of the 3 matrixes against capillary is made.

```{r Histogramofpvalues, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Histogram of the p values each of the linear models against capillary. All histograms show a smooth decrease over the range of pvalues, indicating only a small possible effect from confounding factors"}
TestedSourcenames <-  c("plasma", "RBC", "venous")
par(mfrow=c(1, 3))
for (i in 1:3) {
    hist(res_matrix_neg[, paste0("pvalue.", SourcesTypes[i])],
         main = paste0(TestedSourcenames[i], " vs capillary"),
         xlab = "p-value")
}


```


It looks like the p-values decrease gradually, this suggest that the effect of
any confounding effects are likely to be minor, compared to the overall
differences from the matrices. We do however have a very large number of
features with differences in their abundances.

Based on the models created in the previous section volcano plots are made for
each of the matrices compared to capillary.

```{r VolcanoPlotMultipleRegression, fig.path = IMAGE_PATH, echo = FALSE, fig.cap = "Volcano plots of each of the linear models against capillary. Benjamini-Hochberg adjusted values, against the mean differences for each feature, show a large proportion of the features are statistically significantly different between matrixes", fig.width = 12, fig.height = 4}
#' A new color set is defined to better conform to what is need by the volcano plot
col_volcano <- brewer.pal(4, name = "Set1")
col_volcano[1] <- "#DDDDDD" #' Grey
col_volcano_Trans <- paste0(col_volcano, 80)

pLimit <- 0.01
foldLimit <- 1

#' The filters for pvalue and fold increase is defined
colorFilterLowP <- as.matrix(
    res_matrix_neg[, paste0("adj.pvalue.", SourcesTypes)]) < pLimit
colorFilterHighFold <- abs(as.matrix(
    res_matrix_neg[, paste0("coef.", SourcesTypes)])) > foldLimit

#' The filter for coloring the points is made
colorFilter <- 1 * colorFilterLowP
colorFilter[colorFilterHighFold] <- 2
colorFilter[colorFilterHighFold * colorFilterLowP == 1] <- 3
colorFilter <- colorFilter + 1

#' The significant samples (both high fold difference and low p-value) are
#' defined
sign <- colorFilterLowP & colorFilterHighFold
#' replace the "coef" in the column names with "significant"
colnames(sign) <- sub("adj.pvalue", "significant", colnames(sign))
res_matrix_neg <- cbind(res_matrix_neg, sign)

par(mfrow=c(1, 3), mar = c(4.3, 4.3, 1, 1))
for (i in 1:3){
    plot(res_matrix_neg[, paste0("coef.", SourcesTypes[i])],  cex = 0.6,
         -log10(res_matrix_neg[, paste0("adj.pvalue.", SourcesTypes[i])]),
         pch = 16, main = paste0(TestedSourcenames[i], " vs capillary"), 
         col = col_volcano_Trans[colorFilter[, i]], 
         ylab = expression(log[10]~p[BH]-value), 
         xlab = expression(log[2]~fold~change))
    abline(h = -log10(pLimit))
    abline(v = c(-foldLimit, foldLimit))
    legend("top", c("non-Significant", "lowP", "Highfold", "Significant"),
           col = col_volcano, pch = 16, bg = "white")
}


```


A surprisingly large amount of features were detected to have statistically
different abundances between the compared matrices. This was expected for the
RBC and plasma, but not for the venous one, but is consistent with what was seen
in the PCA plots. The number of significant features for each comparison are
listed below.

```{r count-table_linear_model, results = "asis", echo = FALSE}
cpt <- paste0("Number of features with a significant difference in abundances",
              " (more than two-fold difference in abundances at a 1% FDR.")
pandoc.table(colSums(as.matrix(
    res_matrix_neg[, grep("significant", colnames(res_matrix_neg))]),
    na.rm = TRUE), style = "rmarkdown", caption = cpt)
```



As expected, the largest number of features with differences in abundances was
identified between plasma and capillary samples. The number of significant
features is much smaller for the comparisons between RBC and capillary and for
venous and capillary samples.

The large differences may have many sources. To help identify these it was
decided to probe the underlying distributions of the features based on the
two factors we know. The mass/charge ratio and the retention time. It would be
expected that the venous and capillary would be mostly the same with the plasma
and RBC being different.

Plots showing the distributions of the matrixes based on m/z were made.

```{r pValueDependencyOnmzNoconfounding, fig.cap = "Distribution of m/z values (left) or retention time (right) for detected features in a matrix.", echo = FALSE, fig.width = 12, fig.height = 6}

#' Distributions of mz values for each matrix.
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
plot(density(res_matrix_neg$mzmed[present_call[, 1]], na.rm = TRUE),
     col = paste0(col_source[colnames(present_call)[1]], 80),
     main = "Detected features in a matrix",
     xlab = "m/z", ylim = c(0, 0.002), lwd = 2)
for (i in 2:4) {
    lines(density(res_matrix_neg$mzmed[present_call[, i]], na.rm = TRUE), 
          col = paste0(col_source[colnames(present_call)[i]], 80), lwd = 2)
}
legend("topleft", legend = c(colnames(present_call)),
       col = col_source[colnames(present_call)], pch = 16)
#' retention time
plot(density(res_matrix_neg$rtmed[present_call[, 1]], na.rm = TRUE),
     col = paste0(col_source[colnames(present_call)[1]], 80),
     main = "Detected features in a matrix",
     xlab = "rt", ylim = c(0, 0.015), lwd = 2)
for (i in 2:4) {
    lines(density(res_matrix_neg$rtmed[present_call[, i]], na.rm = TRUE), 
          col = paste0(col_source[colnames(present_call)[i]], 80), lwd = 2)
}

```


Looking at the distribution of detected features as a function of their m/z,
The RBC and capillary, and the plasma and venous. Generally follow the same
distributions.
For RBC and capillary, a higher proportion of low mz features are found.
Which conversely means that for plasma and venous a higher proportion of higher 
mz features is found.

Looking at the retention time instead. It is the same story. The plasma 
generally follow the venous. The capillary and RBC are however more different 
especially in the high retention time region.


Next we evaluate the distribution of m/z and retention times for the features
with significant differences in abundances compared to capillary samples.

```{r mzSignificant, fig.cap = "Distributions of m/z values (left) and retention times (right) for features with significant different abundances compared to the levels in capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6}
par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1, 1))
#' m/z
plot(density(res_matrix_neg$mzmed, na.rm = TRUE),
     col = "#cecece", main = "Significant features",
     xlab = "m/z", ylim = c(0, 0.002), lwd = 2)
for (i in 1:3) {
    lines(density(
        res_matrix_neg$mzmed[res_matrix_neg[, paste0("significant.",
                                                     SourcesTypes[i])]],
        na.rm = TRUE), col = paste0(col_source[TestedSourcenames[i]], 80),
        lwd = 2)
}
legend("topleft", legend=c("All features", TestedSourcenames), 
       col = c("#cecece", col_source[TestedSourcenames]), pch = 16)
#' rt
plot(density(res_matrix_neg$rtmed, na.rm = TRUE),
     col = "#cecece", main = "Significant features",
     xlab = "rt", ylim = c(0, 0.04), lwd = 2)
for (i in 1:3) {
    lines(density(
        res_matrix_neg$rtmed[res_matrix_neg[, paste0("significant.",
                                                     SourcesTypes[i])]],
        na.rm = TRUE), col = paste0(col_source[TestedSourcenames[i]], 80),
        lwd = 2)
}
```

This distribution includes only the features shown to be significantly different
in the previous model. The overall shape of the density lines is mostly similar
with the one for all features with the most notable differences in the m/z range
from 200 to 700.

The distribution of significant features along the retention time dimension
differs considerably between matrices. THe venous being the most striking.
in the low rt range (0-150), the venous has very few significantly different
features compared to capillary, comparing to the overall distribution of values.
At the 150 mark an marked increase is seen and from here to 200 a very large 
difference is seen. Two large peaks at aprox. 160 and 220 is noticeably.
For RBC the distribution mostly follows the overall distribution for the low 
to medium values (0-150) with a slight over representation at 160-200.
Plasma follows the distribution at at low rt (0-100) then like venous it is 
underrepresentated at 100-150, and increased at 150-220. Although not as extreme.


Next we evaluate over- and under-represented features (compared to capillary
samples) separately.

```{r mzCountsSignificant, fig.cap = "Distribution of m/z values (left) and retention time (right) for significant features with a higher or lower abundance compared to capillary samples.", echo = FALSE, fig.width = 12, fig.height = 6}

par(mfrow = c(1, 2), mar = c(4.3, 4.3, 1.5, 1))
plot(x = 1, y = 1, pch = NA, xlim = c(50, 1100), ylim = c(-100, 100),
     xlab = "m/z", ylab = "count", main = "Significant features")
text(x = c(50, 50), y = c(-90, 90), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_neg[
        which(res_matrix_neg[, paste0("significant.source", src)] & 
              res_matrix_neg[, paste0("coef.source", src)] < 0), "mzmed"]
    yLow <- hist(yLow, breaks = seq(50, 1100, 5), plot = FALSE)   
    yHigh <- res_matrix_neg[
        which(res_matrix_neg[, paste0("significant.source", src)] & 
              res_matrix_neg[, paste0("coef.source", src)] > 0), "mzmed"]
    yHigh <- hist(yHigh, breaks = seq(50, 1100, 5), plot = FALSE)
    lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
}
grid()
abline(h = 0)
legend("topright", legend = TestedSourcenames, lwd = 2, lty = 1,
       col = col_source[TestedSourcenames], inset = c(0, 0))
plot(x = 1, y = 1, pch = NA, xlim = c(0, 250), ylim = c(-800, 800),
     xlab = "rt", ylab = "count", main = "Significant features")
text(x = c(0, 0), y = c(-750, 750), pos = c(4, 4),
     label = c("lower abundance than capillary",
               "higher abundance than capillary"))
for (src in TestedSourcenames) {
    yLow <- res_matrix_neg[
        which(res_matrix_neg[, paste0("significant.source", src)] & 
              res_matrix_neg[, paste0("coef.source", src)] < 0), "rtmed"]
    yLow <- hist(yLow, breaks = seq(0, 255, 5), plot = FALSE)   
    yHigh <- res_matrix_neg[
        which(res_matrix_neg[, paste0("significant.source", src)] & 
              res_matrix_neg[, paste0("coef.source", src)] > 0), "rtmed"]
    yHigh <- hist(yHigh, breaks = seq(0, 255, 5), plot = FALSE)
    lines(yHigh$mids, yHigh$counts, col = paste0(col_source[src], 80), lwd = 2)
    lines(yLow$mids, -yLow$counts, col = paste0(col_source[src], 80), lwd = 2)
}
grid()
abline(h = 0)

```
 As opposed to the positive mode the plasma sample is characterized by having
 lower abundance for features in the low m/z range. In the 400-1000 range, the
 venous generally have fewer low abundancy features than the RBC and plasma 
 which follows one another quite well. The RBC samples are generally have fewer
 samples with a higher abundancy throughout the entire range. plasma and venous
 are comparable in this regard.

The distribution of retention times, representing the polarity, for significant
features is very informative too (right plot in the figure above).
For the low RT, the the plasma has a lot of features whuch are both over- and 
underrepresentated. at the 125-220 range most of the plasma mostly follows the
two other matrices. With the exception of 200-250 where it has more 
underrepresentated features. 
The RBC has most of its significant features as underrepresentated in the low 
RT range.
apart from about 150, the RBC follows the two other matrices. Here a large amount
of features are underrepresentated. likewise a very large peak or perfectly 
overlapping size of overrepresentated features for plasma and venous are found,
but not for RBC.
For venous, it seems to act as an average of the venous and RBC, as expected.




# Session information

```{r sessionInfo}
devtools::session_info()
```