---
title: "Preprocessing of the vams vs intraveneous untargeted metabolomics data"
author: "Johannes Rainer, Giuseppe Paglia and Sigurdur Smarason"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
BiocStyle::markdown() 
```

**Modified**: `r file.info("vams_preprocessing.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r settings, echo = FALSE, results = "hide", message = FALSE}

## Set general options
options(useFancyQuotes = FALSE)
set.seed(18011977)

## Define paths:
filename <- "vams_preprocessing"
## Path to save the images; remove all old images.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
## Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
if (!file.exists(RDATA_PATH)) dir.create(RDATA_PATH, recursive = TRUE)

## Define the path where we can find the mzML files:
MZML_PATH <- "~/data/2018/"
if (!file.exists(MZML_PATH))
    stop("Can not find the directory with the mzML files: ", MZML_PATH)

## Wheter all calculations should be performed or cached data is to be used.
use_cached <- FALSE
## Run only the chunks that are required to produce the data.
just_analysis <- FALSE

## Where to cut the chromatogram (i.e. subset to spectra within rt 0-rt_cut)
rt_cut <- 250

```

# Introduction

In this section we perform the pre-processing of the untargeted metabolomics
data of the *MitYOU* project which comprises an initial quality assessment,
chromatographic peak detection, alignment (retention time correction) and
correspondence (chromatographic peak grouping across samples).

The input files represent centroided data, with centroiding performed using the
`MSnbase` package based on the following settings: smoothing of the signal in
retention time dimension by averaging signal from 3 neighboring spectra in a
moving window approach followed by smoothing of each spectrum (along the m/z
dimension) using a Savitzky-Golay filter with a half window size of 6. The data
was subsequently centroided refining in addition the centroid's m/z value with
an intensity-weighted average of m/z values for signal within the peak with at
least 33% of the centroid's intensity.

# Positive polarity

We first focus on the data in positive polarity.

## Raw data import and initial quality assessment

Below we load all required libraries for the analysis as well as the *phenodata*
file defining the experiment's layout. Also, we set up the parallel processing,
define colors to be used for the various experimental groups throughout the
experiment and finally read the raw data.

```{r pos-libraries-data, message = FALSE}
library(xcms)
library(RColorBrewer)
library(pander)
library(doParallel)
registerDoParallel(3)
register(DoparParam(), default = TRUE)

pd <- read.table("data/phenodata.txt", sep = "\t", as.is = TRUE, header = TRUE)
pd_pos <- pd[pd$polarity == "POS", ]

## Define colors for the groups.
col_source <- brewer.pal(5, name = "Set1")
names(col_source) <- c("RBC",           # red
                       "plasma",        # blue
                       "all",           # green
                       "capillary",     # purple
                       "venous")        # orange
## Read the data
data_pos <- readMSData(paste0(MZML_PATH, sub(".mzML", "_csp1_SG6_dp33.mzML",
                                             pd_pos$mzML_file)),
                       pdata = as(AnnotatedDataFrame(pd_pos),
                                  "NAnnotatedDataFrame"), mode = "onDisk")
```

All samples of the experiment are listed below.

```{r pos-phenodata, message = FALSE, results = "asis", echo = FALSE}
validatePdata <- function(x) {
    src <- c(`1` = "plasma", `2` = "venous", `3` = "capillary", RBC = "RBC")
    fn <- strsplit(x$mzML_file, split = "_")
    for (i in 1:nrow(x)) {
        splt <- fn[[i]]
        if (splt[1] != x$batch[i])
            message("Row ", i, ": batch wrong")
        if (length(splt) == 5) {
            if (as.numeric(splt[2]) != x$sample[i])
                message("Row ", i, ": sample wrong")
            if (src[splt[3]] != x$source[i])
                message("Row ", i, ": source wrong")
            if ((splt[4] == "a" & x$replicate[i] != 1) ||
                (splt[4] == "b" & x$replicate[i] != 2))
                message("Row ", i, ": replicate wrong")
        } else {
            if (x$sample[i] != "POOL")
                message("Row ", i, ": sample wrong")
        }
    }
}
## validatePdata(pData(data_pos))

tab <- pData(data_pos)[, -4]
pandoc.table(tab, style = "rmarkdown",
             caption = paste0("Data files and samples for positive polarity"))

```

The data set consists thus of samples from in total 22 individuals, with
measurements performed on plasma, venous and capillary blood samples as well as
on red blood cells for each. Measurements were performed in two runs, with each
sample measured in duplicates within each run. A pool of all samples (all
sources and samples) was used as quality control and repeatedly measured (every
9 injections) during a measurement run. The injection order layout of one
measurement run is visualized below.

```{r pos-run-plot, fig.width = 10, fig.height = 2, fig.path = IMAGE_PATH, fig.cap = "Layout of a measurement run."}
tmp <- pData(data_pos)[data_pos$batch == 31012018, ]
par(mar = c(2, 0, 2, 0))
plot(3, 3, pch = NA, xaxt = "n", yaxt = "n", ylim = c(0, 2),
     xlim = c(0.5, nrow(tmp) + 0.5), bty = "n")
rect(xleft = tmp$inj_idx - 0.5, xright = tmp$inj_idx + 0.5, ybottom = 0,
     ytop = 1, col = col_source[tmp$source], border = "grey")
axis(side = 1, line = 0, las = 2)
text(x = tmp$inj_idx, y = rep(0.5, nrow(tmp)), labels = tmp$sample,
     srt = 90, cex = 0.7)
legend("top", col = col_source, legend = names(col_source), horiz = TRUE,
       pch = 15)
```

Sample source type and individuals are randomly distributed along the injection
order.

Next we perform an initial quality assessment by comparing the distribution of
TIC and BPC values from the spectra between all samples.

- plot the TIC and BPC
- RLA of TIC/BPC values: identify samples with much lower signal.

### Summary

## Chromatographic peak detection

- Perform peak detection.
- Compare numbers of identified peaks across samples.
- Compare intensity distributions across samples.
- Image of peak positions along retention time.
- Check chromatographic peaks for some known controls.

### Summary

## Alignment

## Correspondence

# References
